---
documentclass: jss
author:
  - name: Paul Govan
    orcid: 0000-0002-1821-8492
    affiliation: Independent Researcher
    email: \email{paul.govan2@gmail.com}
    url: https://paulgovan.github.io/PRA/
title:
  formatted: "\\pkg{PRA}: Quantitative Project Risk Analysis in \\proglang{R}"
  plain:     "PRA: Quantitative Project Risk Analysis in R"
  short:     "\\pkg{PRA}: Project Risk Analysis"
abstract: >
  The \pkg{PRA} package for \proglang{R} provides a unified, open-source
  implementation of the core analytical methods used in quantitative project
  risk analysis: the Second Moment Method (SMM), Monte Carlo simulation (MCS),
  earned value management (EVM), contingency and sensitivity analysis,
  Bayesian risk inference and updating, sigmoidal learning curves,
  and design structure matrices (DSM). Building on prior theoretical work
  that introduced causal networks and structural network measures for project
  risk (Govan 2014; Govan and Damnjanovic 2016, 2020), \pkg{PRA} operationalizes
  these frameworks in a minimal, well-tested \proglang{R} package available on
  CRAN. Version 0.4.0 further adds an experimental probabilistic network
  module for forward simulation and Bayesian conditioning along causal chains,
  and an AI agent layer that exposes all analytical functions as tools callable
  by large language models via slash commands, natural language queries, and a
  retrieval-augmented generation knowledge base.
  The package adheres to rOpenSci Statistical Software Review standards and is
  supported by a 640-test suite running on continuous integration across
  macOS, Windows, and Ubuntu.
keywords:
  formatted: [project risk analysis, Monte Carlo simulation, earned value management,
              "Bayesian inference", "design structure matrix", "\\proglang{R}"]
  plain:     [project risk analysis, Monte Carlo simulation, earned value management,
              Bayesian inference, design structure matrix, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: pra-jss.bib
---

```{r setup, include=FALSE}
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center",
                      fig.width = 5, fig.height = 3.5,
                      out.width = "80%")
set.seed(42)
library(PRA)
```

# Introduction

Quantitative risk analysis is a foundational discipline in project management,
enabling organizations to assess schedule and cost uncertainty, identify
high-impact risk drivers, and set evidence-based contingency reserves.
Govan (2014) and Govan and Damnjanovic (2016) introduced the first
implementation of causal networks for project risk analysis grounded in the
resource-based view, demonstrating that organizational capabilities and
shared resources---not just schedule uncertainty---are the structural
pathways through which risks propagate across project tasks.
Extending this work, Govan and Damnjanovic (2020) developed structural
network measures that quantify the topology of project risk networks,
enabling systematic identification of high-centrality risk nodes.

Despite broad industry adoption of quantitative methods, open-source tooling
for integrated project risk analysis in \proglang{R} remains fragmented.
Proprietary platforms such as Oracle Primavera Risk Analysis and Palisade
\@RISK provide integrated workflows but impose licensing costs and limit
reproducibility. Within \proglang{R}, no single package unifies Monte Carlo
schedule risk, earned value management, Bayesian risk inference, sigmoidal
learning curves, and dependency structure analysis under a common API.
The \pkg{mc2d} package \citep{mc2d} provides general-purpose two-dimensional
Monte Carlo simulation but has no project-specific constructs.
The \pkg{sensitivity} package implements variance-based indices for general
functions but does not integrate with schedule or cost models. EVM metrics
are absent from CRAN entirely. Bayesian packages such as \pkg{rstan} and
\pkg{brms} operate at a high level of generality and require substantial
expertise to apply in project risk contexts.

\pkg{PRA} fills this gap. It provides a coherent, CRAN-validated framework
of ten analytical modules covering the full quantitative project risk
workflow---from uncertainty characterization and simulation to performance
monitoring and Bayesian updating---while operationalizing the causal and
network-theoretic frameworks of Govan (2014) and Govan and Damnjanovic
(2016, 2020) computationally. The package targets project analysts who work
in \proglang{R} and require reproducible, auditable outputs compatible with
standard reporting pipelines. It adheres to rOpenSci Statistical Software
Review (SRR) standards \citep{rOpenSciStats2021} and uses minimal hard
dependencies: \pkg{mc2d} \citep{mc2d} for triangular distribution families
and \pkg{minpack.lm} \citep{Elzhov2016} for nonlinear least squares fitting.
All other capabilities (AI agent, Shiny app, MCP server) activate through
optional \code{Suggests}-only dependencies.

The following table summarizes how \pkg{PRA} compares with the most
closely related \proglang{R} packages and the dominant commercial
platform.

```{r comparison-table, echo=FALSE}
comp <- data.frame(
  Feature = c(
    "MCS schedule risk", "EVM (11 metrics)", "Bayesian updating",
    "Sigmoidal learning curves", "Design structure matrices",
    "Probabilistic networks", "AI agent layer",
    "Open-source (CRAN)", "SRR-compliant"
  ),
  PRA         = c("Yes","Yes","Yes","Yes","Yes","Exp.","Yes","Yes","Yes"),
  mc2d        = c("Partial","---","---","---","---","---","---","Yes","---"),
  sensitivity = c("---","---","---","---","---","---","---","Yes","---"),
  `rstan/brms`= c("---","---","Yes","---","---","Yes","---","Yes","---"),
  `@RISK`     = c("Yes","Yes","---","---","---","---","---","---","---"),
  check.names = FALSE
)
knitr::kable(
  comp,
  caption = paste0("Comparison of \\pkg{PRA} with related \\proglang{R} packages",
    " and commercial tools. ``Exp.'' = experimental feature;",
    " ``---'' = not supported."),
  booktabs = TRUE, align = "lccccc"
)
```

The package is installed from CRAN in the standard way:

```{r install, eval=FALSE}
install.packages("PRA")
library("PRA")
```

Optional dependencies for the AI agent layer---\pkg{ellmer}, \pkg{ragnar},
\pkg{shiny}, \pkg{shinychat}, and \pkg{jsonlite}---are only required for
\code{pra\_chat()}, \code{pra\_app()}, and \code{pra\_mcp\_server()}.

The remainder of this article describes each analytical module in turn,
with working code examples drawn from the same \proglang{R} session.
Section~\ref{sec-uncertainty} covers schedule and cost uncertainty methods
(SMM, MCS, contingency, sensitivity). Section~\ref{sec-evm} describes
earned value management. Section~\ref{sec-bayes} covers Bayesian risk
inference and updating. Section~\ref{sec-learning} describes sigmoidal
learning curves. Section~\ref{sec-dsm} covers design structure matrices.
Section~\ref{sec-network} introduces the experimental probabilistic
network module. Section~\ref{sec-agent} describes the AI agent framework.
Section~\ref{sec-casestudy} presents an integrated case study.
Section~\ref{sec-summary} concludes.


# Schedule and Cost Uncertainty {#sec-uncertainty}

Two complementary methods are provided for quantifying schedule and cost
uncertainty: the Second Moment Method for rapid analytical estimates, and
Monte Carlo simulation for full distributional analysis. Contingency
extraction and sensitivity analysis complete the workflow.

## Second Moment Method

The Second Moment Method (SMM) propagates task-level means and variances to
project-level estimates using the Central Limit Theorem \citep{Benjamin2000}.
For a project with $n$ tasks:
$$E[X] = \sum_{i=1}^{n} E[X_i], \quad
  \text{Var}(X) = \sum_{i=1}^{n}\text{Var}(X_i) + 2\sum_{i<j}\rho_{ij}\sigma_i\sigma_j$$
The \code{smm()} function takes vectors of task means and variances plus a
correlation matrix and returns the total mean, variance, and standard deviation.

```{r smm-example}
task_means <- c(10, 15, 20)   # weeks
task_vars  <- c(4,  9,  16)
cor_mat <- matrix(c(
  1.0, 0.5, 0.3,
  0.5, 1.0, 0.4,
  0.3, 0.4, 1.0
), nrow = 3, byrow = TRUE)

result <- smm(task_means, task_vars, cor_mat)
cat("Total mean:", round(result$total_mean, 2), "weeks\n")
cat("Total SD:  ", round(result$total_std,  2), "weeks\n")
ci <- result$total_mean + c(-1, 1) * 1.96 * result$total_std
cat("95% CI:    [", round(ci[1], 1), ",", round(ci[2], 1), "] weeks\n")
```

SMM is useful when only first and second moments are available and a rapid
estimate is needed. When task distributions are non-normal or accurate tail
behavior is required, Monte Carlo simulation is preferred.

## Monte Carlo Simulation

\code{mcs()} propagates uncertainty through a project network by drawing
samples from user-specified task distributions---triangular, normal, or
uniform---over a large number of iterations \citep{Vose2008}. Task correlations
are incorporated via Cholesky decomposition of the supplied correlation matrix.

```{r mcs-setup}
num_simulations   <- 10000
task_distributions <- list(
  list(type = "normal",     mean = 10, sd = 2),
  list(type = "triangular", a = 5, b = 10, c = 15),
  list(type = "uniform",    min = 8, max = 12)
)
```

```{r mcs-run}
results <- mcs(num_simulations, task_distributions, cor_mat)
cat("Simulated mean:", round(results$total_mean, 2), "weeks\n")
cat("Simulated SD:  ", round(results$total_sd,   2), "weeks\n")
```

The \code{mcs()} function returns an S3 object of class \code{"mcs"} with
a \code{print()} method that summarizes percentiles. The full simulated
distribution is available in \code{results\$total\_distribution}.
On a modern laptop, 10,000 simulations of a 10-task network complete in
under one second; increasing \code{num\_simulations} to 100,000 provides
more precise tail estimates at roughly ten-fold additional runtime.
The default of 10,000 balances accuracy and speed for most practical analyses.

```{r mcs-hist, fig.cap="Distribution of simulated total project durations. The dashed vertical line marks the mean."}
hist(results$total_distribution,
  breaks = 50, freq = FALSE,
  main = "MCS: Total Project Duration",
  xlab = "Duration (weeks)", col = "steelblue", border = "white"
)
lines(density(results$total_distribution), col = "tomato", lwd = 2)
abline(v = results$total_mean, lty = 2, lwd = 1.5)
```

## Contingency Analysis

\code{contingency()} extracts a reserve buffer as the difference between a
high-confidence percentile and the base (median) estimate---a standard
approach in project risk practice \citep{PMBOK2021}.

```{r contingency}
reserve <- contingency(results, phigh = 0.80, pbase = 0.50)
cat("Schedule contingency (P80 - P50):", round(reserve, 2), "weeks\n")
```

## Sensitivity Analysis

\code{sensitivity()} decomposes total project variance by task contribution
using rank correlation, producing the inputs for a Tornado chart.

```{r sensitivity, fig.cap="Tornado chart: each bar shows a task's rank correlation with total project duration."}
sens <- sensitivity(task_distributions, cor_mat)
barplot(sort(sens),
  horiz = TRUE, col = "steelblue",
  main = "Sensitivity: Task Contributions",
  xlab = "Rank Correlation with Total"
)
```

Tasks with larger bars are the dominant drivers of total risk; mitigation
resources should be directed there first.


# Earned Value Management {#sec-evm}

The EVM module implements the complete ANSI/EIA-748 suite of eleven
performance metrics \citep{Fleming2010}. The three core quantities---Planned
Value (PV), Earned Value (EV), and Actual Cost (AC)---are computed first;
all performance indices and forecasts follow from them.

```{r evm-setup}
bac         <- 500000                          # budget at completion ($)
schedule    <- c(0.10, 0.25, 0.50, 0.75, 1.00) # cumulative value curve
time_period <- 3
```

```{r evm-core}
pv_val <- pv(bac, schedule, time_period)
ev_val <- ev(bac, actual_per_complete = 0.40)
ac_val <- ac(c(45000, 110000, 135000), time_period, cumulative = FALSE)

cat("PV: $", format(pv_val, big.mark = ","), "\n")
cat("EV: $", format(ev_val, big.mark = ","), "\n")
cat("AC: $", format(ac_val, big.mark = ","), "\n")
```

```{r evm-indices}
spi_val  <- spi(ev_val, pv_val)
cpi_val  <- cpi(ev_val, ac_val)
sv_val   <- sv(ev_val,  pv_val)
cv_val   <- cv(ev_val,  ac_val)

cat("SPI:", round(spi_val, 3), "  CPI:", round(cpi_val, 3), "\n")
cat("SV: $", format(round(sv_val), big.mark = ","),
    "  CV: $", format(round(cv_val), big.mark = ","), "\n")
```

Both SPI $< 1$ and CPI $< 1$ indicate the project is behind schedule and over
budget. The \code{eac()} function provides three completion forecasts under
different efficiency assumptions:

```{r evm-eac}
cat("EAC (typical):  $", format(round(eac(bac, method = "typical",
    cpi = cpi_val)), big.mark = ","), "\n")
cat("EAC (atypical): $", format(round(eac(bac, method = "atypical",
    ac = ac_val, ev = ev_val)), big.mark = ","), "\n")
cat("EAC (combined): $", format(round(eac(bac, method = "combined",
    cpi = cpi_val, ac = ac_val, ev = ev_val, spi = spi_val)),
    big.mark = ","), "\n")
```

The remaining EVM metrics are:

```{r evm-remaining}
eac_typ <- eac(bac, method = "typical", cpi = cpi_val)
cat("ETC:  $", format(round(etc(bac, ev_val, cpi = cpi_val)), big.mark = ","), "\n")
cat("TCPI:", round(tcpi(bac, ev_val, ac_val), 3), "\n")
cat("VAC:  $", format(round(vac(bac, eac_typ)), big.mark = ","), "\n")
```

The To-Complete Performance Index (TCPI) reports the cost efficiency required
on all remaining work to finish within the original budget; values substantially
above 1.0 signal an unrealistic recovery target.


# Bayesian Risk Methods {#sec-bayes}

Building on the causal structure introduced in Govan (2014) and
Govan and Damnjanovic (2016), the Bayesian module models risk event
probabilities from root causes and supports formal updating as project
observations become available \citep{Gelman2013,Pearl2009}.

## Prior Risk Probability

\code{risk\_prob()} computes the marginal probability of a risk event $R$
from multiple independent root causes using the total probability theorem:
$$P(R) = \sum_{i} P(R \mid C_i)\,P(C_i) + P(R \mid \bar{C}_i)\,P(\bar{C}_i)$$

```{r prior-risk}
cause_probs            <- c(0.30, 0.20)
risks_given_causes     <- c(0.80, 0.60)
risks_given_not_causes <- c(0.20, 0.40)

prior <- risk_prob(cause_probs, risks_given_causes, risks_given_not_causes)
cat("Prior P(R):", round(prior, 3), "\n")
```

## Bayesian Updating

\code{risk\_post\_prob()} updates the risk probability after observing cause
status on site. Unobserved causes are passed as \code{NA} and are excluded
from the update.

```{r posterior-risk}
observed_causes <- c(1, NA)  # Cause 1 present; Cause 2 not yet assessed
post <- risk_post_prob(cause_probs, risks_given_causes,
                       risks_given_not_causes, observed_causes)
cat("Posterior P(R):", round(post, 3), "\n")
cat("Change:        ", round(post - prior, 3), "\n")
```

Observing Cause~1 (which has a strong link to $R$) raises the risk probability
substantially. The posterior informs contingency updates mid-project.

## Cost Risk Distributions

\code{cost\_pdf()} and \code{cost\_post\_pdf()} sample prior and posterior
cost distributions by simulating which risks materialize and summing their
cost impacts.

```{r cost-dist, fig.cap="Prior and posterior cost distributions. Observing Cause 1 shifts the distribution rightward, increasing the expected cost impact."}
risk_probs        <- c(0.3, 0.5, 0.2)
means_given_risks <- c(10000, 15000, 5000)
sds_given_risks   <- c(2000,  1000,  1000)
base_cost         <- 2000

prior_costs <- cost_pdf(5000, risk_probs, means_given_risks,
                        sds_given_risks, base_cost)
post_costs  <- cost_post_pdf(5000,
  observed_risks = c(1, NA, 0),   # Risk 1 present, Risk 2 unobserved, Risk 3 absent
  means_given_risks, sds_given_risks, base_cost
)

plot(density(prior_costs), col = "steelblue", lwd = 2,
     main = "Cost Risk: Prior vs. Posterior",
     xlab = "Cost ($)", xlim = range(c(prior_costs, post_costs)))
lines(density(post_costs), col = "tomato", lwd = 2)
legend("topright", legend = c("Prior", "Posterior"),
       col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

## Correlation Matrix Construction

\code{cor\_matrix()} constructs a valid positive-definite correlation matrix
from a vector of upper-triangle entries, for use as input to \code{mcs()}.

```{r cor-matrix}
dists <- list(
  normal  = function(n) rnorm(n, mean = 10, sd = 2),
  triang  = function(n) mc2d::rpert(n, min = 5, mode = 10, max = 15),
  uniform = function(n) runif(n, min = 8, max = 12)
)
C_emp <- cor_matrix(num_samples = 1000, num_vars = 3, dists = dists)
print(round(C_emp, 3))
```


# Learning Curves {#sec-learning}

Sigmoidal (S-curve) models capture the characteristic pattern of project
progress: slow start, rapid acceleration, and plateau approaching 100\%
completion. \pkg{PRA} provides three model types---Logistic, Pearl, and
Gompertz---fitted via the Levenberg--Marquardt nonlinear least-squares
algorithm in \pkg{minpack.lm} \citep{Elzhov2016}.

```{r sigmoidal-data}
progress <- data.frame(
  time       = 1:9,
  completion = c(5, 15, 40, 60, 70, 75, 80, 85, 90)
)
```

```{r sigmoidal-fit}
fit_log  <- fit_sigmoidal(progress, "time", "completion", "logistic")
fit_gomp <- fit_sigmoidal(progress, "time", "completion", "gompertz")

cat("Logistic RSE: ", round(summary(fit_log)$sigma,  2), "\n")
cat("Gompertz RSE: ", round(summary(fit_gomp)$sigma, 2), "\n")
```

```{r sigmoidal-plot, fig.cap="Logistic learning curve with 95\\% confidence band. The shaded region widens as extrapolation extends beyond the observed data range."}
plot_sigmoidal(fit_log, progress, "time", "completion", "logistic",
               conf_level = 0.95,
               main = "Logistic Learning Curve",
               xlab = "Week", ylab = "Completion (%)")
```

\code{predict\_sigmoidal()} generates forecasts with confidence bounds at
arbitrary future time points:

```{r sigmoidal-predict}
preds <- predict_sigmoidal(fit_log, seq(10, 12, by = 0.5),
                            "logistic", conf_level = 0.95)
print(round(preds, 1))
```


# Dependency Structure Analysis {#sec-dsm}

Building on Govan and Damnjanovic (2016, 2020) and the foundational DSM
literature \citep{Steward1981, Browning2001}, the DSM module quantifies
structural coupling between project tasks through shared resources and risks.

## Design Structure Matrices

The **Parent DSM** is derived from the Resource--Task matrix $\mathbf{S}$
(rows = resources, columns = tasks) as $\mathbf{P} = \mathbf{S}^\top\mathbf{S}$.
Off-diagonal entry $P_{jk}$ counts the number of resources shared between
tasks $j$ and $k$---the structural pathway through which disruptions propagate
\citep{Govan2016}.

```{r dsm-S}
S <- matrix(c(
  1, 0, 1, 0,
  1, 1, 0, 0,
  0, 1, 0, 1,
  0, 0, 1, 1,
  0, 1, 1, 0
), nrow = 4, ncol = 5)
rownames(S) <- paste0("R", 1:4)
colnames(S) <- paste0("T", 1:5)
```

```{r parent-dsm, fig.cap="Parent DSM heatmap. Darker cells indicate more shared resources and stronger coupling between task pairs."}
p <- parent_dsm(S)
plot(p)
```

The **Grandparent DSM** adds a risk layer via the Risk--Resource matrix
$\mathbf{R}$, tracing dependencies from risks through resources to tasks.
Off-diagonal entry $G_{jk}$ counts the number of risks shared between
task $j$ and task $k$ through the resource chain---a direct operationalization
of the structural network measures developed in Govan and Damnjanovic (2020).

```{r dsm-R}
R_mat <- matrix(c(
  1, 0, 1,
  1, 1, 0,
  0, 1, 0,
  0, 0, 1
), nrow = 3, ncol = 4)
rownames(R_mat) <- paste0("Risk", 1:3)
colnames(R_mat) <- paste0("R", 1:4)
```

```{r grandparent-dsm, fig.cap="Grandparent DSM heatmap. Shows risk-based coupling: task pairs that share risk exposure through common resources tend to fail together."}
g <- grandparent_dsm(S, R_mat)
plot(g)
```

Task pairs with high off-diagonal values in the Grandparent DSM represent
concentrated risk exposure; these pairs are prime candidates for contingency
buffering or resource decoupling.


# Probabilistic Networks {#sec-network}

**Experimental.** The probabilistic network module extends DSM structural
analysis by attaching probability distributions to each node, enabling
forward simulation of cost uncertainty and Bayesian conditioning on
observed evidence. A probabilistic network is a directed acyclic graph in
which nodes represent project variables (risks, resources, tasks, and the
project total) and edges encode conditional dependencies
\citep{Pearl2009, Govan2014}. Where a DSM counts shared resources between
task pairs, a probabilistic network propagates distributions through the
causal chain from risks to total project cost.

## Building a Network

\code{prob\_net()} constructs the network from a node data frame, a links
data frame, and a named list of distributions. Distributions may be
\code{normal}, \code{lognormal}, \code{uniform}, \code{discrete}, or
\code{conditional} (switching between two sub-distributions depending on
a discrete parent node); \code{aggregate} nodes sum their parents.

```{r net-setup}
nodes <- data.frame(
  id = c("A","B","C","D","E","F","G","H","I"),
  label = c("Risk-1","Risk-2","Resource-1","Resource-2","Resource-3",
            "Task-1","Task-2","Task-3","Project"),
  stringsAsFactors = FALSE
)
links <- data.frame(
  source = c("A","B","C","D","E","F","G","H"),
  target = c("C","D","F","G","H","I","I","I"),
  stringsAsFactors = FALSE
)
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)),
  D = list(type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = "C"),
  G = list(type = "aggregate", nodes = "D"),
  H = list(type = "aggregate", nodes = "E"),
  I = list(type = "aggregate", nodes = c("F","G","H"))
)
net <- prob_net(nodes, links, distributions = distributions)
```

## Forward Simulation

\code{prob\_net\_sim()} draws samples from every node in topological order
and returns a data frame with one column per node.

```{r net-sim}
sim_net <- prob_net_sim(net, num_samples = 10000)
cat("Mean project cost: $", format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("P80 project cost:  $", format(round(quantile(sim_net$I, 0.80)),
    big.mark = ","), "\n")
```

## Learning: Conditioning on Evidence

\code{prob\_net\_learn()} clamps observed nodes to fixed values and
re-simulates the downstream distribution---an observational update in the
sense of \citet{Pearl2009}. The following call fixes Risk-2 (node B) to
absent (value 0), ruling out Technical Complexity as a concern:

```{r net-learn, fig.cap="Total project cost before and after observing that Risk-2 (Technical Complexity) did not occur. Conditioning on Risk-2 = 0 eliminates the heavy upper tail and shifts the distribution leftward by roughly \\$30,000 in expectation."}
learn_net <- prob_net_learn(net, observations = list(B = 0),
                            num_samples = 10000)

plot(density(sim_net$I),   col = "steelblue", lwd = 2,
     main = "Total Project Cost: Prior vs. Posterior",
     xlab = "Cost ($)")
lines(density(learn_net$I), col = "tomato", lwd = 2)
legend("topright",
  legend = c("Prior (Risk-2 uncertain)", "Posterior (Risk-2 = 0)"),
  col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

\code{prob\_net\_update()} performs structural surgery on the network by
adding or removing edges and replacing node distributions, corresponding
to the \emph{do}-operator in Pearl's causal calculus
\citep{Pearl2009}---distinct from \code{prob\_net\_learn()}, which
conditions without altering the graph. The probabilistic network API
is marked experimental and may evolve in future releases.


# AI Agent Framework {#sec-agent}

Version 0.4.0 of \pkg{PRA} includes an optional AI agent layer that wraps
all analytical functions as tools callable by large language models. Three
interaction modes are supported, summarized below.

```{r routing-table, echo=FALSE}
routing <- data.frame(
  `Input type`   = c("\\code{/command} syntax", "Numerical data",
                     "Conceptual question"),
  Route          = c("Deterministic (no LLM)", "LLM tool call",
                     "RAG knowledge base"),
  Example        = c("\\code{/mcs n=10000 tasks=[\\ldots]}",
                     "\"Simulate 3 tasks: Normal(10,2)\\ldots\"",
                     "\"What is earned value?\""),
  check.names    = FALSE
)
knitr::kable(
  routing,
  caption = "Agent routing modes in \\pkg{PRA}.",
  booktabs = TRUE, escape = FALSE
)
```

\begin{enumerate}
  \item \textbf{Slash commands} (\code{/mcs}, \code{/evm}, \code{/risk}, etc.)
    for deterministic, LLM-free execution via \code{PRA:::execute\_command()}.
  \item \textbf{Natural language} through \code{pra\_chat()}, which connects
    to a local Ollama model (default: \code{llama3.2}) and uses
    \pkg{ellmer} \citep{ellmer} tool-calling to route queries to the
    appropriate analytical function.
  \item \textbf{Retrieval-augmented generation (RAG)} via \pkg{ragnar}
    \citep{ragnar}: conceptual questions are answered from a bundled domain
    knowledge base with hybrid BM25 + vector search and source attribution.
\end{enumerate}

An interactive Shiny application (\code{pra\_app()}) integrates all three
modes with a chat interface, data upload, and inline plots.
\code{pra\_mcp\_server()} exposes all tools via the Model Context Protocol
\citep{MCP2024}, enabling integration with Claude, Claude Code, and other
MCP-compatible clients. The server is registered in Claude Code with a
single shell command:

```{r mcp-code, eval=FALSE, engine="bash"}
claude mcp add -s project pra -- Rscript -e "PRA::pra_mcp_server()"
```

For Claude Desktop, add the following block to the MCP configuration file:

\begin{verbatim}
{
  "mcpServers": {
    "pra": { "command": "Rscript",
             "args": ["-e", "PRA::pra_mcp_server()"] }
  }
}
\end{verbatim}

The bundled knowledge base covers all core methods. Analysts can extend it
with project-specific documents using \code{build\_knowledge\_base()} and
\code{add\_documents()}:

```{r rag-extend, eval=FALSE}
store <- build_knowledge_base()
add_documents(store, "path/to/risk_register.md")
add_documents(store, "path/to/project_docs/")
```

The AI layer is strictly optional: all 36 exported analytical functions
operate without any AI dependency. Tool definitions built with
\code{ellmer::tool()} are reused identically for both the chat interface and
the MCP server, eliminating duplication. A preliminary evaluation of 25
scenarios across single-tool, multi-tool, and open-ended query tiers
yielded 72.0\% overall accuracy (87.5\% single-tool, 55.6\% multi-tool,
75.0\% open-ended), with the RAG layer providing the largest accuracy gain
on open-ended conceptual queries. A full empirical evaluation is the subject
of a forthcoming publication in \emph{Risk Analysis}.


# Illustrative Case Study {#sec-casestudy}

This section demonstrates an integrated, end-to-end project risk analysis
for a three-task construction project, progressing through uncertainty
quantification, performance monitoring, Bayesian risk updating, and
dependency structure analysis in a single \proglang{R} session.

## Step 1: Characterize Task Uncertainty

```{r cs-mcs}
tasks <- list(
  list(type = "triangular", a = 8,  b = 12, c = 18),   # earthworks
  list(type = "normal",     mean = 20, sd = 3),          # structural frame
  list(type = "triangular", a = 5,  b = 8,  c = 12)    # MEP
)
C_proj <- matrix(c(1, 0.4, 0.2,
                   0.4, 1, 0.3,
                   0.2, 0.3, 1), nrow = 3, byrow = TRUE)

sim <- mcs(10000, tasks, C_proj)
res <- contingency(sim, phigh = 0.80, pbase = 0.50)
cat("P50 schedule:", round(sim$percentiles["50%"], 1), "weeks\n")
cat("P80 schedule:", round(sim$percentiles["80%"], 1), "weeks\n")
cat("Contingency (P80-P50):", round(res, 2), "weeks\n")
```

## Step 2: Monitor Earned Value at Mid-Project

```{r cs-evm}
bac_cs   <- 1200000
sched_cs <- c(0.15, 0.35, 0.60, 0.85, 1.00)
pv_cs    <- pv(bac_cs, sched_cs, time_period = 3)
ev_cs    <- ev(bac_cs, actual_per_complete = 0.50)
ac_cs    <- ac(c(180000, 250000, 310000), time_period = 3, cumulative = FALSE)

cat("SPI:", round(spi(ev_cs, pv_cs), 3),
    " CPI:", round(cpi(ev_cs, ac_cs), 3), "\n")
cat("EAC (typical): $", format(round(eac(bac_cs,
    method = "typical", cpi = cpi(ev_cs, ac_cs))), big.mark = ","), "\n")
```

## Step 3: Update Risk Probability After Site Inspection

```{r cs-bayes}
prior_r <- risk_prob(c(0.25, 0.15), c(0.70, 0.50), c(0.15, 0.30))
post_r  <- risk_post_prob(c(0.25, 0.15), c(0.70, 0.50),
                          c(0.15, 0.30), observed_causes = c(1, NA))
cat("Prior P(R):", round(prior_r, 3),
    "-> Posterior P(R):", round(post_r, 3), "\n")
```

## Step 4: Identify Coupled Tasks via DSM

```{r cs-dsm, fig.cap="Grandparent DSM for the case-study project. Task pairs T3--T5 share the most risk exposure through common resources."}
S_cs <- matrix(c(
  1, 1, 0, 0,
  0, 1, 1, 0,
  1, 0, 0, 1,
  0, 0, 1, 1,
  0, 1, 0, 1
), nrow = 4, ncol = 5)
R_cs <- matrix(c(
  1, 0, 1,
  0, 1, 0,
  1, 1, 0,
  0, 0, 1
), nrow = 3, ncol = 4)
g_cs <- grandparent_dsm(S_cs, R_cs)
plot(g_cs)
```

The Grandparent DSM reveals which task pairs share risk exposure through
common resources, informing targeted contingency allocation and resource
decoupling decisions.


# Summary {#sec-summary}

\pkg{PRA} is a CRAN-available \proglang{R} package that provides a unified,
open-source implementation of quantitative project risk analysis methods.
It operationalizes the causal network and structural network frameworks
developed in Govan (2014) and Govan and Damnjanovic (2016, 2020) through
ten integrated analytical modules: SMM, MCS, contingency analysis,
sensitivity analysis, EVM (11 metrics), Bayesian risk inference and updating,
cost risk distributions, sigmoidal learning curves, correlation matrix
construction, and design structure matrices. An experimental eleventh
module provides probabilistic networks with forward simulation and
Bayesian conditioning.

Version 0.4.0 adds an AI agent layer that exposes all functions as tools
callable via slash commands, natural language, and RAG-powered queries
through a local Shiny application and an MCP server. All 36 exported
functions adhere to rOpenSci SRR standards, with 640+ tests verifying
correctness, edge-case handling, and numerical stability across three
operating systems. The package is licensed under MIT and the source is
available at \url{https://github.com/paulgovan/PRA}; documentation and
a companion online book are at \url{https://paulgovan.github.io/PRA/}.

## Limitations and Future Work

Several limitations apply. First, \pkg{PRA} focuses on quantitative
uncertainty propagation; it does not model schedule logic (critical path,
activity dependencies, or resource leveling), which remain the domain of
dedicated scheduling engines. Second, the probabilistic network module is
experimental: the API may change and no convergence diagnostics are
provided for the Monte Carlo estimators. Third, AI agent accuracy (72.0\%
on 25 scenarios) is preliminary; the evaluation set is small and the
routing heuristics have not been stress-tested on large or ambiguous inputs.
Fourth, the package currently accepts data as \proglang{R} vectors and lists;
direct import from project management file formats (Microsoft Project,
Primavera P6, CSV exports from Jira or Asana) is not supported and is
planned for a future release.

A forthcoming companion article in \emph{Risk Analysis} presents a full
empirical evaluation of the AI agent framework, including a 25-scenario
benchmark, RAG retrieval quality analysis, and a comparison to purely
LLM-based assistants.

# References {-}
