---
documentclass: jss
author:
  - name: Paul Govan
    orcid: 0000-0002-1821-8492
    affiliation: GE Aerospace
    email: \email{paul.govan2@gmail.com}
    url: paulgovan.com
title:
  formatted: "Quantitative Project Risk Analysis with \\pkg{PRA}"
  plain:     "Quantitative Project Risk Analysis with PRA"
  short:     "\\pkg{PRA}: Project Risk Analysis"
abstract: >
  The \pkg{PRA} package for \proglang{R} is the first package of its kind to unify a core set of widely adopted quantitative project risk analysis methods in a single, open-source, Comprehensive R Archive Network (CRAN)-available implementation: the Second Moment Method (SMM), Monte Carlo simulation (MCS), earned value management (EVM), contingency and sensitivity analysis, Bayesian risk inference and updating, sigmoidal learning curves, and design structure matrices (DSM). Beyond this foundational core, PRA makes two further contributions. As a secondary contribution, version 0.5.0 provides an open-source implementation of the causal- and probabilistic-network methods for project risk, supporting forward simulation of cost uncertainty and Bayesian conditioning along causal chains. As a third contribution, it exposes every analytical function as a tool over the Model Context Protocol (MCP), paired with a documented set of Agent Skills, so that large-language-model agents can run project risk analyses directly. The package adheres to rOpenSci Statistical Software Review standards and is supported by a continuous-integration test suite running across macOS, Windows, and Ubuntu.
  
keywords:
  formatted: [project risk analysis, Monte Carlo simulation, earned value management,
              "Bayesian inference", "design structure matrix", "\\proglang{R}"]
  plain:     [project risk analysis, Monte Carlo simulation, earned value management,
              Bayesian inference, design structure matrix, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
  \usepackage{longtable}
output: rticles::jss_article
bibliography: pra-jss.bib
---

```{r setup, include=FALSE}
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center",
                      fig.width = 5, fig.height = 3.5,
                      out.width = "80%")
set.seed(42)
library(PRA)
```

# Introduction

Quantitative risk analysis is a foundational discipline in project management,
enabling organizations to assess schedule and cost uncertainty, identify
high-impact risk drivers, and set evidence-based contingency reserves.
This article presents \pkg{PRA}, an open-source \proglang{R} package that
unifies the core quantitative methods of project risk analysis, schedule and
cost uncertainty, earned value management, sensitivity and contingency
analysis, learning curves, and dependency structure analysis, under a single,
reproducible, CRAN-validated API.

Despite broad industry adoption of quantitative methods, open-source tooling
for integrated project risk analysis in \proglang{R} remains fragmented.
Proprietary platforms such as Oracle Primavera Risk Analysis and Palisade
\@RISK provide integrated workflows but impose licensing costs and limit
reproducibility. Within \proglang{R}, no single package unifies Monte Carlo
schedule risk, earned value management, Bayesian risk inference, sigmoidal
learning curves, and dependency structure analysis under a common API.
The \pkg{mc2d} package \citep{mc2d} provides general-purpose two-dimensional
Monte Carlo simulation but has no project-specific constructs.
The \pkg{sensitivity} package \citep{sensitivity} implements variance-based
indices for general functions but does not integrate with schedule or cost
models. EVM metrics are absent from CRAN entirely. Bayesian packages such as
\pkg{rstan} \citep{rstan} and \pkg{brms} \citep{brms} operate at a high level
of generality and require substantial expertise to apply in project risk
contexts.

The primary contribution addresses the gap above directly: To the author’s knowledge, \pkg{PRA} is the first CRAN package to integrate these project risk analysis methods in a single API. The comparison in Table~\ref{tab:comparison} covers the closest general-purpose Monte Carlo and sensitivity packages and the dominant commercial platform; a broader survey of project-specific R packages is given below. Where the existing ecosystem is fragmented across general-purpose packages, none of them project-specific, \pkg{PRA} provides one reproducible, auditable API for the full workflow, integrating outputs across modules in a way the fragmented ecosystem does not. For example, MCS results feed directly into contingency and sensitivity analyses, and Bayesian risk updates can re-parameterize task distributions for subsequent simulations. The package targets project analysts who work in \proglang{R} and require outputs compatible with standard reporting pipelines. It adheres to rOpenSci
Statistical Software Review (SRR) standards \citep{rOpenSciStats2021} and uses
minimal hard dependencies: \pkg{mc2d} \citep{mc2d} for triangular distribution
families and \pkg{minpack.lm} \citep{Elzhov2016} for nonlinear least squares
fitting.

As a secondary contribution, \pkg{PRA} provides an open-source
\proglang{R} implementation, to the author's knowledge the first, of the
causal- and probabilistic-network methods for project risk introduced by
\citet{Govan2014} and \citet{Govan2016}. That work grounds causal networks in the resource-based
view, showing that organizational capabilities and shared resources, not
schedule uncertainty alone, are the structural pathways through which risks
propagate across project tasks. \citet{Govan2020} extend the approach with
structural network measures that quantify the topology of project risk
networks. The probabilistic network module
(Section~\ref{sec-network}) operationalizes these ideas, propagating
distributions through the causal chain from risks to total project cost.

As a third contribution, \pkg{PRA} is the first project risk analysis package, and to the author’s knowledge one of the first R packages in any domain, to expose its analytical methods as tools callable by LLM agents over the Model Context Protocol (MCP) and paired with a documented set of Agent Skills (Section~\ref{sec-agent}). The optional MCP server activates through \code{Suggests}-only dependencies, so the analytical core carries no additional runtime cost.

The following table underscores the primary contribution: no existing
\proglang{R} package or commercial platform covers the foundational project
risk workflow that \pkg{PRA} unifies. It summarizes how \pkg{PRA} compares with
the most closely related \proglang{R} packages and the dominant commercial
platform.

```{r comparison-table, echo=FALSE, results="asis"}
comp <- data.frame(
  Feature = c(
    "MCS schedule risk", "EVM (11 metrics)", "Bayesian updating",
    "Sigmoidal learning curves", "Design structure matrices",
    "Probabilistic networks", "MCP tool server",
    "Open-source (CRAN)", "SRR-compliant"
  ),
  PRA         = c("Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes"),
  mc2d        = c("Partial","---","---","---","---","---","---","Yes","---"),
  sensitivity = c("---","---","---","---","---","---","---","Yes","---"),
  `rstan/brms`= c("---","---","Yes","---","---","Partial","---","Yes","---"),
  `@RISK`     = c("Yes","Yes","---","---","---","---","---","---","---"),
  check.names = FALSE
)
knitr::kable(
  comp,
  format = "latex",
  caption = paste0("\\label{tab:comparison}Comparison of \\pkg{PRA} with related",
    " \\proglang{R} packages and commercial tools. ``Partial'' = only",
    " general-purpose support, requiring custom modeling and not",
    " project-specific; ``---'' = not supported. \\pkg{PRA}'s MCS supports task",
    " correlation, but its current handling is approximate",
    " (Section~\\ref{sec-summary})."),
  booktabs = TRUE, align = "lccccc"
)
```

A small number of CRAN packages do target project management specifically, but
each addresses a single slice of the workflow rather than the integrated whole.
The \pkg{ProjectManagement} package \citep{ProjectManagement} is the closest in
spirit: it implements deterministic and stochastic scheduling, PERT/CPM,
critical-path identification, and Monte Carlo simulation over activity
durations, and is complementary to the schedule-network extension planned for
\pkg{PRA} (Section~\ref{sec-summary}), but it does not provide earned value
management, Bayesian risk updating, learning curves, dependency structure
matrices, or probabilistic networks. The \pkg{plan} package \citep{plan}
produces Gantt and burn-down charts for project visualization but performs no
risk quantification. More generally, decision-analysis packages such as
\pkg{decisionSupport} \citep{decisionSupport} apply Monte Carlo simulation to
cost--benefit models under uncertainty, but are not organized around
project-schedule or earned-value constructs. To the author's knowledge, none of
these packages, nor any single package on CRAN, spans the foundational project
risk workflow that \pkg{PRA} unifies; this survey is not exhaustive, and a
formal systematic review of the ecosystem remains future work.

The package is installed from CRAN in the standard way:

```{r install, eval=FALSE}
install.packages("PRA")
library("PRA")
```

Optional dependencies, \pkg{ellmer} \citep{ellmer}, \pkg{mcptools}, and
\pkg{jsonlite}, are only required for the MCP server, \code{pra\_mcp\_server()}.

The remainder of this article describes each analytical module in turn,
with working code examples drawn from the same \proglang{R} session.
Section~\ref{sec-uncertainty} covers schedule and cost uncertainty methods
(SMM, MCS, contingency, sensitivity, and correlation matrix construction).
Section~\ref{sec-evm} describes earned value management.
Section~\ref{sec-learning} describes sigmoidal
learning curves. Section~\ref{sec-bayes} covers Bayesian risk inference and
updating. Section~\ref{sec-dsm} covers design structure matrices.
Section~\ref{sec-network} introduces the probabilistic
network module. Section~\ref{sec-agent} describes the Model Context Protocol
integration. Section~\ref{sec-casestudy} presents an integrated case study.
Section~\ref{sec-summary} concludes.


# Schedule and Cost Uncertainty {#sec-uncertainty}

Two complementary methods are provided for quantifying schedule and cost
uncertainty: the Second Moment Method for rapid analytical estimates, and
Monte Carlo simulation for full distributional analysis. Contingency
extraction and sensitivity analysis complete the workflow.

We illustrate the workflow with \code{building\_project}, a bundled example
dataset describing a mid-rise commercial building with six work packages,
adapted from the construction-project examples in \citet{Damnjanovic2020}.

```{r bp-load}
data("building_project", package = "PRA")
bp <- building_project
bp$task_names
```

## Second Moment Method

The Second Moment Method (SMM) propagates task-level means and variances to
project-level estimates using the Central Limit Theorem \citep{Benjamin2000}.
For a project with $n$ tasks:
$$E[X] = \sum_{i=1}^{n} E[X_i], \quad
  \text{Var}(X) = \sum_{i=1}^{n}\text{Var}(X_i) + 2\sum_{i<j}\rho_{ij}\sigma_i\sigma_j$$
The six task durations are triangular, so we obtain their means and variances
in closed form and propagate them with \code{smm()}:

```{r smm-example}
tri_mean <- function(d) (d$a + d$b + d$c) / 3
tri_var  <- function(d) (d$a^2 + d$b^2 + d$c^2 -
                         d$a * d$b - d$a * d$c - d$b * d$c) / 18
task_means <- vapply(bp$task_distributions, tri_mean, numeric(1))
task_vars  <- vapply(bp$task_distributions, tri_var,  numeric(1))

result <- smm(task_means, task_vars)   # independent tasks
cat("Total mean:", round(result$total_mean, 2), "weeks\n")
cat("Total SD:  ", round(result$total_std,  2), "weeks\n")
```

Positive correlation between tasks inflates the total variance; \code{smm()}
accepts a correlation matrix and adds the covariance terms analytically:

```{r smm-cor}
result_cor <- smm(task_means, task_vars, bp$cor_mat)
cat("Total SD (correlated):", round(result_cor$total_std, 2), "weeks\n")
```

SMM is useful when only first and second moments are available and a rapid
estimate is needed. When task distributions are non-normal or accurate tail
behavior is required, Monte Carlo simulation is preferred.

## Monte Carlo Simulation

\code{mcs()} propagates uncertainty through a project network by drawing
samples from user-specified task distributions, triangular, normal, or
uniform, over a large number of iterations \citep{Vose2008}. By default the
tasks are sampled independently, where an optional correlation matrix is supported,
though its current handling is approximate (Section~\ref{sec-summary}).

```{r mcs-run}
set.seed(42)
results <- mcs(10000, bp$task_distributions)
cat("Simulated mean:", round(results$total_mean, 2), "weeks\n")
cat("Simulated SD:  ", round(results$total_sd,   2), "weeks\n")
```

The \code{mcs()} function returns an S3 object of class \code{"mcs"} with
a \code{print()} method that summarizes percentiles. The full simulated
distribution is available in \code{results\$total\_distribution}.
On a modern laptop, 10,000 simulations of a six-task network complete in
under one second; increasing the count to 100,000 provides more precise tail
estimates at roughly ten-fold additional runtime. The default of 10,000
balances accuracy and speed for most practical analyses.

### Numerical Validation

For independent tasks the SMM estimates are the exact first two moments of the
total: the total mean equals the sum of the task means by linearity of
expectation, and the total variance equals the sum of the task variances by
additivity of variance under independence. (The Central Limit Theorem governs
only the asymptotic \emph{shape} of the total, not the equality of its
moments.) Monte Carlo simulation should therefore reproduce these moments, and
the two agree to within Monte Carlo error, validating the sampling engine:

```{r validation}
cat(sprintf("Mean:  SMM %.2f  vs  MCS %.2f\n",
            result$total_mean, results$total_mean))
cat(sprintf("SD:    SMM %.2f  vs  MCS %.2f\n",
            result$total_std,  results$total_sd))
```

```{r mcs-hist, fig.cap="Distribution of simulated total project durations. The dashed vertical line marks the mean."}
hist(results$total_distribution,
  breaks = 50, freq = FALSE,
  main = "MCS: Total Project Duration",
  xlab = "Duration (weeks)", col = "steelblue", border = "white"
)
lines(density(results$total_distribution), col = "tomato", lwd = 2)
abline(v = results$total_mean, lty = 2, lwd = 1.5)
```

## Correlation Matrix Construction

Both \code{smm()} and \code{mcs()} accept a correlation matrix to model
dependence between tasks. \code{cor\_matrix()} constructs a valid
positive-definite correlation matrix by empirically sampling from a set of
task distributions, providing a convenient way to derive the correlation
input from the same distributional assumptions used elsewhere in the workflow.

```{r cor-matrix}
dists <- list(
  normal  = function(n) rnorm(n, mean = 10, sd = 2),
  triang  = function(n) mc2d::rpert(n, min = 5, mode = 10, max = 15),
  uniform = function(n) runif(n, min = 8, max = 12)
)
C_emp <- cor_matrix(num_samples = 1000, num_vars = 3, dists = dists)
print(round(C_emp, 3))
```

## Contingency Analysis

\code{contingency()} extracts a reserve buffer as the difference between a
high-confidence percentile and the base (median) estimate, a standard
approach in project risk practice \citep{PMBOK2021}.

```{r contingency}
reserve <- contingency(results, phigh = 0.80, pbase = 0.50)
cat("Schedule contingency (P80 - P50):", round(reserve, 2), "weeks\n")
```

## Sensitivity Analysis

\code{sensitivity()} decomposes total project variance by task contribution,
producing the inputs for a Tornado chart. Each task's index is its own
variance plus its covariance with every other task, expressed as a
proportion of total project variance.

```{r sensitivity, fig.cap="Tornado chart: each bar shows a task's share of total project duration variance."}
sens <- sensitivity(bp$task_distributions)
names(sens) <- paste0("T", seq_along(sens))
barplot(sort(sens),
  horiz = TRUE, col = "steelblue",
  main = "Sensitivity: Task Contributions",
  xlab = "Proportion of Total Variance"
)
```

Tasks with larger bars are the dominant drivers of total risk; mitigation
resources should be directed there first.


# Earned Value Management {#sec-evm}

The EVM module implements the complete ANSI/EIA-748 suite of eleven
performance metrics \citep{Fleming2010}. The three core quantities, Planned
Value (PV), Earned Value (EV), and Actual Cost (AC), are computed first.
All performance indices and forecasts follow from them.

```{r evm-setup}
bac         <- 500000                          # budget at completion ($)
schedule    <- c(0.10, 0.25, 0.50, 0.75, 1.00) # cumulative value curve
time_period <- 3
```

```{r evm-core}
pv_val <- pv(bac, schedule, time_period)
ev_val <- ev(bac, actual_per_complete = 0.40)
ac_val <- ac(c(45000, 110000, 135000), time_period, cumulative = FALSE)

cat("PV: $", format(pv_val, big.mark = ","), "\n")
cat("EV: $", format(ev_val, big.mark = ","), "\n")
cat("AC: $", format(ac_val, big.mark = ","), "\n")
```

```{r evm-indices}
spi_val  <- spi(ev_val, pv_val)
cpi_val  <- cpi(ev_val, ac_val)
sv_val   <- sv(ev_val,  pv_val)
cv_val   <- cv(ev_val,  ac_val)

cat("SPI:", round(spi_val, 3), "  CPI:", round(cpi_val, 3), "\n")
cat("SV: $", format(round(sv_val), big.mark = ","),
    "  CV: $", format(round(cv_val), big.mark = ","), "\n")
```

Both SPI $< 1$ and CPI $< 1$ indicate the project is behind schedule and over
budget. The \code{eac()} function provides three completion forecasts under
different efficiency assumptions \citep{Fleming2010}:
$$\text{EAC}_{\text{typical}} = \frac{\text{BAC}}{\text{CPI}}, \quad
  \text{EAC}_{\text{atypical}} = \text{AC} + (\text{BAC} - \text{EV}), \quad
  \text{EAC}_{\text{combined}} = \text{AC} +
  \frac{\text{BAC} - \text{EV}}{\text{CPI} \times \text{SPI}}$$
The typical method assumes remaining work proceeds at the cumulative cost
efficiency to date; the atypical method assumes the overrun to date was a
one-off and future work runs on plan; the combined method deflates the
remaining budget by both cost and schedule performance.

```{r evm-eac}
cat("EAC (typical):  $", format(round(eac(bac, method = "typical",
    cpi = cpi_val)), big.mark = ","), "\n")
cat("EAC (atypical): $", format(round(eac(bac, method = "atypical",
    ac = ac_val, ev = ev_val)), big.mark = ","), "\n")
cat("EAC (combined): $", format(round(eac(bac, method = "combined",
    cpi = cpi_val, ac = ac_val, ev = ev_val, spi = spi_val)),
    big.mark = ","), "\n")
```

The remaining EVM metrics are:

```{r evm-remaining}
eac_typ <- eac(bac, method = "typical", cpi = cpi_val)
cat("ETC:  $", format(round(etc(bac, ev_val, cpi = cpi_val)), big.mark = ","), "\n")
cat("TCPI:", round(tcpi(bac, ev_val, ac_val), 3), "\n")
cat("VAC:  $", format(round(vac(bac, eac_typ)), big.mark = ","), "\n")
```

The To-Complete Performance Index (TCPI) reports the cost efficiency required
on all remaining work to finish within the original budget; values substantially
above 1.0 signal an unrealistic recovery target.


# Learning Curves {#sec-learning}

Sigmoidal (S-curve) models capture the characteristic pattern of project
progress: slow start, rapid acceleration, and plateau approaching 100\%
completion. \pkg{PRA} provides three model types, Logistic, Pearl, and
Gompertz, fitted via the Levenberg--Marquardt nonlinear least-squares
algorithm in \pkg{minpack.lm} \citep{Elzhov2016}.

```{r sigmoidal-data}
progress <- data.frame(
  time       = 1:9,
  completion = c(5, 15, 40, 60, 70, 75, 80, 85, 90)
)
```

```{r sigmoidal-fit}
fit_log  <- fit_sigmoidal(progress, "time", "completion", "logistic")
fit_gomp <- fit_sigmoidal(progress, "time", "completion", "gompertz")

cat("Logistic RSE: ", round(summary(fit_log)$sigma,  2), "\n")
cat("Gompertz RSE: ", round(summary(fit_gomp)$sigma, 2), "\n")
```

```{r sigmoidal-plot, fig.cap="Logistic learning curve with 95\\% confidence band. The shaded region widens as extrapolation extends beyond the observed data range."}
plot_sigmoidal(fit_log, progress, "time", "completion", "logistic",
               conf_level = 0.95,
               main = "Logistic Learning Curve",
               xlab = "Week", ylab = "Completion (%)")
```

\code{predict\_sigmoidal()} generates forecasts with confidence bounds at
arbitrary future time points:

```{r sigmoidal-predict}
preds <- predict_sigmoidal(fit_log, seq(10, 12, by = 0.5),
                            "logistic", conf_level = 0.95)
print(round(preds, 1))
```


# Bayesian Risk Inference {#sec-bayes}

Project risks are frequently triggered by one or more underlying root causes
whose presence is uncertain. \pkg{PRA} models this with a noisy-OR risk network
\citep{Pearl2009}: each root cause $C_i$ acts on a risk event $R$ independently,
and $R$ occurs if it is triggered by at least one cause. We illustrate with the
schedule-delay risk in \code{building\_project}, which has two candidate causes.

## Prior Risk Probability

\code{risk\_prob()} first computes each cause's marginal contribution by the law
of total probability,
$$m_i = P(R \mid C_i)\,P(C_i) + P(R \mid \bar{C}_i)\,\bigl(1 - P(C_i)\bigr),$$
and then combines the independent causes with a noisy-OR,
$$P(R) = 1 - \prod_{i=1}^{k}\bigl(1 - m_i\bigr),$$
which always lies in $[0, 1]$ and reduces to the single-cause marginal when
$k = 1$.

```{r bayes-prior}
bp <- building_project
prior_delay <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                         bp$risks_given_not_causes)
cat("Prior P(schedule delay):", round(prior_delay, 3), "\n")
```

## Posterior Risk Probability

When some causes are subsequently observed, for example, a site inspection
confirms one cause is present while the other remains unknown, \code{risk\_post\_prob()}
updates the estimate. An observed cause is fixed to its conditional probability,
while an unobserved cause (coded \code{NA}) retains its marginal contribution
$m_i$; the causes are recombined with the same noisy-OR, so prior and posterior
share a common scale:
$$P(R \mid \text{obs}) = 1 - \prod_{i=1}^{k}\bigl(1 - c_i\bigr), \qquad
  c_i = \begin{cases}
    P(R \mid C_i)       & C_i \text{ observed present},\\
    P(R \mid \bar{C}_i) & C_i \text{ observed absent},\\
    m_i                 & C_i \text{ unobserved}.
  \end{cases}$$

```{r bayes-post}
post_delay <- risk_post_prob(bp$cause_probs, bp$risks_given_causes,
                             bp$risks_given_not_causes, bp$observed_causes)
cat("Posterior P(schedule delay):", round(post_delay, 3),
    " (observed causes:", paste(bp$observed_causes, collapse = ", "), ")\n")
```

Observing an aggravating cause raises the posterior risk probability; observing
a mitigating one lowers it. This closed-form update is inexpensive and
complements the sample-based conditioning of the probabilistic network module
(Section~\ref{sec-network}), which operates on full cost distributions rather
than event probabilities.


# Dependency Structure Analysis {#sec-dsm}

Building on \citet{Govan2016, Govan2020} and the foundational DSM
literature \citep{Steward1981, Browning2001}, the DSM module quantifies
structural coupling between project tasks through shared resources and risks.

## Design Structure Matrices

The **Parent DSM** is derived from the Resource--Task matrix $\mathbf{S}$
(rows = resources, columns = tasks) as $\mathbf{P} = \mathbf{S}^\top\mathbf{S}$.
Off-diagonal entry $P_{jk}$ counts the number of resources shared between
tasks $j$ and $k$, the structural pathway through which disruptions propagate
\citep{Govan2016}.

```{r dsm-S}
S <- matrix(c(
  1, 0, 1, 0,
  1, 1, 0, 0,
  0, 1, 0, 1,
  0, 0, 1, 1,
  0, 1, 1, 0
), nrow = 4, ncol = 5)
rownames(S) <- paste0("R", 1:4)
colnames(S) <- paste0("T", 1:5)
```

```{r parent-dsm, fig.cap="Parent DSM heatmap. Darker cells indicate more shared resources and stronger coupling between task pairs."}
p <- parent_dsm(S)
plot(p)
```

The **Grandparent DSM** adds a risk layer via the Risk--Resource matrix
$\mathbf{R}$, tracing dependencies from risks through resources to tasks.
Off-diagonal entry $G_{jk}$ counts the number of risks shared between
task $j$ and task $k$ through the resource chain, a direct operationalization
of the structural network measures developed in \citet{Govan2020}.

```{r dsm-R}
R_mat <- matrix(c(
  1, 0, 1,
  1, 1, 0,
  0, 1, 0,
  0, 0, 1
), nrow = 3, ncol = 4)
rownames(R_mat) <- paste0("Risk", 1:3)
colnames(R_mat) <- paste0("R", 1:4)
```

```{r grandparent-dsm, fig.cap="Grandparent DSM heatmap. Shows risk-based coupling: task pairs that share risk exposure through common resources tend to fail together."}
g <- grandparent_dsm(S, R_mat)
plot(g)
```

Task pairs with high off-diagonal values in the Grandparent DSM represent
concentrated risk exposure; these pairs are prime candidates for contingency
buffering or resource decoupling.


# Probabilistic Networks {#sec-network}

This module is the secondary contribution of \pkg{PRA}: an
open-source implementation, to the author's knowledge the first, of the
causal-network methods for project risk introduced by \citet{Govan2014} and
\citet{Govan2016, Govan2020}.
The probabilistic network module extends DSM structural
analysis by attaching probability distributions to each node, enabling
forward simulation of cost uncertainty and Bayesian conditioning on
observed evidence. A probabilistic network is a directed acyclic graph in
which nodes represent project variables (risks, resources, tasks, and the
project total) and edges encode conditional dependencies
\citep{Pearl2009, Govan2014}. Where a DSM counts shared resources between
task pairs, a probabilistic network propagates distributions through the
causal chain from risks to total project cost.

## Building a Network

\code{prob\_net()} constructs the network from a node data frame, a links
data frame, and a named list of distributions. Distributions may be
\code{normal}, \code{lognormal}, \code{uniform}, \code{discrete}, or
\code{conditional} (switching between two sub-distributions depending on
a discrete parent node); \code{aggregate} nodes sum their parents.

```{r net-setup}
nodes <- data.frame(
  id = c("A","B","C","D","E","F","G","H","I"),
  label = c("Risk-1","Risk-2","Resource-1","Resource-2","Resource-3",
            "Task-1","Task-2","Task-3","Project"),
  stringsAsFactors = FALSE
)
links <- data.frame(
  source = c("A","B","C","D","E","F","G","H"),
  target = c("C","D","F","G","H","I","I","I"),
  stringsAsFactors = FALSE
)
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)),
  D = list(type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = "C"),
  G = list(type = "aggregate", nodes = "D"),
  H = list(type = "aggregate", nodes = "E"),
  I = list(type = "aggregate", nodes = c("F","G","H"))
)
net <- prob_net(nodes, links, distributions = distributions)
```

## Forward Simulation

\code{prob\_net\_sim()} draws samples from every node in topological order
and returns a data frame with one column per node.

```{r net-sim}
sim_net <- prob_net_sim(net, num_samples = 10000)
cat("Mean project cost: $", format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("P80 project cost:  $", format(round(quantile(sim_net$I, 0.80)),
    big.mark = ","), "\n")
```

## Learning: Conditioning on Evidence

\code{prob\_net\_learn()} clamps observed nodes to fixed values and
re-simulates the downstream distribution, an observational update in the
sense of \citet{Pearl2009}. The following call fixes Risk-2 (node B) to
absent (value 0), ruling out Technical Complexity as a concern:

```{r net-learn, fig.cap="Total project cost before and after observing that Risk-2 (Technical Complexity) did not occur. Conditioning on Risk-2 = 0 eliminates the heavy upper tail and shifts the distribution leftward by roughly \\$18,000 in expectation."}
learn_net <- prob_net_learn(net, observations = list(B = 0),
                            num_samples = 10000)

prior_net_dens <- density(sim_net$I)
post_net_dens  <- density(learn_net$I)

plot(prior_net_dens, col = "steelblue", lwd = 2,
     main = "Total Project Cost: Prior vs. Posterior",
     xlab = "Cost ($)",
     xlim = range(c(sim_net$I, learn_net$I)),
     ylim = range(c(prior_net_dens$y, post_net_dens$y)))
lines(post_net_dens, col = "tomato", lwd = 2)
legend("topright",
  legend = c("Prior (Risk-2 uncertain)", "Posterior (Risk-2 = 0)"),
  col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

## Intervention: The do-Operator

Conditioning answers "what if we \emph{observe} that a risk did not occur."
A distinct question is "what if we \emph{act} to eliminate it."
\code{prob\_net\_update()} performs structural surgery on the
network, adding or removing edges and replacing node distributions, which
corresponds to the \emph{do}-operator in Pearl's causal calculus
\citep{Pearl2009}, as opposed to \code{prob\_net\_learn()}, which conditions
without altering the graph. Suppose management funds a mitigation that
decouples Resource-2 from Risk-2: remove the Risk-2 $\to$ Resource-2 edge and
fix Resource-2 to its baseline cost.

```{r net-do}
inter_net <- prob_net_update(net,
  remove_links = data.frame(source = "B", target = "D"),
  update_distributions = list(
    D = list(type = "normal", mean = 50000, sd = 10000)
  )
)
do_net <- prob_net_sim(inter_net, num_samples = 10000)

cat("Prior mean:            $", format(round(mean(sim_net$I)),   big.mark = ","), "\n")
cat("Seeing (Risk-2 = 0):   $", format(round(mean(learn_net$I)), big.mark = ","), "\n")
cat("Doing  (do(Risk-2=0)): $", format(round(mean(do_net$I)),    big.mark = ","), "\n")
```

Here \emph{seeing} (\code{prob\_net\_learn()}) and \emph{doing}
(\code{prob\_net\_update()}) coincide because Risk-2 is an independent root
cause with no shared upstream confounder, so conditioning on it and
intervening on it induce the same downstream distribution. The two diverge,
however, when a risk shares a common cause with project cost, as the next
example shows.

## When Seeing and Doing Diverge

Suppose a latent common cause, adverse site conditions, drives both a rework
risk and, through a separate pathway, the foundation cost. Observing that rework
did \emph{not} occur is evidence that site conditions were favorable, which in
turn lowers the expected foundation cost; \emph{intervening} to prevent rework
severs the risk but leaves site conditions, and hence the foundation cost,
untouched. The network below encodes this structure: the common cause \code{U}
feeds both the rework risk \code{Rk} and the foundation cost \code{Wf}, while
the schedule cost \code{Vs} depends only on \code{Rk}.

```{r net-confound}
set.seed(42)
cf_nodes <- data.frame(
  id = c("U", "Rk", "Wf", "Vs", "Tc"), stringsAsFactors = FALSE)
cf_links <- data.frame(
  source = c("U", "U", "Rk", "Wf", "Vs"),
  target = c("Rk", "Wf", "Vs", "Tc", "Tc"), stringsAsFactors = FALSE)
cf_dists <- list(
  U  = list(type = "discrete", values = c(1, 0), probs = c(0.5, 0.5)),
  Rk = list(type = "conditional", condition = "U",
    true_dist  = list(type = "discrete", values = c(1, 0), probs = c(0.9, 0.1)),
    false_dist = list(type = "discrete", values = c(1, 0), probs = c(0.1, 0.9))),
  Wf = list(type = "conditional", condition = "U",
    true_dist  = list(type = "normal", mean = 60000, sd = 5000),
    false_dist = list(type = "normal", mean = 20000, sd = 5000)),
  Vs = list(type = "conditional", condition = "Rk",
    true_dist  = list(type = "normal", mean = 40000, sd = 5000),
    false_dist = list(type = "normal", mean = 10000, sd = 5000)),
  Tc = list(type = "aggregate", nodes = c("Wf", "Vs")))
cf_net <- prob_net(cf_nodes, cf_links, distributions = cf_dists)

cf_prior <- prob_net_sim(cf_net, num_samples = 10000)
cf_see   <- prob_net_learn(cf_net, observations = list(Rk = 0),
                           num_samples = 10000)
cf_do    <- prob_net_sim(prob_net_update(cf_net,
  remove_links = data.frame(source = "U", target = "Rk"),
  update_distributions = list(
    Rk = list(type = "discrete", values = c(1, 0), probs = c(0, 1)))),
  num_samples = 10000)

cat("Prior mean total:        $", format(round(mean(cf_prior$Tc)), big.mark = ","), "\n")
cat("Seeing (Rk = 0) mean:    $", format(round(mean(cf_see$Tc)),   big.mark = ","),
    "  E[U | Rk=0] =", round(mean(cf_see$U), 2), "\n")
cat("Doing  do(Rk = 0) mean:  $", format(round(mean(cf_do$Tc)),    big.mark = ","),
    "  E[U] =", round(mean(cf_do$U), 2), "\n")
```

Conditioning propagates the evidence back through the shared cause, the
expected site-condition indicator falls from its prior of 0.5 to about 0.1, so
\emph{seeing} that rework did not occur pulls the foundation cost down as well
and yields a markedly lower expected total than \emph{doing}. Because
\code{prob\_net\_learn()} conditions by rejection sampling over whole joint
draws, it captures this backdoor propagation that a clamp-and-forward update
cannot; this is precisely where observation and intervention must be
distinguished, and where the \emph{do}-operator becomes indispensable.

## Ranking Risks by Intervention

Repeating the intervention for each root risk and measuring the reduction in
project-cost variance yields a principled, network-level mitigation
ranking, a causal analog of the Tornado chart of
Section~\ref{sec-uncertainty}:

```{r net-importance, fig.cap="Risk importance for the probabilistic network: the reduction in total project-cost variance achieved by intervening to eliminate each root risk. Taller bars indicate higher-priority mitigation targets."}
base_var <- var(sim_net$I)

do_A <- prob_net_sim(prob_net_update(net,
  remove_links = data.frame(source = "A", target = "C"),
  update_distributions = list(C = list(type = "normal", mean = 15000, sd = 3000))),
  num_samples = 10000)

importance <- c(
  "Risk-1" = base_var - var(do_A$I),
  "Risk-2" = base_var - var(do_net$I)
)
barplot(sort(importance), horiz = TRUE, col = "steelblue",
  main = "Risk Importance: Variance Eliminated by do(risk = 0)",
  xlab = "Reduction in Project-Cost Variance")
```

The probabilistic network module is newer than the analytical core, and its
API may still evolve in future releases.


# Model Context Protocol Integration {#sec-agent}

Version 0.5.0 of \pkg{PRA} exposes all analytical functions as tools callable
by large-language-model (LLM) agents through the Model Context Protocol (MCP)
\citep{MCP2024}. \code{pra\_mcp\_server()} starts a server that advertises each
tool, Monte Carlo simulation, EVM, Bayesian risk, learning curves, DSM, and
the rest, so that Claude, Claude Code, and other MCP-compatible clients can
invoke them as native tools. The server is registered in Claude Code with a
single shell command:

```{r mcp-code, eval=FALSE, engine="bash"}
claude mcp add -s project pra -- Rscript -e "PRA::pra_mcp_server()"
```

For Claude Desktop, add the following block to the MCP configuration file:

\begin{verbatim}
{
  "mcpServers": {
    "pra": { "command": "Rscript",
             "args": ["-e", "PRA::pra_mcp_server()"] }
  }
}
\end{verbatim}

The tool definitions are built once with \code{ellmer::tool()} \citep{ellmer}
and served by \pkg{mcptools}. Each tool validates its JSON inputs and returns a
plain-text summary for the LLM together with, where applicable, a rich HTML
result with inline plots for clients that render it.

To help an agent decide \emph{when} and \emph{how} to call each tool,
\pkg{PRA} bundles a set of Agent Skills under
\code{system.file("skills", package = "PRA")}. Each skill is a \code{SKILL.md}
file with structured guidance, when to use the method, which tool to call, the
JSON argument format, and how to interpret the result, covering Monte Carlo
simulation, sensitivity and contingency, EVM, Bayesian risk, learning curves,
and dependency structure analysis, plus a routing overview. Pairing the skills
with the MCP server lets an agent carry out a complete project risk workflow
(for example, \code{mcs\_tool} followed by \code{contingency\_tool} and
\code{sensitivity\_tool}) from a natural-language request, without embedding an
LLM in the package itself.

The MCP layer is strictly optional: every analytical function operates without
any optional dependency. Only \code{pra\_mcp\_server()} requires \pkg{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}.


# Illustrative Case Study {#sec-casestudy}

This section demonstrates an integrated, end-to-end project risk analysis
for the \code{building\_project} dataset introduced in
Section~\ref{sec-uncertainty}, progressing through uncertainty
quantification, risk-driver ranking, performance monitoring, completion
forecasting, Bayesian risk updating, dependency structure analysis, and
network cost propagation in a single \proglang{R} session. The workflow
reuses objects built in the preceding sections (for example the closed-form
task moments, the fitted learning curve, and the probabilistic network),
illustrating how the modules compose within one analysis.

## Step 1: Characterize Task Uncertainty

Combine the rapid analytical estimate from the Second Moment Method, reusing
the closed-form triangular moments \code{task\_means} and \code{task\_vars}
from Section~\ref{sec-uncertainty} with the dataset correlation matrix, with
the full simulated distribution from Monte Carlo simulation:

```{r cs-mcs}
bp <- building_project
set.seed(42)

smm_cs <- smm(task_means, task_vars, bp$cor_mat)   # analytic, correlated
sim     <- mcs(10000, bp$task_distributions)        # full distribution
res     <- contingency(sim, phigh = 0.80, pbase = 0.50)

cat("SMM total (correlated): mean", round(smm_cs$total_mean, 1),
    "SD", round(smm_cs$total_std, 2), "weeks\n")
cat("P50 schedule:", round(quantile(sim$total_distribution, 0.50), 1), "weeks\n")
cat("P80 schedule:", round(quantile(sim$total_distribution, 0.80), 1), "weeks\n")
cat("Contingency (P80-P50):", round(res, 2), "weeks\n")
```

## Step 2: Rank Risk Drivers

\code{sensitivity()} ranks the tasks by their share of total schedule
variance, focusing mitigation on the dominant drivers:

```{r cs-sens}
drivers <- sensitivity(bp$task_distributions)
names(drivers) <- bp$task_names
print(round(sort(drivers, decreasing = TRUE), 3))
```

## Step 3: Monitor Earned Value at Mid-Project

```{r cs-evm}
pv_cs <- pv(bp$bac, bp$schedule, bp$time_period)
ev_cs <- ev(bp$bac, bp$actual_per_complete)
ac_cs <- ac(bp$actual_costs, bp$time_period, cumulative = FALSE)

cat("SPI:", round(spi(ev_cs, pv_cs), 3),
    " CPI:", round(cpi(ev_cs, ac_cs), 3), "\n")
cat("EAC (typical): $", format(round(eac(bp$bac,
    method = "typical", cpi = cpi(ev_cs, ac_cs))), big.mark = ","), "\n")
```

## Step 4: Forecast Completion

Reusing the logistic learning curve \code{fit\_log} fitted in
Section~\ref{sec-learning}, project percent-complete at future weeks with
confidence bounds:

```{r cs-forecast}
preds_cs <- predict_sigmoidal(fit_log, seq(10, 12, by = 1),
                              "logistic", conf_level = 0.95)
print(round(preds_cs, 1))
```

## Step 5: Update Risk Probability After Site Inspection

```{r cs-bayes}
prior_r <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                     bp$risks_given_not_causes)
post_r  <- risk_post_prob(bp$cause_probs, bp$risks_given_causes,
                          bp$risks_given_not_causes, bp$observed_causes)
cat("Prior P(delay):", round(prior_r, 3),
    "-> Posterior P(delay):", round(post_r, 3), "\n")
```

## Step 6: Identify Coupled Tasks via DSM

```{r cs-dsm, fig.cap="Grandparent DSM for the case-study project, showing which task pairs share the most risk exposure through common resources."}
g_cs <- grandparent_dsm(bp$resource_task, bp$risk_resource)
plot(g_cs)
```

The Grandparent DSM reveals which task pairs share risk exposure through
common resources, informing targeted contingency allocation and resource
decoupling decisions.

## Step 7: Propagate Cost Through the Network

Finally, use the probabilistic network \code{net} from
Section~\ref{sec-network} to propagate cost uncertainty to the project total,
then condition on the observation that Risk-2 (Technical Complexity) did not
occur:

```{r cs-net}
cat("Prior mean project cost: $",
    format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("Prior P80 project cost:  $",
    format(round(quantile(sim_net$I, 0.80)), big.mark = ","), "\n")
cat("Posterior mean (Risk-2 = 0): $",
    format(round(mean(learn_net$I)), big.mark = ","), "\n")
```

Conditioning on the resolved risk lowers the expected project cost, closing
the loop from uncertainty characterization to evidence-based updating.


# Summary {#sec-summary}

\pkg{PRA} is a CRAN-available \proglang{R} package whose primary contribution
is being the first of its kind to unify the foundational methods of quantitative
project risk analysis in a single, open-source framework of ten integrated
analytical modules: SMM, MCS, contingency analysis, sensitivity analysis,
EVM (11 metrics), Bayesian risk inference and updating, cost risk
distributions, sigmoidal learning curves, correlation matrix construction, and
design structure matrices. As a secondary contribution, an
eleventh module delivers an open-source implementation, to the author's
knowledge the first, of the causal-network and probabilistic-network methods
for project risk developed in \citet{Govan2014} and \citet{Govan2016, Govan2020},
with forward simulation and Bayesian conditioning.

As a third contribution, version 0.5.0 exposes every analytical function as a
tool over the Model Context Protocol, complemented by a bundled set of Agent
Skills, so that LLM agents can drive the package without any embedded
model, the first project risk analysis package to do so. All exported
functions adhere to rOpenSci SRR standards, with a test suite verifying
correctness, edge-case handling, and numerical stability across three
operating systems. The package is licensed under MIT and the source is
available at \url{https://github.com/paulgovan/PRA}; documentation and
a companion online book are at \url{https://paulgovan.github.io/PRA/}.

## Limitations and Future Work

Several limitations apply. First, \pkg{PRA} currently propagates uncertainty at the task level and does not yet model schedule logic, critical path, activity precedence, and resource leveling. Full schedule-network risk analysis, in which schedule Monte Carlo runs over an activity-precedence network, such as Program Evaluation and Review Technique / Critical Path Method (PERT/CPM) to identify the critical path and criticality indices under uncertainty, is a planned future direction that would extend \pkg{PRA} from task-level to network-level schedule risk. Second, \code{mcs()} induces task correlation by
applying a Cholesky factor to the sampled values, which reproduces the target
rank correlation but distorts the marginal means; the examples in this article
therefore treat tasks as independent for Monte Carlo simulation and use
\code{smm()}, which handles correlation analytically, when correlation matters.
A copula-based implementation that preserves the marginals is planned. Third,
the probabilistic network module is newer than the analytical core: its API may
still evolve, and it does not yet provide convergence diagnostics for the Monte
Carlo estimators. Fourth,
the package currently accepts data as \proglang{R} vectors and lists; direct
import from project management file formats (Microsoft Project, Primavera P6,
CSV exports from Jira or Asana) is not supported and is planned for a future
release.

# Computational details {-}

The results in this article were produced using \proglang{R} \citep{RCore}
with the \pkg{PRA} package (version 0.5.0) and its hard dependencies
\pkg{mc2d} \citep{mc2d} and \pkg{minpack.lm} \citep{Elzhov2016}. The optional
Model Context Protocol server additionally uses \pkg{ellmer} \citep{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}. \proglang{R} itself and all packages used
are available from the CRAN at \url{https://CRAN.R-project.org/}. A random seed of 42 is set at the start of the session so that the simulation output is reproducible.

# References {-}
