---
documentclass: jss
author:
  - name: Paul Govan
    orcid: 0000-0002-1821-8492
    affiliation: Independent Researcher
    email: \email{paul.govan2@gmail.com}
    url: https://paulgovan.github.io/PRA/
title:
  formatted: "\\pkg{PRA}: Quantitative Project Risk Analysis in \\proglang{R}"
  plain:     "PRA: Quantitative Project Risk Analysis in R"
  short:     "\\pkg{PRA}: Project Risk Analysis"
abstract: >
  The \pkg{PRA} package for \proglang{R} provides a unified, open-source
  implementation of the core analytical methods used in quantitative project
  risk analysis: the Second Moment Method (SMM), Monte Carlo simulation (MCS),
  earned value management (EVM), contingency and sensitivity analysis,
  Bayesian risk inference and updating, sigmoidal learning curves,
  and design structure matrices (DSM). It operationalizes causal-network and
  structural-network frameworks for project risk in a dependency-light
  \proglang{R} package available on CRAN. Version 0.5.0 further adds an
  experimental probabilistic network module for forward simulation and
  Bayesian conditioning along causal chains, and exposes every analytical
  function as a tool over the Model Context Protocol (MCP), so that
  large-language-model agents can run project risk analyses through a
  documented set of skills. The package adheres to rOpenSci Statistical
  Software Review standards and is supported by a continuous-integration test
  suite running across macOS, Windows, and Ubuntu.
keywords:
  formatted: [project risk analysis, Monte Carlo simulation, earned value management,
              "Bayesian inference", "design structure matrix", "\\proglang{R}"]
  plain:     [project risk analysis, Monte Carlo simulation, earned value management,
              Bayesian inference, design structure matrix, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: pra-jss.bib
---

```{r setup, include=FALSE}
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center",
                      fig.width = 5, fig.height = 3.5,
                      out.width = "80%")
set.seed(42)
library(PRA)
```

# Introduction

Quantitative risk analysis is a foundational discipline in project management,
enabling organizations to assess schedule and cost uncertainty, identify
high-impact risk drivers, and set evidence-based contingency reserves.
Govan (2014) and Govan and Damnjanovic (2016) introduced the first
implementation of causal networks for project risk analysis grounded in the
resource-based view, demonstrating that organizational capabilities and
shared resources---not just schedule uncertainty---are the structural
pathways through which risks propagate across project tasks.
Extending this work, Govan and Damnjanovic (2020) developed structural
network measures that quantify the topology of project risk networks,
enabling systematic identification of high-centrality risk nodes.

Despite broad industry adoption of quantitative methods, open-source tooling
for integrated project risk analysis in \proglang{R} remains fragmented.
Proprietary platforms such as Oracle Primavera Risk Analysis and Palisade
\@RISK provide integrated workflows but impose licensing costs and limit
reproducibility. Within \proglang{R}, no single package unifies Monte Carlo
schedule risk, earned value management, Bayesian risk inference, sigmoidal
learning curves, and dependency structure analysis under a common API.
The \pkg{mc2d} package \citep{mc2d} provides general-purpose two-dimensional
Monte Carlo simulation but has no project-specific constructs.
The \pkg{sensitivity} package \citep{sensitivity} implements variance-based
indices for general functions but does not integrate with schedule or cost
models. EVM metrics are absent from CRAN entirely. Bayesian packages such as
\pkg{rstan} \citep{rstan} and \pkg{brms} \citep{brms} operate at a high level
of generality and require substantial expertise to apply in project risk
contexts.

\pkg{PRA} addresses this gap. It provides a coherent, CRAN-validated framework
of ten analytical modules covering the full quantitative project risk
workflow---from uncertainty characterization and simulation to performance
monitoring and Bayesian updating---while operationalizing the causal and
network-theoretic frameworks of Govan (2014) and Govan and Damnjanovic
(2016, 2020) computationally. The package targets project analysts who work
in \proglang{R} and require reproducible, auditable outputs compatible with
standard reporting pipelines. It adheres to rOpenSci Statistical Software
Review (SRR) standards \citep{rOpenSciStats2021} and uses minimal hard
dependencies: \pkg{mc2d} \citep{mc2d} for triangular distribution families
and \pkg{minpack.lm} \citep{Elzhov2016} for nonlinear least squares fitting.
The optional Model Context Protocol (MCP) server activates through
\code{Suggests}-only dependencies.

The following table summarizes how \pkg{PRA} compares with the most
closely related \proglang{R} packages and the dominant commercial
platform.

```{r comparison-table, echo=FALSE}
comp <- data.frame(
  Feature = c(
    "MCS schedule risk", "EVM (11 metrics)", "Bayesian updating",
    "Sigmoidal learning curves", "Design structure matrices",
    "Probabilistic networks", "MCP tool server",
    "Open-source (CRAN)", "SRR-compliant"
  ),
  PRA         = c("Yes","Yes","Yes","Yes","Yes","Exp.","Yes","Yes","Yes"),
  mc2d        = c("Partial","---","---","---","---","---","---","Yes","---"),
  sensitivity = c("---","---","---","---","---","---","---","Yes","---"),
  `rstan/brms`= c("---","---","Yes","---","---","Yes","---","Yes","---"),
  `@RISK`     = c("Yes","Yes","---","---","---","---","---","---","---"),
  check.names = FALSE
)
knitr::kable(
  comp,
  caption = paste0("Comparison of \\pkg{PRA} with related \\proglang{R} packages",
    " and commercial tools. ``Exp.'' = experimental feature;",
    " ``---'' = not supported."),
  booktabs = TRUE, align = "lccccc"
)
```

The package is installed from CRAN in the standard way:

```{r install, eval=FALSE}
install.packages("PRA")
library("PRA")
```

Optional dependencies---\pkg{ellmer} \citep{ellmer}, \pkg{mcptools}, and
\pkg{jsonlite}---are only required for the MCP server, \code{pra\_mcp\_server()}.

The remainder of this article describes each analytical module in turn,
with working code examples drawn from the same \proglang{R} session.
Section~\ref{sec-uncertainty} covers schedule and cost uncertainty methods
(SMM, MCS, contingency, sensitivity). Section~\ref{sec-evm} describes
earned value management. Section~\ref{sec-bayes} covers Bayesian risk
inference and updating. Section~\ref{sec-learning} describes sigmoidal
learning curves. Section~\ref{sec-dsm} covers design structure matrices.
Section~\ref{sec-network} introduces the experimental probabilistic
network module. Section~\ref{sec-agent} describes the Model Context Protocol
integration. Section~\ref{sec-casestudy} presents an integrated case study.
Section~\ref{sec-summary} concludes.


# Schedule and Cost Uncertainty {#sec-uncertainty}

Two complementary methods are provided for quantifying schedule and cost
uncertainty: the Second Moment Method for rapid analytical estimates, and
Monte Carlo simulation for full distributional analysis. Contingency
extraction and sensitivity analysis complete the workflow.

We illustrate the workflow with \code{building\_project}, a bundled example
dataset describing a mid-rise commercial building with six work packages,
adapted from the construction-project examples in \citet{Damnjanovic2020}.

```{r bp-load}
data("building_project", package = "PRA")
bp <- building_project
bp$task_names
```

## Second Moment Method

The Second Moment Method (SMM) propagates task-level means and variances to
project-level estimates using the Central Limit Theorem \citep{Benjamin2000}.
For a project with $n$ tasks:
$$E[X] = \sum_{i=1}^{n} E[X_i], \quad
  \text{Var}(X) = \sum_{i=1}^{n}\text{Var}(X_i) + 2\sum_{i<j}\rho_{ij}\sigma_i\sigma_j$$
The six task durations are triangular, so we obtain their means and variances
in closed form and propagate them with \code{smm()}:

```{r smm-example}
tri_mean <- function(d) (d$a + d$b + d$c) / 3
tri_var  <- function(d) (d$a^2 + d$b^2 + d$c^2 -
                         d$a * d$b - d$a * d$c - d$b * d$c) / 18
task_means <- vapply(bp$task_distributions, tri_mean, numeric(1))
task_vars  <- vapply(bp$task_distributions, tri_var,  numeric(1))

result <- smm(task_means, task_vars)   # independent tasks
cat("Total mean:", round(result$total_mean, 2), "weeks\n")
cat("Total SD:  ", round(result$total_std,  2), "weeks\n")
```

Positive correlation between tasks inflates the total variance; \code{smm()}
accepts a correlation matrix and adds the covariance terms analytically:

```{r smm-cor}
result_cor <- smm(task_means, task_vars, bp$cor_mat)
cat("Total SD (correlated):", round(result_cor$total_std, 2), "weeks\n")
```

SMM is useful when only first and second moments are available and a rapid
estimate is needed. When task distributions are non-normal or accurate tail
behavior is required, Monte Carlo simulation is preferred.

## Monte Carlo Simulation

\code{mcs()} propagates uncertainty through a project network by drawing
samples from user-specified task distributions---triangular, normal, or
uniform---over a large number of iterations \citep{Vose2008}. By default the
tasks are sampled independently; an optional correlation matrix is supported,
though its current handling is approximate (Section~\ref{sec-summary}).

```{r mcs-run}
set.seed(42)
results <- mcs(10000, bp$task_distributions)
cat("Simulated mean:", round(results$total_mean, 2), "weeks\n")
cat("Simulated SD:  ", round(results$total_sd,   2), "weeks\n")
```

The \code{mcs()} function returns an S3 object of class \code{"mcs"} with
a \code{print()} method that summarizes percentiles. The full simulated
distribution is available in \code{results\$total\_distribution}.
On a modern laptop, 10,000 simulations of a six-task network complete in
under one second; increasing the count to 100,000 provides more precise tail
estimates at roughly ten-fold additional runtime. The default of 10,000
balances accuracy and speed for most practical analyses.

### Numerical Validation

For independent tasks, the Central Limit Theorem makes the SMM estimates the
exact moments of the total, so the Monte Carlo simulation should reproduce
them. The two agree to within Monte Carlo error, validating the sampling
engine:

```{r validation}
cat(sprintf("Mean:  SMM %.2f  vs  MCS %.2f\n",
            result$total_mean, results$total_mean))
cat(sprintf("SD:    SMM %.2f  vs  MCS %.2f\n",
            result$total_std,  results$total_sd))
```

```{r mcs-hist, fig.cap="Distribution of simulated total project durations. The dashed vertical line marks the mean."}
hist(results$total_distribution,
  breaks = 50, freq = FALSE,
  main = "MCS: Total Project Duration",
  xlab = "Duration (weeks)", col = "steelblue", border = "white"
)
lines(density(results$total_distribution), col = "tomato", lwd = 2)
abline(v = results$total_mean, lty = 2, lwd = 1.5)
```

## Contingency Analysis

\code{contingency()} extracts a reserve buffer as the difference between a
high-confidence percentile and the base (median) estimate---a standard
approach in project risk practice \citep{PMBOK2021}.

```{r contingency}
reserve <- contingency(results, phigh = 0.80, pbase = 0.50)
cat("Schedule contingency (P80 - P50):", round(reserve, 2), "weeks\n")
```

## Sensitivity Analysis

\code{sensitivity()} decomposes total project variance by task contribution
using rank correlation, producing the inputs for a Tornado chart.

```{r sensitivity, fig.cap="Tornado chart: each bar shows a task's rank correlation with total project duration."}
sens <- sensitivity(bp$task_distributions)
names(sens) <- paste0("T", seq_along(sens))
barplot(sort(sens),
  horiz = TRUE, col = "steelblue",
  main = "Sensitivity: Task Contributions",
  xlab = "Rank Correlation with Total"
)
```

Tasks with larger bars are the dominant drivers of total risk; mitigation
resources should be directed there first.


# Earned Value Management {#sec-evm}

The EVM module implements the complete ANSI/EIA-748 suite of eleven
performance metrics \citep{Fleming2010}. The three core quantities---Planned
Value (PV), Earned Value (EV), and Actual Cost (AC)---are computed first;
all performance indices and forecasts follow from them.

```{r evm-setup}
bac         <- 500000                          # budget at completion ($)
schedule    <- c(0.10, 0.25, 0.50, 0.75, 1.00) # cumulative value curve
time_period <- 3
```

```{r evm-core}
pv_val <- pv(bac, schedule, time_period)
ev_val <- ev(bac, actual_per_complete = 0.40)
ac_val <- ac(c(45000, 110000, 135000), time_period, cumulative = FALSE)

cat("PV: $", format(pv_val, big.mark = ","), "\n")
cat("EV: $", format(ev_val, big.mark = ","), "\n")
cat("AC: $", format(ac_val, big.mark = ","), "\n")
```

```{r evm-indices}
spi_val  <- spi(ev_val, pv_val)
cpi_val  <- cpi(ev_val, ac_val)
sv_val   <- sv(ev_val,  pv_val)
cv_val   <- cv(ev_val,  ac_val)

cat("SPI:", round(spi_val, 3), "  CPI:", round(cpi_val, 3), "\n")
cat("SV: $", format(round(sv_val), big.mark = ","),
    "  CV: $", format(round(cv_val), big.mark = ","), "\n")
```

Both SPI $< 1$ and CPI $< 1$ indicate the project is behind schedule and over
budget. The \code{eac()} function provides three completion forecasts under
different efficiency assumptions:

```{r evm-eac}
cat("EAC (typical):  $", format(round(eac(bac, method = "typical",
    cpi = cpi_val)), big.mark = ","), "\n")
cat("EAC (atypical): $", format(round(eac(bac, method = "atypical",
    ac = ac_val, ev = ev_val)), big.mark = ","), "\n")
cat("EAC (combined): $", format(round(eac(bac, method = "combined",
    cpi = cpi_val, ac = ac_val, ev = ev_val, spi = spi_val)),
    big.mark = ","), "\n")
```

The remaining EVM metrics are:

```{r evm-remaining}
eac_typ <- eac(bac, method = "typical", cpi = cpi_val)
cat("ETC:  $", format(round(etc(bac, ev_val, cpi = cpi_val)), big.mark = ","), "\n")
cat("TCPI:", round(tcpi(bac, ev_val, ac_val), 3), "\n")
cat("VAC:  $", format(round(vac(bac, eac_typ)), big.mark = ","), "\n")
```

The To-Complete Performance Index (TCPI) reports the cost efficiency required
on all remaining work to finish within the original budget; values substantially
above 1.0 signal an unrealistic recovery target.


# Bayesian Risk Methods {#sec-bayes}

Building on the causal structure introduced in Govan (2014) and
Govan and Damnjanovic (2016), the Bayesian module models risk event
probabilities from root causes and supports formal updating as project
observations become available \citep{Gelman2013,Pearl2009}.

## Prior Risk Probability

\code{risk\_prob()} computes the probability of a risk event $R$ from multiple
independent root causes. Each cause contributes a marginal probability by the
law of total probability, and the causes are combined with a noisy-OR---the
probability that at least one cause triggers $R$---so the result always lies in
$[0,1]$:
$$P(R) = 1 - \prod_{i}\Bigl[1 - \bigl(P(R \mid C_i)\,P(C_i)
  + P(R \mid \bar{C}_i)\,P(\bar{C}_i)\bigr)\Bigr]$$
For the \code{building\_project} schedule-delay risk with two root causes
(adverse weather and late design changes):

```{r prior-risk}
prior <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                   bp$risks_given_not_causes)
cat("Prior P(delay):", round(prior, 3), "\n")
```

## Bayesian Updating

\code{risk\_post\_prob()} updates the risk probability after observing cause
status on site. An observed cause is fixed to its conditional probability;
unobserved causes are passed as \code{NA} and retain their prior (marginal)
contribution.

```{r posterior-risk}
# Weather observed to have occurred; design changes not yet assessed
post <- risk_post_prob(bp$cause_probs, bp$risks_given_causes,
                       bp$risks_given_not_causes, bp$observed_causes)
cat("Posterior P(delay):", round(post, 3), "\n")
cat("Change:            ", round(post - prior, 3), "\n")
```

Confirming that adverse weather has occurred raises the delay probability
relative to the prior. The posterior informs contingency updates mid-project.

## Cost Risk Distributions

\code{cost\_pdf()} and \code{cost\_post\_pdf()} sample prior and posterior
cost distributions by simulating which risks materialize and summing their
cost impacts.

```{r cost-dist, fig.cap="Prior and posterior cost distributions. Observing Cause 1 shifts the distribution rightward, increasing the expected cost impact."}
risk_probs        <- c(0.3, 0.5, 0.2)
means_given_risks <- c(10000, 15000, 5000)
sds_given_risks   <- c(2000,  1000,  1000)
base_cost         <- 2000

prior_costs <- cost_pdf(5000, risk_probs, means_given_risks,
                        sds_given_risks, base_cost)
post_costs  <- cost_post_pdf(5000,
  observed_risks = c(1, NA, 0),   # Risk 1 present, Risk 2 unobserved, Risk 3 absent
  means_given_risks, sds_given_risks, base_cost
)

plot(density(prior_costs), col = "steelblue", lwd = 2,
     main = "Cost Risk: Prior vs. Posterior",
     xlab = "Cost ($)", xlim = range(c(prior_costs, post_costs)))
lines(density(post_costs), col = "tomato", lwd = 2)
legend("topright", legend = c("Prior", "Posterior"),
       col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

## Correlation Matrix Construction

\code{cor\_matrix()} constructs a valid positive-definite correlation matrix
from a vector of upper-triangle entries, for use as input to \code{mcs()}.

```{r cor-matrix}
dists <- list(
  normal  = function(n) rnorm(n, mean = 10, sd = 2),
  triang  = function(n) mc2d::rpert(n, min = 5, mode = 10, max = 15),
  uniform = function(n) runif(n, min = 8, max = 12)
)
C_emp <- cor_matrix(num_samples = 1000, num_vars = 3, dists = dists)
print(round(C_emp, 3))
```


# Learning Curves {#sec-learning}

Sigmoidal (S-curve) models capture the characteristic pattern of project
progress: slow start, rapid acceleration, and plateau approaching 100\%
completion. \pkg{PRA} provides three model types---Logistic, Pearl, and
Gompertz---fitted via the Levenberg--Marquardt nonlinear least-squares
algorithm in \pkg{minpack.lm} \citep{Elzhov2016}.

```{r sigmoidal-data}
progress <- data.frame(
  time       = 1:9,
  completion = c(5, 15, 40, 60, 70, 75, 80, 85, 90)
)
```

```{r sigmoidal-fit}
fit_log  <- fit_sigmoidal(progress, "time", "completion", "logistic")
fit_gomp <- fit_sigmoidal(progress, "time", "completion", "gompertz")

cat("Logistic RSE: ", round(summary(fit_log)$sigma,  2), "\n")
cat("Gompertz RSE: ", round(summary(fit_gomp)$sigma, 2), "\n")
```

```{r sigmoidal-plot, fig.cap="Logistic learning curve with 95\\% confidence band. The shaded region widens as extrapolation extends beyond the observed data range."}
plot_sigmoidal(fit_log, progress, "time", "completion", "logistic",
               conf_level = 0.95,
               main = "Logistic Learning Curve",
               xlab = "Week", ylab = "Completion (%)")
```

\code{predict\_sigmoidal()} generates forecasts with confidence bounds at
arbitrary future time points:

```{r sigmoidal-predict}
preds <- predict_sigmoidal(fit_log, seq(10, 12, by = 0.5),
                            "logistic", conf_level = 0.95)
print(round(preds, 1))
```


# Dependency Structure Analysis {#sec-dsm}

Building on Govan and Damnjanovic (2016, 2020) and the foundational DSM
literature \citep{Steward1981, Browning2001}, the DSM module quantifies
structural coupling between project tasks through shared resources and risks.

## Design Structure Matrices

The **Parent DSM** is derived from the Resource--Task matrix $\mathbf{S}$
(rows = resources, columns = tasks) as $\mathbf{P} = \mathbf{S}^\top\mathbf{S}$.
Off-diagonal entry $P_{jk}$ counts the number of resources shared between
tasks $j$ and $k$---the structural pathway through which disruptions propagate
\citep{Govan2016}.

```{r dsm-S}
S <- matrix(c(
  1, 0, 1, 0,
  1, 1, 0, 0,
  0, 1, 0, 1,
  0, 0, 1, 1,
  0, 1, 1, 0
), nrow = 4, ncol = 5)
rownames(S) <- paste0("R", 1:4)
colnames(S) <- paste0("T", 1:5)
```

```{r parent-dsm, fig.cap="Parent DSM heatmap. Darker cells indicate more shared resources and stronger coupling between task pairs."}
p <- parent_dsm(S)
plot(p)
```

The **Grandparent DSM** adds a risk layer via the Risk--Resource matrix
$\mathbf{R}$, tracing dependencies from risks through resources to tasks.
Off-diagonal entry $G_{jk}$ counts the number of risks shared between
task $j$ and task $k$ through the resource chain---a direct operationalization
of the structural network measures developed in Govan and Damnjanovic (2020).

```{r dsm-R}
R_mat <- matrix(c(
  1, 0, 1,
  1, 1, 0,
  0, 1, 0,
  0, 0, 1
), nrow = 3, ncol = 4)
rownames(R_mat) <- paste0("Risk", 1:3)
colnames(R_mat) <- paste0("R", 1:4)
```

```{r grandparent-dsm, fig.cap="Grandparent DSM heatmap. Shows risk-based coupling: task pairs that share risk exposure through common resources tend to fail together."}
g <- grandparent_dsm(S, R_mat)
plot(g)
```

Task pairs with high off-diagonal values in the Grandparent DSM represent
concentrated risk exposure; these pairs are prime candidates for contingency
buffering or resource decoupling.


# Probabilistic Networks {#sec-network}

**Experimental.** The probabilistic network module extends DSM structural
analysis by attaching probability distributions to each node, enabling
forward simulation of cost uncertainty and Bayesian conditioning on
observed evidence. A probabilistic network is a directed acyclic graph in
which nodes represent project variables (risks, resources, tasks, and the
project total) and edges encode conditional dependencies
\citep{Pearl2009, Govan2014}. Where a DSM counts shared resources between
task pairs, a probabilistic network propagates distributions through the
causal chain from risks to total project cost.

## Building a Network

\code{prob\_net()} constructs the network from a node data frame, a links
data frame, and a named list of distributions. Distributions may be
\code{normal}, \code{lognormal}, \code{uniform}, \code{discrete}, or
\code{conditional} (switching between two sub-distributions depending on
a discrete parent node); \code{aggregate} nodes sum their parents.

```{r net-setup}
nodes <- data.frame(
  id = c("A","B","C","D","E","F","G","H","I"),
  label = c("Risk-1","Risk-2","Resource-1","Resource-2","Resource-3",
            "Task-1","Task-2","Task-3","Project"),
  stringsAsFactors = FALSE
)
links <- data.frame(
  source = c("A","B","C","D","E","F","G","H"),
  target = c("C","D","F","G","H","I","I","I"),
  stringsAsFactors = FALSE
)
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)),
  D = list(type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = "C"),
  G = list(type = "aggregate", nodes = "D"),
  H = list(type = "aggregate", nodes = "E"),
  I = list(type = "aggregate", nodes = c("F","G","H"))
)
net <- prob_net(nodes, links, distributions = distributions)
```

## Forward Simulation

\code{prob\_net\_sim()} draws samples from every node in topological order
and returns a data frame with one column per node.

```{r net-sim}
sim_net <- prob_net_sim(net, num_samples = 10000)
cat("Mean project cost: $", format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("P80 project cost:  $", format(round(quantile(sim_net$I, 0.80)),
    big.mark = ","), "\n")
```

## Learning: Conditioning on Evidence

\code{prob\_net\_learn()} clamps observed nodes to fixed values and
re-simulates the downstream distribution---an observational update in the
sense of \citet{Pearl2009}. The following call fixes Risk-2 (node B) to
absent (value 0), ruling out Technical Complexity as a concern:

```{r net-learn, fig.cap="Total project cost before and after observing that Risk-2 (Technical Complexity) did not occur. Conditioning on Risk-2 = 0 eliminates the heavy upper tail and shifts the distribution leftward by roughly \\$30,000 in expectation."}
learn_net <- prob_net_learn(net, observations = list(B = 0),
                            num_samples = 10000)

plot(density(sim_net$I),   col = "steelblue", lwd = 2,
     main = "Total Project Cost: Prior vs. Posterior",
     xlab = "Cost ($)")
lines(density(learn_net$I), col = "tomato", lwd = 2)
legend("topright",
  legend = c("Prior (Risk-2 uncertain)", "Posterior (Risk-2 = 0)"),
  col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

\code{prob\_net\_update()} performs structural surgery on the network by
adding or removing edges and replacing node distributions, corresponding
to the \emph{do}-operator in Pearl's causal calculus
\citep{Pearl2009}---distinct from \code{prob\_net\_learn()}, which
conditions without altering the graph. The probabilistic network API
is marked experimental and may evolve in future releases.


# Model Context Protocol Integration {#sec-agent}

Version 0.5.0 of \pkg{PRA} exposes all analytical functions as tools callable
by large-language-model (LLM) agents through the Model Context Protocol (MCP)
\citep{MCP2024}. \code{pra\_mcp\_server()} starts a server that advertises each
tool---Monte Carlo simulation, EVM, Bayesian risk, learning curves, DSM, and
the rest---so that Claude, Claude Code, and other MCP-compatible clients can
invoke them as native tools. The server is registered in Claude Code with a
single shell command:

```{r mcp-code, eval=FALSE, engine="bash"}
claude mcp add -s project pra -- Rscript -e "PRA::pra_mcp_server()"
```

For Claude Desktop, add the following block to the MCP configuration file:

\begin{verbatim}
{
  "mcpServers": {
    "pra": { "command": "Rscript",
             "args": ["-e", "PRA::pra_mcp_server()"] }
  }
}
\end{verbatim}

The tool definitions are built once with \code{ellmer::tool()} \citep{ellmer}
and served by \pkg{mcptools}. Each tool validates its JSON inputs and returns a
plain-text summary for the LLM together with, where applicable, a rich HTML
result with inline plots for clients that render it.

To help an agent decide \emph{when} and \emph{how} to call each tool,
\pkg{PRA} bundles a set of Agent Skills under
\code{system.file("skills", package = "PRA")}. Each skill is a \code{SKILL.md}
file with structured guidance---when to use the method, which tool to call, the
JSON argument format, and how to interpret the result---covering Monte Carlo
simulation, sensitivity and contingency, EVM, Bayesian risk, learning curves,
and dependency structure analysis, plus a routing overview. Pairing the skills
with the MCP server lets an agent carry out a complete project risk workflow
(for example, \code{mcs\_tool} followed by \code{contingency\_tool} and
\code{sensitivity\_tool}) from a natural-language request, without embedding an
LLM in the package itself.

The MCP layer is strictly optional: every analytical function operates without
any optional dependency. Only \code{pra\_mcp\_server()} requires \pkg{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}.


# Illustrative Case Study {#sec-casestudy}

This section demonstrates an integrated, end-to-end project risk analysis
for the \code{building\_project} dataset introduced in
Section~\ref{sec-uncertainty}, progressing through uncertainty
quantification, performance monitoring, Bayesian risk updating, and
dependency structure analysis in a single \proglang{R} session.

## Step 1: Characterize Task Uncertainty

```{r cs-mcs}
bp <- building_project
set.seed(42)
sim <- mcs(10000, bp$task_distributions)
res <- contingency(sim, phigh = 0.80, pbase = 0.50)
cat("P50 schedule:", round(quantile(sim$total_distribution, 0.50), 1), "weeks\n")
cat("P80 schedule:", round(quantile(sim$total_distribution, 0.80), 1), "weeks\n")
cat("Contingency (P80-P50):", round(res, 2), "weeks\n")
```

## Step 2: Monitor Earned Value at Mid-Project

```{r cs-evm}
pv_cs <- pv(bp$bac, bp$schedule, bp$time_period)
ev_cs <- ev(bp$bac, bp$actual_per_complete)
ac_cs <- ac(bp$actual_costs, bp$time_period, cumulative = FALSE)

cat("SPI:", round(spi(ev_cs, pv_cs), 3),
    " CPI:", round(cpi(ev_cs, ac_cs), 3), "\n")
cat("EAC (typical): $", format(round(eac(bp$bac,
    method = "typical", cpi = cpi(ev_cs, ac_cs))), big.mark = ","), "\n")
```

## Step 3: Update Risk Probability After Site Inspection

```{r cs-bayes}
prior_r <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                     bp$risks_given_not_causes)
post_r  <- risk_post_prob(bp$cause_probs, bp$risks_given_causes,
                          bp$risks_given_not_causes, bp$observed_causes)
cat("Prior P(delay):", round(prior_r, 3),
    "-> Posterior P(delay):", round(post_r, 3), "\n")
```

## Step 4: Identify Coupled Tasks via DSM

```{r cs-dsm, fig.cap="Grandparent DSM for the case-study project, showing which task pairs share the most risk exposure through common resources."}
g_cs <- grandparent_dsm(bp$resource_task, bp$risk_resource)
plot(g_cs)
```

The Grandparent DSM reveals which task pairs share risk exposure through
common resources, informing targeted contingency allocation and resource
decoupling decisions.


# Summary {#sec-summary}

\pkg{PRA} is a CRAN-available \proglang{R} package that provides a unified,
open-source implementation of quantitative project risk analysis methods.
It operationalizes the causal network and structural network frameworks
developed in Govan (2014) and Govan and Damnjanovic (2016, 2020) through
ten integrated analytical modules: SMM, MCS, contingency analysis,
sensitivity analysis, EVM (11 metrics), Bayesian risk inference and updating,
cost risk distributions, sigmoidal learning curves, correlation matrix
construction, and design structure matrices. An experimental eleventh
module provides probabilistic networks with forward simulation and
Bayesian conditioning.

Version 0.5.0 exposes every analytical function as a tool over the Model
Context Protocol, complemented by a bundled set of Agent Skills, so that
LLM agents can drive the package without any embedded model. All exported
functions adhere to rOpenSci SRR standards, with a test suite verifying
correctness, edge-case handling, and numerical stability across three
operating systems. The package is licensed under MIT and the source is
available at \url{https://github.com/paulgovan/PRA}; documentation and
a companion online book are at \url{https://paulgovan.github.io/PRA/}.

## Limitations and Future Work

Several limitations apply. First, \pkg{PRA} focuses on quantitative
uncertainty propagation; it does not model schedule logic (critical path,
activity dependencies, or resource leveling), which remain the domain of
dedicated scheduling engines. Second, \code{mcs()} induces task correlation by
applying a Cholesky factor to the sampled values, which reproduces the target
rank correlation but distorts the marginal means; the examples in this article
therefore treat tasks as independent for Monte Carlo simulation and use
\code{smm()}, which handles correlation analytically, when correlation matters.
A copula-based implementation that preserves the marginals is planned. Third,
the probabilistic network module is experimental: the API may change and no
convergence diagnostics are provided for the Monte Carlo estimators. Fourth,
the package currently accepts data as \proglang{R} vectors and lists; direct
import from project management file formats (Microsoft Project, Primavera P6,
CSV exports from Jira or Asana) is not supported and is planned for a future
release.

# Computational details {-}

The results in this article were produced using \proglang{R} \citep{RCore}
with the \pkg{PRA} package (version 0.5.0) and its hard dependencies
\pkg{mc2d} \citep{mc2d} and \pkg{minpack.lm} \citep{Elzhov2016}. The optional
Model Context Protocol server additionally uses \pkg{ellmer} \citep{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}. \proglang{R} itself and all packages used
are available from the Comprehensive \proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}. A random seed of 42 is set at the start of
the session so that the simulation output is reproducible.

# References {-}
