---
documentclass: jss
author:
  - name: Paul Govan
    orcid: 0000-0002-1821-8492
    affiliation: GE Aerospace
    address: 3290 Patterson Ave SE, Grand Rapids, MI 49512, USA
    email: \email{paul.govan@geaerospace.com}
    url: https://paulgovan.com
title:
  formatted: "Quantitative Project Risk Analysis with \\pkg{PRA}"
  plain:     "Quantitative Project Risk Analysis with PRA"
  short:     "\\pkg{PRA}: Project Risk Analysis"
abstract: >
  The \pkg{PRA} package for \proglang{R} is the first package to unify a core set of widely adopted quantitative project risk analysis methods in a single, open-source implementation: the Second Moment Method (SMM), Monte Carlo simulation (MCS), earned value management (EVM), contingency and sensitivity analysis, noisy-OR risk inference and updating, sigmoidal learning curves, and design structure matrices (DSM). Beyond this foundational core, \pkg{PRA} provides an open-source implementation of probabilistic network methods for project risk, supporting the forward simulation of cost uncertainty and Bayesian conditioning along causal chains. \pkg{PRA} also exposes every analytical function as a tool over the Model Context Protocol (MCP), paired with a documented set of Agent Skills, so that large-language-model (LLM) agents can run project risk analyses directly. The package is annotated against the rOpenSci general standards for statistical software and is supported by a continuous integration test suite running on macOS, Windows, and Ubuntu.
  
keywords:
  formatted: [project risk analysis, Monte Carlo simulation, earned value management,
              "Bayesian inference", "design structure matrix", "\\proglang{R}"]
  plain:     [project risk analysis, Monte Carlo simulation, earned value management,
              Bayesian inference, design structure matrix, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{longtable}
output: rticles::jss_article
bibliography: pra-jss.bib
---

```{r setup, include=FALSE}
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center",
                      fig.width = 5, fig.height = 3.5,
                      out.width = "80%")
set.seed(42)
library(PRA)
```

# Introduction

Quantitative risk analysis is a foundational discipline in project management,
enabling organizations to assess schedule and cost uncertainty, identify
high-impact risk drivers, and set evidence-based contingency reserves.
This article presents \pkg{PRA}, an open-source \proglang{R} package that
unifies the core quantitative methods of project risk analysis, schedule and
cost uncertainty, earned value management, sensitivity and contingency
analysis, learning curves, and dependency structure analysis, under a single,
reproducible API.

Despite broad industry adoption of quantitative methods, open-source tooling
for integrated project risk analysis in \proglang{R} remains fragmented.
Proprietary platforms such as Oracle Primavera Risk Analysis and Palisade
\@RISK provide integrated workflows but impose licensing costs and limit
reproducibility. Within \proglang{R}, no single package unifies Monte Carlo
risk analysis, earned value management, noisy-OR risk inference, sigmoidal
learning curves, and dependency structure analysis under a common API.
The \pkg{mc2d} package \citep{mc2d} provides general-purpose two-dimensional
Monte Carlo simulation but has no project-specific constructs.
The \pkg{sensitivity} package \citep{sensitivity} implements variance-based
indices for general functions but does not integrate with schedule or cost
models. EVM metrics are absent from CRAN entirely. Bayesian packages such as
\pkg{rstan} \citep{rstan} and \pkg{brms} \citep{brms} operate at a high level
of generality and require substantial expertise to apply in project risk
contexts. For probabilistic networks specifically, \pkg{bnlearn}
\citep{bnlearn} and \pkg{gRain} \citep{gRain} are the established
\proglang{R} tools: \pkg{bnlearn} provides structure and parameter learning
along with graph-surgery interventions through \code{mutilated()}, and
\pkg{gRain} performs exact junction-tree inference. Both are more general and
more capable inference engines than the sampling-based module of
Section \ref{sec-network}. Neither, however, carries a project-risk vocabulary
of risks, resources, and tasks, and neither connects to schedule, cost, or
earned value; \pkg{PRA}'s contribution here is that project layer rather than
the underlying inference.

\pkg{PRA} is the first package available on CRAN to integrate these project risk analysis methods in a single API. The comparison in Table \ref{tab:comparison} covers the closest general-purpose Monte Carlo and sensitivity packages and the dominant commercial platform, while a broader survey of project-specific \proglang{R} packages is provided below. Where the existing ecosystem is fragmented across general-purpose packages, none of which are project-specific, \pkg{PRA} provides a single reproducible, auditable API for the full workflow, integrating outputs across modules in a way that the fragmented ecosystem does not. For example, MCS results are directly fed into contingency and sensitivity analyses, and noisy-OR risk updates can re-parameterize task distributions for subsequent simulations. The package targets project analysts who work in \proglang{R} and require outputs that are compatible with standard reporting pipelines. It is annotated against the rOpenSci Statistical Software Review (SRR) general standards \citep{rOpenSciStats2021} and uses minimal hard dependencies: \pkg{mc2d} \citep{mc2d} for triangular distribution families and \pkg{minpack.lm} \citep{Elzhov2016} for nonlinear least-squares fitting.

\pkg{PRA} also provides an \proglang{R} implementation of probabilistic-network methods for project risk that grounds causal networks in the resource-based view, showing that organizational capabilities and shared resources, not schedule uncertainty alone, are the structural pathways through which risks propagate across project tasks \citep{Govan2014, Govan2016}. This approach is extended with structural network measures that quantify the topology of project risk networks \citep{Govan2020}. The probabilistic network module (Section \ref{sec-network}) operationalizes these ideas by propagating distributions through the causal chain from risks to the total project risk.

\pkg{PRA} is the first project risk analysis package to expose its analytical methods as tools callable by LLM agents over the Model Context Protocol (MCP). The optional MCP server is activated through \code{Suggests}-only dependencies; therefore, the analytical core incurs no additional runtime cost.

The following table underscores the primary contribution: no existing
\proglang{R} package or commercial platform covers the foundational project
risk workflow that \pkg{PRA} unifies. It summarizes how \pkg{PRA} compares with
the most closely related \proglang{R} packages and the dominant commercial
platform.

```{r comparison-table, echo=FALSE, results="asis"}
comp <- data.frame(
  Feature = c(
    "MCS schedule risk", "EVM (11 metrics)", "Noisy-OR updating",
    "Learning curves", "Design structure matrices",
    "Probabilistic networks", "MCP tool server",
    "Open-source (CRAN)", "SRR general standards"
  ),
  PRA         = c("Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes"),
  mc2d        = c("Partial","---","---","---","---","---","---","Yes","---"),
  sensitivity = c("---","---","---","---","---","---","---","Yes","---"),
  `rstan/brms`= c("---","---","Yes","---","---","Partial","---","Yes","---"),
  `bnlearn/gRain` = c("---","---","Partial","---","---","Yes","---",
                      "Yes","---"),
  `@RISK`     = c("Yes","Yes","---","---","---","---","---","---","---"),
  check.names = FALSE
)
tab <- knitr::kable(
  comp,
  format = "latex",
  caption = paste0("\\label{tab:comparison}Comparison of \\pkg{PRA} with",
    " related",
    " \\proglang{R} packages and commercial tools. ``Partial'' = only",
    " general-purpose support, requiring custom modeling and not",
    " project-specific; ``---'' = not supported."),
  booktabs = TRUE, align = "lcccccc"
)
# Seven columns overrun the JSS text width at the default size.
cat(sub("\\centering", "\\centering\n\\small", tab, fixed = TRUE))
```

A small number of CRAN packages do target project management specifically, but
each addresses a single slice of the workflow rather than the integrated whole.
The \pkg{ProjectManagement} package \citep{ProjectManagement} is the closest in
spirit: it implements deterministic and stochastic scheduling, PERT/CPM,
critical-path identification, and Monte Carlo simulation over activity
durations, and is complementary to the schedule-network extension planned for
\pkg{PRA} (Section \ref{sec-summary}), but it does not provide earned value
management, noisy-OR risk updating, learning curves, dependency structure
matrices, or probabilistic networks. The \pkg{plan} package \citep{plan}
produces Gantt and burn-down charts for project visualization but performs no
risk quantification. More generally, decision-analysis packages such as
\pkg{decisionSupport} \citep{decisionSupport} apply Monte Carlo simulation to
cost--benefit models under uncertainty, but are not organized around
project-schedule or earned-value constructs. None of
these packages, nor any single package on CRAN, spans the foundational project
risk workflow that \pkg{PRA} unifies. 

The gap is not limited to \proglang{R}: open-source tooling for integrated project risk
analysis remains sparse across languages, and no package in another ecosystem
matches \pkg{PRA}'s combined scope either. In \proglang{Python},
\pkg{PyCostTools} \citep{PyCostTools} implements learning-curve and
cost-estimating-relationship models similar in
spirit to Section \ref{sec-learning}, but remains a low-adoption project without peer-reviewed documentation. The
general-purpose \pkg{PyMC} \citep{PyMC2016} supports arbitrary Bayesian models
but, unlike the probabilistic-network module of Section \ref{sec-network}, has
no project-risk-specific vocabulary of risks, resources, and tasks. In
\proglang{Julia}, \pkg{MCHammer.jl} \citep{MCHammerJL} provides Monte Carlo
business-risk modeling with correlated sampling and sensitivity tornado charts
in the style of Crystal Ball and \@RISK, but has seen minimal maintenance
activity in recent years. Design structure matrix tools outside \proglang{R} target software and systems
architecture rather than project risk, and no citable academic survey of
open-source project-risk software outside \proglang{R} could be located; this
survey is therefore drawn directly from the individual tools' own repositories
rather than a secondary source.

The package is installed from CRAN in the standard way:

```{r install, eval=FALSE}
install.packages("PRA")
library("PRA")
```

Optional dependencies, \pkg{ellmer} \citep{ellmer}, \pkg{mcptools} \citep{mcptools}, and
\pkg{jsonlite} \citep{jsonlite}, are only required for the MCP server, \code{pra\_mcp\_server()}.

The remainder of this article describes each analytical module in turn,
with working code examples drawn from the same \proglang{R} session.
Section \ref{sec-uncertainty} covers schedule and cost uncertainty methods
(SMM, MCS, contingency, sensitivity, and correlation matrix construction).
Section \ref{sec-evm} describes earned value management.
Section \ref{sec-learning} describes sigmoidal
learning curves. Section \ref{sec-bayes} covers noisy-OR risk inference and
updating. Section \ref{sec-network} introduces the probabilistic
network module. Section \ref{sec-dsm} covers design structure matrices,
built from that same network example. Section \ref{sec-agent} describes
the Model Context Protocol integration. Section \ref{sec-casestudy}
presents an integrated case study.
Section \ref{sec-summary} concludes.


# Schedule and Cost Uncertainty {#sec-uncertainty}

Two complementary methods are provided for quantifying schedule and cost
uncertainty: the Second Moment Method for rapid analytical estimates, and
Monte Carlo simulation for full distributional analysis. Contingency
extraction and sensitivity analysis complete the workflow.

To illustrate the workflow, \pkg{PRA} includes \code{building\_project}, a bundled example
dataset describing a mid-rise commercial building with six work packages,
adapted from the construction-project examples in \citet{Damnjanovic2020}.

```{r bp-load}
data("building_project", package = "PRA")
bp <- building_project
bp$task_names
```

## Second Moment Method

The Second Moment Method (SMM) propagates task-level means and variances to
project-level estimates using the Central Limit Theorem \citep{Benjamin2000}.
For a project with $n$ tasks:
$$E[X] = \sum_{i=1}^{n} E[X_i], \quad
  \text{Var}(X) = \sum_{i=1}^{n}\text{Var}(X_i) + 2\sum_{i<j}\rho_{ij}\sigma_i\sigma_j$$
The six task durations are triangular, so their means and variances may be obtained closed-form and propagated with \code{smm()} at the project level.

```{r smm-example}
tri_mean <- function(d) (d$a + d$b + d$c) / 3
tri_var  <- function(d) (d$a^2 + d$b^2 + d$c^2 -
                         d$a * d$b - d$a * d$c - d$b * d$c) / 18
task_means <- vapply(bp$task_distributions, tri_mean, numeric(1))
task_vars  <- vapply(bp$task_distributions, tri_var,  numeric(1))

result <- smm(task_means, task_vars)
cat("Total mean:", round(result$total_mean, 2), "weeks\n")
cat("Total SD:  ", round(result$total_std,  2), "weeks\n")
```

Positive correlation between tasks inflates the total variance; \code{smm()}
accepts a correlation matrix and adds the covariance terms analytically:

```{r smm-cor}
result_cor <- smm(task_means, task_vars, bp$cor_mat)
cat("Total SD (correlated):", round(result_cor$total_std, 2), "weeks\n")
```

SMM is useful when only first and second moments are available and a rapid
estimate is needed. When task distributions are non-normal or accurate tail
behavior is required, Monte Carlo simulation is preferred.

## Monte Carlo Simulation

\code{mcs()} propagates uncertainty through a project network by drawing
samples from user-specified task distributions, triangular, normal, or
uniform, over a large number of iterations \citep{Vose2008}. By default the
tasks are sampled independently, with an optional correlation matrix that induces
dependence between them.
For a project with $n$ tasks, one simulated draw of the total duration is
$$T^{(s)} = \sum_{i=1}^{n} Y_i^{(s)}, \qquad s = 1, \ldots, S,$$
where $Y_i^{(s)}$ is the $s$-th draw of task $i$'s duration, and the project-level
estimates are the Monte Carlo moments of the $S$ simulated totals,
$$\hat{E}[T] = \frac{1}{S}\sum_{s=1}^{S} T^{(s)}, \qquad
  \widehat{\text{Var}}(T) = \frac{1}{S-1}\sum_{s=1}^{S}\bigl(T^{(s)} - \hat{E}[T]\bigr)^2.$$
When a correlation matrix $\mathbf{C}$ is supplied, dependence between tasks is
induced by Cholesky decomposition. The correlation
matrix is factored as $\mathbf{U}^\top\mathbf{U} = \mathbf{C}$, and a matrix
$\mathbf{Z}$ of independent standard normal draws (columns = tasks) is
right-multiplied by the Cholesky factor $\mathbf{U}$, producing normal scores
that carry exactly the target correlation. Because task distributions are in
general not normal, the correlated scores are then returned to each task's own
scale: the standard normal CDF $\Phi$ maps each score to a percentile, and each
task's inverse CDF $F_i^{-1}$ maps that percentile back to a duration,
$$\mathbf{W} = \mathbf{Z}\mathbf{U}, \qquad
  U_i^{(s)} = \Phi\bigl(W_i^{(s)}\bigr), \qquad
  Y_i^{(s)} = F_i^{-1}\bigl(U_i^{(s)}\bigr).$$


```{r mcs-run}
set.seed(42)
results <- mcs(10000, bp$task_distributions)
cat("Simulated mean:", round(results$total_mean, 2), "weeks\n")
cat("Simulated SD:  ", round(results$total_sd,   2), "weeks\n")
```

The \code{mcs()} function returns an S3 object of class \code{"mcs"} with
a \code{print()} method that summarizes percentiles. The full simulated
distribution is available in \code{results\$total\_distribution}.
On a modern laptop, 10,000 simulations of a six-task network complete in
under one second; increasing the count to 100,000 provides more precise tail
estimates at roughly ten-fold additional runtime. The default of 10,000
balances accuracy and speed for most practical analyses.

Supplying the dataset's correlation matrix induces dependence between the tasks
through the transform above:

```{r mcs-cor}
set.seed(42)
results_cor <- mcs(10000, bp$task_distributions, bp$cor_mat)
cat("Simulated SD (correlated):", round(results_cor$total_sd, 2), "weeks\n")
```

### Numerical Validation

For independent tasks the SMM estimates are the exact first two moments of the
total: the total mean equals the sum of the task means by linearity of
expectation, and the total variance equals the sum of the task variances by
additivity of variance under independence. Supplying the correlation matrix adds
the covariance terms analytically, giving a second exact target. Monte Carlo
simulation should reproduce both, and it does, to within Monte Carlo error. The
correlated comparison is the sharper test of the two: an error in the Cholesky
factorization or in the rank-to-scale mapping would surface here as a mismatched
total variance, while leaving the independent case untouched.

```{r validation}
cat(sprintf("Mean:            SMM %.2f  vs  MCS %.2f\n",
            result$total_mean, results$total_mean))
cat(sprintf("SD:              SMM %.2f  vs  MCS %.2f\n",
            result$total_std,  results$total_sd))
cat(sprintf("SD (correlated): SMM %.2f  vs  MCS %.2f\n",
            result_cor$total_std, results_cor$total_sd))
```

```{r mcs-hist, fig.cap="Distribution of simulated total project durations. The dashed vertical line marks the mean."}
hist(results$total_distribution,
  breaks = 50, freq = FALSE,
  main = "MCS: Total Project Duration",
  xlab = "Duration (weeks)", col = "steelblue", border = "white"
)
lines(density(results$total_distribution), col = "tomato", lwd = 2)
abline(v = results$total_mean, lty = 2, lwd = 1.5)
```

## Correlation Matrix Construction

Both \code{smm()} and \code{mcs()} accept a correlation matrix to model
dependence between tasks. \code{cor\_matrix()} returns the empirical Pearson
correlation matrix of $m$ columns sampled from a supplied set of task
distributions. Because the columns are drawn independently, the off-diagonal
entries are sampling noise of order $N^{-1/2}$ about zero: the function is a
generator of correctly shaped, positive-definite correlation input for testing
and for establishing the near-independent baseline, not an estimator of
dependence between tasks. Correlations that reflect genuine shared drivers must
come from historical data or expert elicitation.
For samples $x_{ki}$, $k = 1,\ldots,N$, drawn from each of $m$ task distributions
$i = 1,\ldots,m$, \code{cor\_matrix()} returns the empirical Pearson correlation
matrix $\mathbf{C}$ with entries
$$C_{ij} = \frac{\sum_{k=1}^{N}(x_{ki}-\bar{x}_i)(x_{kj}-\bar{x}_j)}
  {\sqrt{\sum_{k=1}^{N}(x_{ki}-\bar{x}_i)^2}\;\sqrt{\sum_{k=1}^{N}(x_{kj}-\bar{x}_j)^2}}.$$

```{r cor-matrix}
set.seed(42)
dists <- list(
  normal  = function(n) rnorm(n, mean = 10, sd = 2),
  triang  = function(n) mc2d::rpert(n, min = 5, mode = 10, max = 15),
  uniform = function(n) runif(n, min = 8, max = 12)
)
C_emp <- cor_matrix(num_samples = 1000, num_vars = 3, dists = dists)
print(round(C_emp, 3))
```

## Contingency Analysis

\code{contingency()} extracts a reserve buffer as the difference between a
high-confidence percentile and the base (median) estimate, a standard
approach in project risk practice \citep{PMBOK2021}. Writing $Q_p(T)$ for the
empirical $p$-th quantile of the simulated total duration $T$, the
contingency reserve is
$$\text{CR} = Q_{p_{\text{high}}}(T) - Q_{p_{\text{base}}}(T).$$

```{r contingency}
reserve <- contingency(results, phigh = 0.80, pbase = 0.50)
cat("Schedule contingency (P80 - P50):", round(reserve, 2), "weeks\n")
```

## Sensitivity Analysis

\code{sensitivity()} decomposes total project variance by task contribution,
producing the inputs for a Tornado chart. Each task's index is its own
variance plus its covariance with every other task, expressed as a
proportion of total project variance. Let $\text{Var}(X_i)$ be task $i$'s
variance and $\text{Cov}(X_i, X_j) = \rho_{ij}\sqrt{\text{Var}(X_i)\text{Var}(X_j)}$
its covariance with task $j$ (zero if no correlation matrix is supplied).
Total project variance decomposes as
$$\text{Var}(T) = \sum_{i=1}^{n}\text{Var}(X_i) + 2\sum_{i<j}\text{Cov}(X_i, X_j),$$
and \code{sensitivity()} reports each task's share of that total,
$$S_i = \frac{\text{Var}(X_i) + \sum_{j \neq i}\text{Cov}(X_i, X_j)}{\text{Var}(T)},
  \qquad \sum_{i=1}^{n} S_i = 1.$$

```{r sensitivity, fig.cap="Tornado chart: each bar shows a task's share of total project duration variance, sorted ascending so the largest driver appears at the top."}
sens <- sensitivity(bp$task_distributions)
names(sens) <- paste0("T", seq_along(sens))
barplot(sort(sens),
  horiz = TRUE, col = "steelblue",
  main = "Sensitivity: Task Contributions",
  xlab = "Proportion of Total Variance"
)
```

Tasks with larger bars are the dominant drivers of total risk; mitigation
resources should be directed there first.


# Earned Value Management {#sec-evm}

The EVM module implements the complete ANSI/EIA-748 suite of eleven
performance metrics \citep{Fleming2010}. The three core quantities, Planned
Value (PV), Earned Value (EV), and Actual Cost (AC), are computed first.
All performance indices and forecasts follow from them. Consider a project
with a \$500,000 budget at completion, evaluated at period 3 of a 5-period
schedule with cumulative planned value 10/25/50/75/100\%:

```{r evm-setup}
bac         <- 500000
schedule    <- c(0.10, 0.25, 0.50, 0.75, 1.00)
time_period <- 3
```

At this reporting point, 40\% of scope is actually complete, and actual costs
of \$45,000, \$110,000, and \$135,000 have been incurred across the three
periods to date:

```{r evm-core}
pv_val <- pv(bac, schedule, time_period)
ev_val <- ev(bac, actual_per_complete = 0.40)
ac_val <- ac(c(45000, 110000, 135000), time_period, cumulative = FALSE)

cat("PV: $", format(pv_val, big.mark = ","), "\n")
cat("EV: $", format(ev_val, big.mark = ","), "\n")
cat("AC: $", format(ac_val, big.mark = ","), "\n")
```

The Schedule Performance Index, Cost Performance Index, Schedule Variance,
and Cost Variance follow directly from PV, EV, and AC:
$$\text{SPI} = \frac{\text{EV}}{\text{PV}}, \quad
  \text{CPI} = \frac{\text{EV}}{\text{AC}}, \quad
  \text{SV} = \text{EV} - \text{PV}, \quad
  \text{CV} = \text{EV} - \text{AC}$$

```{r evm-indices}
spi_val  <- spi(ev_val, pv_val)
cpi_val  <- cpi(ev_val, ac_val)
sv_val   <- sv(ev_val,  pv_val)
cv_val   <- cv(ev_val,  ac_val)

cat("SPI:", round(spi_val, 3), "  CPI:", round(cpi_val, 3), "\n")
cat("SV: $", format(round(sv_val), big.mark = ","),
    "  CV: $", format(round(cv_val), big.mark = ","), "\n")
```

Both SPI $< 1$ and CPI $< 1$ indicate the project is behind schedule and over
budget. The \code{eac()} function provides three completion forecasts under
different efficiency assumptions \citep{Fleming2010}:
$$\text{EAC}_{\text{typical}} = \frac{\text{BAC}}{\text{CPI}}, \quad
  \text{EAC}_{\text{atypical}} = \text{AC} + (\text{BAC} - \text{EV}), \quad
  \text{EAC}_{\text{combined}} = \text{AC} +
  \frac{\text{BAC} - \text{EV}}{\text{CPI} \times \text{SPI}}$$
The typical method assumes remaining work proceeds at the cumulative cost
efficiency to date; the atypical method assumes the overrun to date was a
one-off and future work runs on plan; the combined method deflates the
remaining budget by both cost and schedule performance.

```{r evm-eac}
cat("EAC (typical):  $", format(round(eac(bac, method = "typical",
    cpi = cpi_val)), big.mark = ","), "\n")
cat("EAC (atypical): $", format(round(eac(bac, method = "atypical",
    ac = ac_val, ev = ev_val)), big.mark = ","), "\n")
cat("EAC (combined): $", format(round(eac(bac, method = "combined",
    cpi = cpi_val, ac = ac_val, ev = ev_val, spi = spi_val)),
    big.mark = ","), "\n")
```

The remaining EVM metrics are:

```{r evm-remaining}
eac_typ <- eac(bac, method = "typical", cpi = cpi_val)
cat("ETC:  $",
    format(round(etc(bac, ev_val, cpi = cpi_val)), big.mark = ","), "\n")
cat("TCPI:", round(tcpi(bac, ev_val, ac_val), 3), "\n")
cat("VAC:  $", format(round(vac(bac, eac_typ)), big.mark = ","), "\n")
```

The To-Complete Performance Index (TCPI) reports the cost efficiency required
on all remaining work to finish within the original budget; values substantially
above 1.0 signal an unrealistic recovery target. The Estimate to Complete (ETC)
gives the projected cost of the remaining work under the typical-method
efficiency assumption, and the negative Variance at Completion (VAC) confirms
the project is tracking toward a budget overrun rather than an underrun.


# Learning Curves {#sec-learning}

Sigmoidal (S-curve) models capture the characteristic pattern of project
progress: slow start, rapid acceleration, and plateau approaching 100\%
completion. \pkg{PRA} provides two model types, Logistic and Gompertz, fitted
via the Levenberg--Marquardt nonlinear least-squares algorithm in
\pkg{minpack.lm} \citep{Elzhov2016}. The Logistic model has the closed form
$$y(t) = \frac{K}{1 + \exp\bigl(-r(t - t_0)\bigr)},$$
while the Gompertz model is
$$y(t) = A\exp\bigl(-b\exp(-ct)\bigr).$$
Each model is fit by nonlinear least squares,
$$\hat{\theta} = \arg\min_{\theta}\sum_{k=1}^{m}\bigl(y_k - f(t_k;\theta)\bigr)^2,$$
where $\theta$ is $(K, r, t_0)$ or $(A, b, c)$ and $f$ is the corresponding
curve.

```{r sigmoidal-data}
progress <- data.frame(
  time       = 1:9,
  completion = c(5, 15, 40, 60, 70, 75, 80, 85, 90)
)
```

```{r sigmoidal-fit}
fit_log  <- fit_sigmoidal(progress, "time", "completion", "logistic")
fit_gomp <- fit_sigmoidal(progress, "time", "completion", "gompertz")

cat("Logistic RSE: ", round(summary(fit_log)$sigma,  2), "\n")
cat("Gompertz RSE: ", round(summary(fit_gomp)$sigma, 2), "\n")
```

The Gompertz model has the lower residual standard error of the two here,
reflecting the asymmetric approach to the 90\% plateau in the observed data, and
is therefore carried forward for the confidence band and forecasts below. The
Logistic form remains available via \code{fit\_sigmoidal()}; its symmetric
acceleration and deceleration is the more common default assumption for project
progress curves, and comparing the two residual standard errors is the
recommended way to choose between them in practice.

```{r sigmoidal-plot, fig.cap="Gompertz learning curve over the observed data range, with a 95\\% confidence band. The band is narrowest in the middle of the range, where the observations most tightly constrain the fit, and widens toward both ends."}
plot_sigmoidal(fit_gomp, progress, "time", "completion", "gompertz",
               conf_level = 0.95,
               main = "Gompertz Learning Curve",
               xlab = "Week", ylab = "Completion (%)")
```

\code{predict\_sigmoidal()} generates forecasts with confidence bounds at
arbitrary future time points:

```{r sigmoidal-predict}
preds <- predict_sigmoidal(fit_gomp, seq(10, 12, by = 0.5),
                            "gompertz", conf_level = 0.95)
print(round(preds, 1))
```

The forecast extends the fitted curve past the last observed week (9) to weeks
10 through 12. The point estimates rise only slightly across that span, because
the curve has already reached the neighborhood of its fitted asymptote
$\hat{A} \approx 87$. The interval is symmetric about the point estimate,
$\text{pred} \pm t\,\text{se}$, and widens in both directions with the horizon as
extrapolation uncertainty grows; the lower bound only appears flat because the
point estimate climbs at almost exactly the rate the half-width does.
A plateau below the last observed value is a real property of these nine
observations, which are still climbing roughly linearly at week 9: the data do
not yet identify where the S-curve turns over, and the fitted asymptote should
be treated as a lower bound on eventual completion rather than a prediction of
it.


# Noisy-OR Risk Inference {#sec-bayes}

Project risks are frequently triggered by one or more underlying root causes
whose presence is uncertain. \pkg{PRA} models this with a noisy-OR risk network
\citep{Pearl2009}: each root cause $C_i$ acts on a risk event $R$ independently,
and $R$ occurs if it is triggered by at least one cause. Write $q_i^{+}$ for the
probability that cause $i$ alone triggers $R$ when it is present and $q_i^{-}$
for the corresponding probability when it is absent; these are the link
probabilities supplied as \code{risks\_given\_causes} and
\code{risks\_given\_not\_causes}. Note that $q_i^{+}$ is not the model's
conditional $P(R \mid C_i)$, which is larger because the remaining causes may
also fire. We illustrate with the
schedule-delay risk in \code{building\_project}, which has two candidate causes.

## Prior Risk Probability

\code{risk\_prob()} first computes each cause's marginal contribution by the law
of total probability,
$$m_i = q_i^{+}\,P(C_i) + q_i^{-}\,\bigl(1 - P(C_i)\bigr),$$
and then combines the independent causes with a noisy-OR,
$$P(R) = 1 - \prod_{i=1}^{k}\bigl(1 - m_i\bigr),$$
which always lies in $[0, 1]$ and reduces to the single-cause marginal when
$k = 1$.

```{r bayes-prior}
bp <- building_project
prior_delay <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                         bp$risks_given_not_causes)
cat("Prior P(schedule delay):", round(prior_delay, 3), "\n")
```

## Posterior Risk Probability

When some causes are subsequently observed, for example, a site inspection
confirms one cause is present while the other remains unknown, \code{risk\_post\_prob()}
updates the estimate. An observed cause is fixed to its conditional probability,
while an unobserved cause (coded \code{NA}) retains its marginal contribution
$m_i$; the causes are recombined with the same noisy-OR, so prior and posterior
share a common scale:
$$P(R \mid \text{obs}) = 1 - \prod_{i=1}^{k}\bigl(1 - c_i\bigr), \qquad
  c_i = \begin{cases}
    q_i^{+} & C_i \text{ observed present},\\
    q_i^{-} & C_i \text{ observed absent},\\
    m_i     & C_i \text{ unobserved}.
  \end{cases}$$

```{r bayes-post}
post_delay <- risk_post_prob(
  bp$cause_probs, bp$risks_given_causes,
  bp$risks_given_not_causes, bp$observed_causes
)
cat("Posterior P(schedule delay):", round(post_delay, 3),
    " (observed causes:", paste(bp$observed_causes, collapse = ", "), ")\n")
```

Observing an aggravating cause raises the posterior risk probability; observing
a mitigating one lowers it. The update conditions on the observed causes within
a fixed model rather than revising beliefs about the model's own parameters, and
the diagnostic direction, $P(C_i \mid R)$, is not computed. This closed-form
update is inexpensive and complements the sample-based conditioning of the
probabilistic network module (Section \ref{sec-network}), which operates on full
cost distributions rather than event probabilities.


# Probabilistic Networks {#sec-network}

This module is an implementation of
probabilistic-network methods for project risk
\citep{Govan2014, Govan2016, Govan2020}. It is not a general Bayesian-network
engine, and does not aim to be one: \pkg{bnlearn} and \pkg{gRain} already fill
that role in \proglang{R}, with exact inference and structure learning. What
follows is the project-risk layer, in which nodes are risks, resources, tasks,
and the project total, and in which the quantity propagated is cost.
A probabilistic network is a directed acyclic graph in
which nodes represent project variables (risks, resources, tasks, and the
project total) and edges encode conditional dependencies
\citep{Pearl2009}, attaching a probability distribution to each
node so that cost uncertainty can be propagated forward along the causal
chain from risks to total project cost and conditioned on observed evidence.
Section \ref{sec-dsm} revisits this same example network from a structural
perspective, asking which tasks share resources or risk exposure rather than
how their costs propagate.

## Building a Network

\code{prob\_net()} constructs the network from a node data frame, a links
data frame, and a named list of distributions. Distributions may be
\code{normal}, \code{lognormal}, \code{uniform}, \code{discrete}, or
\code{conditional} (switching between two sub-distributions depending on
a discrete parent node); \code{aggregate} nodes sum their parents.
A probabilistic network encodes a joint distribution over nodes $X_1,\ldots,X_n$
that factorizes according to the DAG's parent structure,
$$p(x_1, \ldots, x_n) = \prod_{i=1}^{n} p\bigl(x_i \mid \mathrm{pa}(x_i)\bigr),$$
where $\mathrm{pa}(x_i)$ denotes the values of $X_i$'s parent nodes (the
\code{condition} node for a conditional distribution, or the summed parents for
an aggregate node) \citep{Pearl2009}. The links are load-bearing rather than
descriptive: \code{prob\_net()} requires the edges to match the dependencies the
distributions declare, so a graph that disagrees with its distributions is
rejected at construction, and it requires the nodes to be supplied in a
topological order, which is the order \code{prob\_net\_sim()} samples in and which
also guarantees the graph is acyclic.

The example network below has nine nodes: two risks (Risk-1, Risk-2), three
resources (Resource-1, Resource-2, Resource-3), three tasks (Task-1, Task-2,
Task-3), and the project total. Risk-1 occurs with probability 0.70 and
Risk-2 with probability 0.60; each risk feeds one resource, whose cost
distribution switches to a higher-mean, higher-variance branch when its risk
occurs, while Resource-3 carries no risk dependency. Each resource
aggregates one-to-one into a task (Task-1 from Resource-1, Task-2 from
Resource-2, Task-3 from Resource-3), and the three tasks sum into the
project total.

```{r net-setup}
nodes <- data.frame(
  id = c("A","B","C","D","E","F","G","H","I"),
  label = c("Risk-1","Risk-2","Resource-1","Resource-2","Resource-3",
            "Task-1","Task-2","Task-3","Project"),
  stringsAsFactors = FALSE
)
links <- data.frame(
  source = c("A","B","C","D","E","F","G","H"),
  target = c("C","D","F","G","H","I","I","I"),
  stringsAsFactors = FALSE
)
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)),
  D = list(type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = "C"),
  G = list(type = "aggregate", nodes = "D"),
  H = list(type = "aggregate", nodes = "E"),
  I = list(type = "aggregate", nodes = c("F","G","H"))
)
net <- prob_net(nodes, links, distributions = distributions)
```

## Forward Simulation

\code{prob\_net\_sim()} draws samples from every node in the order the nodes are
supplied, which must be a topological order of the graph, and returns a data
frame with one column per node.

```{r net-sim}
set.seed(42)
sim_net <- prob_net_sim(net, num_samples = 10000)
cat("Mean project cost: $",
    format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("P80 project cost:  $", format(round(quantile(sim_net$I, 0.80)),
    big.mark = ","), "\n")
```

The gap between the mean and the P80 project cost reflects the upper tail
introduced by the two risks (nodes A and B): when a risk occurs, its downstream
resource is drawn from the higher-cost conditional branch, pulling the tail of
the project total upward relative to the mean.

## Learning: Conditioning on Evidence

\code{prob\_net\_learn()} draws whole joint samples from the prior and retains
only those consistent with the observed values, an observational update in the
sense of \citet{Pearl2009}. Given $M$ prior draws
$x^{(1)}, \ldots, x^{(M)} \sim p(x)$ and observed nodes $X_E = e$, it estimates
the conditional expectation of any function $g$ of the network by rejection
sampling,
$$\widehat{E}\bigl[g(X) \mid X_E = e\bigr] =
  \frac{\sum_{m=1}^{M} \mathbb{1}\bigl\{x_E^{(m)} = e\bigr\}\, g\bigl(x^{(m)}\bigr)}
       {\sum_{m=1}^{M} \mathbb{1}\bigl\{x_E^{(m)} = e\bigr\}}.$$
The following call conditions on Risk-2 (node B) being
absent (value 0), ruling out Technical Complexity as a concern:

```{r net-learn, fig.cap="Total project cost before and after observing that Risk-2 (Technical Complexity) did not occur. Conditioning on Risk-2 = 0 eliminates the heavy upper tail and shifts the distribution leftward by roughly \\$18,000 in expectation."}
learn_net <- prob_net_learn(net, observations = list(B = 0),
                            num_samples = 10000)

prior_net_dens <- density(sim_net$I)
post_net_dens  <- density(learn_net$I)

plot(prior_net_dens, col = "steelblue", lwd = 2,
     main = "Total Project Cost: Prior vs. Posterior",
     xlab = "Cost ($)",
     xlim = range(c(sim_net$I, learn_net$I)),
     ylim = range(c(prior_net_dens$y, post_net_dens$y)))
lines(post_net_dens, col = "tomato", lwd = 2)
legend("topright",
  legend = c("Prior (Risk-2 uncertain)", "Posterior (Risk-2 = 0)"),
  col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

## Intervention: The do-Operator

Conditioning answers "what if we \emph{observe} that a risk did not occur?"
A distinct question is "what if we \emph{act} to eliminate it?"
\code{prob\_net\_update()} performs structural surgery on the
network, adding or removing edges and replacing node distributions, which
corresponds to the \emph{do}-operator in Pearl's causal calculus
\citep{Pearl2009}, as opposed to \code{prob\_net\_learn()}, which conditions
without altering the graph. Intervening to fix $X_k = x_k^{*}$ replaces $X_k$'s
conditional distribution with a point mass at $x_k^{*}$ and removes its incoming
edges, giving Pearl's truncated factorization,
$$p_{\mathrm{do}(X_k = x_k^{*})}(x_1, \ldots, x_n) =
  \mathbb{1}\bigl\{x_k = x_k^{*}\bigr\}
  \prod_{i \neq k} p\bigl(x_i \mid \mathrm{pa}(x_i)\bigr).$$
Suppose management funds a mitigation that
decouples Resource-2 from Risk-2: remove the Risk-2 $\to$ Resource-2 edge and
fix Resource-2 to its baseline cost.

```{r net-do}
inter_net <- prob_net_update(net,
  remove_links = data.frame(source = "B", target = "D"),
  update_distributions = list(
    D = list(type = "normal", mean = 50000, sd = 10000)
  )
)
do_net <- prob_net_sim(inter_net, num_samples = 10000)

cat("Prior mean:            $",
    format(round(mean(sim_net$I)),   big.mark = ","), "\n")
cat("Seeing (Risk-2 = 0):   $",
    format(round(mean(learn_net$I)), big.mark = ","), "\n")
cat("Doing  (do(Risk-2=0)): $",
    format(round(mean(do_net$I)),    big.mark = ","), "\n")
```

Here \emph{seeing} (\code{prob\_net\_learn()}) and \emph{doing}
(\code{prob\_net\_update()}) coincide because Risk-2 is an independent root
cause with no shared upstream confounder, so conditioning on it and
intervening on it induce the same downstream distribution. The two diverge,
however, when a risk shares a common cause with project cost, as the next
example shows.

## When Seeing and Doing Diverge

Suppose a latent common cause, adverse site conditions, drives both a rework
risk and, through a separate pathway, the foundation cost. Observing that rework
did \emph{not} occur is evidence that site conditions were favorable, which in
turn lowers the expected foundation cost; \emph{intervening} to prevent rework
severs the risk but leaves site conditions, and hence the foundation cost,
untouched. The network below encodes this structure: the common cause \code{U}
feeds both the rework risk \code{Rk} and the foundation cost \code{Wf}, while
the schedule cost \code{Vs} depends only on \code{Rk}.

```{r net-confound}
set.seed(42)
cf_nodes <- data.frame(
  id = c("U", "Rk", "Wf", "Vs", "Tc"), stringsAsFactors = FALSE)
cf_links <- data.frame(
  source = c("U", "U", "Rk", "Wf", "Vs"),
  target = c("Rk", "Wf", "Vs", "Tc", "Tc"), stringsAsFactors = FALSE)
cf_dists <- list(
  U  = list(type = "discrete", values = c(1, 0), probs = c(0.5, 0.5)),
  Rk = list(type = "conditional", condition = "U",
    true_dist  = list(type = "discrete",
      values = c(1, 0), probs = c(0.9, 0.1)),
    false_dist = list(type = "discrete",
      values = c(1, 0), probs = c(0.1, 0.9))),
  Wf = list(type = "conditional", condition = "U",
    true_dist  = list(type = "normal", mean = 60000, sd = 5000),
    false_dist = list(type = "normal", mean = 20000, sd = 5000)),
  Vs = list(type = "conditional", condition = "Rk",
    true_dist  = list(type = "normal", mean = 40000, sd = 5000),
    false_dist = list(type = "normal", mean = 10000, sd = 5000)),
  Tc = list(type = "aggregate", nodes = c("Wf", "Vs")))
cf_net <- prob_net(cf_nodes, cf_links, distributions = cf_dists)

cf_prior <- prob_net_sim(cf_net, num_samples = 10000)
cf_see   <- prob_net_learn(cf_net, observations = list(Rk = 0),
                           num_samples = 10000)
cf_do    <- prob_net_sim(prob_net_update(cf_net,
  remove_links = data.frame(source = "U", target = "Rk"),
  update_distributions = list(
    Rk = list(type = "discrete", values = c(1, 0), probs = c(0, 1)))),
  num_samples = 10000)

cat("Prior mean total:        $",
    format(round(mean(cf_prior$Tc)), big.mark = ","), "\n")
cat("Seeing (Rk = 0) mean:    $",
    format(round(mean(cf_see$Tc)),   big.mark = ","),
    "  E[U | Rk=0] =", round(mean(cf_see$U), 2), "\n")
cat("Doing  do(Rk = 0) mean:  $",
    format(round(mean(cf_do$Tc)),    big.mark = ","),
    "  E[U] =", round(mean(cf_do$U), 2), "\n")
```

Conditioning propagates the evidence back through the shared cause, the
expected site-condition indicator falls from its prior of 0.5 to about 0.1, so
\emph{seeing} that rework did not occur pulls the foundation cost down as well
and yields a markedly lower expected total than \emph{doing}. Because
\code{prob\_net\_learn()} conditions by rejection sampling over whole joint
draws, it captures this backdoor propagation that a clamp-and-forward update
cannot; this is precisely where observation and intervention must be
distinguished, and where the \emph{do}-operator becomes indispensable.

## Ranking Risks by Intervention

Repeating the intervention for each root risk and measuring the reduction in
project-cost variance yields a principled, network-level mitigation
ranking, a causal analog of the Tornado chart of
Section \ref{sec-uncertainty}:

```{r net-importance, fig.cap="Risk importance for the probabilistic network: the reduction in total project-cost variance achieved by intervening to eliminate each root risk. Taller bars indicate higher-priority mitigation targets."}
base_var <- var(sim_net$I)

do_A <- prob_net_sim(prob_net_update(net,
  remove_links = data.frame(source = "A", target = "C"),
  update_distributions = list(
    C = list(type = "normal", mean = 15000, sd = 3000))),
  num_samples = 10000)

importance <- c(
  "Risk-1" = base_var - var(do_A$I),
  "Risk-2" = base_var - var(do_net$I)
)
barplot(sort(importance), horiz = TRUE, col = "steelblue",
  main = "Risk Importance: Variance Eliminated by do(risk = 0)",
  xlab = "Reduction in Project-Cost Variance")
```

The probabilistic network module is newer than the analytical core, and its
API may still evolve in future releases.


# Dependency Structure Analysis {#sec-dsm}

Building on foundational DSM
literature \citep{Steward1981, Browning2001}, the DSM module quantifies
structural coupling between project tasks through shared resources and
risks. This section revisits the probabilistic network of
Section \ref{sec-network}, with its two risks (Risk-1, Risk-2), three
resources (Resource-1, Resource-2, Resource-3), and three tasks (Task-1,
Task-2, Task-3), from a structural rather than causal perspective. The
network's cost-aggregation edges route each resource's cost to exactly one
task for simulation simplicity. In practice, resources and risk exposure
are shared more broadly, and the assignment below captures that broader
structural coupling.

## Design Structure Matrices

The **Parent DSM** is derived from the Resource--Task matrix $\mathbf{S}$
(rows = resources, columns = tasks) as $\mathbf{P} = \mathbf{S}^\top\mathbf{S}$.
Off-diagonal entry $P_{jk}$ counts the number of resources shared between
tasks $j$ and $k$, the structural pathway through which disruptions propagate
\citep{Govan2016}.

```{r dsm-S}
S <- matrix(c(
  1, 0, 0,
  0, 1, 0,
  1, 1, 1
), nrow = 3, ncol = 3, byrow = TRUE)
rownames(S) <- c("Resource-1", "Resource-2", "Resource-3")
colnames(S) <- c("Task-1", "Task-2", "Task-3")
```

```{r parent-dsm, fig.cap="Parent DSM heatmap for the probabilistic network example above. Every task pair shares Resource-3, the common structural bottleneck."}
p <- parent_dsm(S)
plot(p)
```

The **Grandparent DSM** adds a risk layer via the Risk--Resource matrix
$\mathbf{R}$ (rows = risks, columns = resources), tracing dependencies from
risks through resources to tasks as
$\mathbf{G} = (\mathbf{R}\mathbf{S})^\top(\mathbf{R}\mathbf{S})$.
Off-diagonal entry $G_{jk} = \sum_i (\mathbf{R}\mathbf{S})_{ij}(\mathbf{R}\mathbf{S})_{ik}$
scores the risk exposure that tasks $j$ and $k$ share through the resource
chain, where $(\mathbf{R}\mathbf{S})_{ij}$ counts the risk-to-task paths from
risk $i$ to task $j$. Entries therefore weight a shared risk by the number of
common resources carrying it, and reduce to a plain count of shared risks only
when $\mathbf{R}\mathbf{S}$ is binary. This is a direct operationalization
of the structural network measures developed by \citet{Govan2020}.

```{r dsm-R}
R_mat <- matrix(c(
  1, 0, 1,
  0, 1, 1
), nrow = 2, ncol = 3, byrow = TRUE)
rownames(R_mat) <- c("Risk-1", "Risk-2")
colnames(R_mat) <- c("Resource-1", "Resource-2", "Resource-3")
```

```{r grandparent-dsm, fig.cap="Grandparent DSM heatmap for the probabilistic network example above. Task pairs inherit shared risk exposure through Resource-3."}
g <- grandparent_dsm(S, R_mat)
plot(g)
```

Resource-3 is the dominant coupling driver: every task pair shares it, and
both Risk-1 and Risk-2 propagate through it, concentrating risk exposure
across the whole project. It is therefore the prime candidate for
contingency buffering or, if feasible, decoupling by assigning a dedicated
resource to at least one of the three tasks.


# Model Context Protocol Integration {#sec-agent}

\pkg{PRA} exposes all analytical functions as tools callable
by large-language-model (LLM) agents through the Model Context Protocol (MCP)
\citep{MCP2024}. \code{pra\_mcp\_server()} starts a server that advertises each
tool, Monte Carlo simulation, EVM, noisy-OR risk, learning curves, DSM, and
the rest, so that MCP-compatible clients can
invoke them as native tools. The server is registered in Claude Code with a
single shell command:

\begin{verbatim}
claude mcp add -s project pra -- Rscript -e "PRA::pra_mcp_server()"
\end{verbatim}

For Claude Desktop, add the following block to the MCP configuration file:

\begin{verbatim}
{
  "mcpServers": {
    "pra": { "command": "Rscript",
             "args": ["-e", "PRA::pra_mcp_server()"] }
  }
}
\end{verbatim}

The tool definitions are built once with \code{ellmer::tool()} \citep{ellmer}
and served by \pkg{mcptools}. Each tool validates its JSON inputs and returns a
plain-text summary for the LLM together with, where applicable, a rich HTML
result with inline plots for clients that render it.

The MCP layer is strictly optional: every analytical function operates without
any optional dependency. Only \code{pra\_mcp\_server()} requires \pkg{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}.


# Illustrative Case Study {#sec-casestudy}

This section demonstrates an integrated, end-to-end project risk analysis
for the \code{building\_project} dataset introduced in
Section \ref{sec-uncertainty}, progressing through uncertainty
quantification, risk-driver ranking, performance monitoring, forecasting cost
at completion, noisy-OR risk updating, dependency structure analysis, and
network cost propagation in a single \proglang{R} session. The workflow
reuses objects built in the preceding sections (for example the closed-form
task moments and the probabilistic network), illustrating how the modules
compose within one analysis.

## Step 1: Characterize Task Uncertainty

Combine the rapid analytical estimate from the Second Moment Method, reusing
the closed-form triangular moments \code{task\_means} and \code{task\_vars}
from Section \ref{sec-uncertainty}, with the full simulated distribution from
Monte Carlo simulation. Both are given the dataset correlation matrix, so the
analytical and simulated views rest on the same dependence assumption:

```{r cs-mcs}
bp <- building_project
set.seed(42)

smm_cs <- smm(task_means, task_vars, bp$cor_mat)
sim     <- mcs(10000, bp$task_distributions, bp$cor_mat)
res     <- contingency(sim, phigh = 0.80, pbase = 0.50)

cat("SMM total (correlated): mean", round(smm_cs$total_mean, 1),
    "SD", round(smm_cs$total_std, 2), "weeks\n")
cat("P50 schedule:",
    round(quantile(sim$total_distribution, 0.50), 1), "weeks\n")
cat("P80 schedule:",
    round(quantile(sim$total_distribution, 0.80), 1), "weeks\n")
cat("Contingency (P80-P50):", round(res, 2), "weeks\n")
```

The P50 estimate is the schedule a planner would commit to with even odds of
meeting it; the contingency buffer above it is the additional duration needed
to raise that confidence to 80\%, and is the reserve a project manager would
add to the baseline schedule. Correlation matters here in practical terms:
treating the six work packages as independent shrinks the same reserve to
3.98 weeks, because independent overruns partially cancel. Shared crews,
weather, and a common supply chain make that assumption optimistic, and the
correlated reserve is the defensible one.

## Step 2: Rank Risk Drivers

\code{sensitivity()} ranks the tasks by their share of total schedule
variance, focusing mitigation on the dominant drivers:

```{r cs-sens}
drivers <- sensitivity(bp$task_distributions)
names(drivers) <- bp$task_names
print(round(sort(drivers, decreasing = TRUE), 3))
```

The Structural Frame accounts for the largest single share of total schedule
variance, but it does not dominate: together with the two tasks ranked
immediately behind it, it accounts for roughly 70\% of the total, and mitigation
effort is best spread across all three. The two smallest contributors, at about
8\% each, are the least productive targets.

## Step 3: Monitor Earned Value at Mid-Project

Compute planned value, earned value, and actual cost to date for the
case-study project:

```{r cs-evm}
pv_cs <- pv(bp$bac, bp$schedule, bp$time_period)
ev_cs <- ev(bp$bac, bp$actual_per_complete)
ac_cs <- ac(bp$actual_costs, bp$time_period, cumulative = FALSE)

cat("SPI:", round(spi(ev_cs, pv_cs), 3),
    " CPI:", round(cpi(ev_cs, ac_cs), 3), "\n")
cat("EAC (typical): $", format(round(eac(bp$bac,
    method = "typical", cpi = cpi(ev_cs, ac_cs))), big.mark = ","), "\n")
```

An SPI and CPI both below 1.0 indicate the case-study project is simultaneously
behind schedule and over budget at this reporting point, and the typical-method
EAC translates that cost inefficiency into a revised completion-cost forecast
above the original budget at completion.

## Step 4: Forecast Cost at Completion

The performance indices above translate into a completion forecast. The
Estimate to Complete prices the remaining work at the cost efficiency achieved
so far, the To-Complete Performance Index gives the efficiency the remaining
work would have to achieve to finish within the original budget, and the
Variance at Completion states the expected shortfall:

```{r cs-forecast}
cpi_cs  <- cpi(ev_cs, ac_cs)
eac_cs  <- eac(bp$bac, method = "typical", cpi = cpi_cs)

cat("ETC:  $", format(round(etc(bp$bac, ev_cs, cpi = cpi_cs)),
    big.mark = ","), "\n")
cat("TCPI:", round(tcpi(bp$bac, ev_cs, ac_cs), 3), "\n")
cat("VAC:  $", format(round(vac(bp$bac, eac_cs)), big.mark = ","), "\n")
```

The TCPI is only slightly above 1.0, so the budget is still recoverable: the
remaining work has to run about one percent more efficiently than planned, which
is a realistic target rather than the kind of step change that signals an
unrecoverable overrun. Left uncorrected, the cost efficiency achieved so far
projects the modest overrun reported by the Variance at Completion.

## Step 5: Update Risk Probability After Site Inspection

Recompute the noisy-OR prior probability of schedule delay, then update it
on the site inspection observation:

```{r cs-bayes}
prior_r <- risk_prob(bp$cause_probs, bp$risks_given_causes,
                     bp$risks_given_not_causes)
post_r  <- risk_post_prob(bp$cause_probs, bp$risks_given_causes,
                          bp$risks_given_not_causes, bp$observed_causes)
cat("Prior P(delay):", round(prior_r, 3),
    "-> Posterior P(delay):", round(post_r, 3), "\n")
```

The shift from prior to posterior quantifies how much the site inspection
evidence should move the project team's confidence in a schedule delay,
information that feeds directly into whether the contingency reserve computed
in Step 1 needs to be revisited.

## Step 6: Propagate Cost Through the Network

Use the probabilistic network \code{net} from
Section \ref{sec-network} to propagate cost uncertainty to the project total,
then condition on the observation that Risk-2 (Technical Complexity) did not
occur:

```{r cs-net}
cat("Prior mean project cost: $",
    format(round(mean(sim_net$I)), big.mark = ","), "\n")
cat("Prior P80 project cost:  $",
    format(round(quantile(sim_net$I, 0.80)), big.mark = ","), "\n")
cat("Posterior mean (Risk-2 = 0): $",
    format(round(mean(learn_net$I)), big.mark = ","), "\n")
```

Conditioning on the resolved risk lowers the expected project cost, closing
the loop from uncertainty characterization to evidence-based updating.

## Step 7: Identify Coupled Tasks via DSM

```{r cs-dsm, fig.cap="Grandparent DSM for the case-study project, showing which task pairs share the most risk exposure through common resources."}
g_cs <- grandparent_dsm(bp$resource_task, bp$risk_resource)
plot(g_cs)
```

The Grandparent DSM reveals which task pairs share risk exposure through
common resources, informing targeted contingency allocation and resource
decoupling decisions.


# Summary {#sec-summary}

\pkg{PRA} is an \proglang{R} package whose primary contribution is being the first package to unify the foundational methods of quantitative project risk analysis in a single, open-source framework of ten integrated analytical modules: SMM, MCS, contingency analysis, sensitivity analysis, EVM (11 metrics), noisy-OR risk inference and updating, cost risk distributions, sigmoidal learning curves, correlation matrix construction, and design structure matrices. The eleventh module applies probabilistic network methods to project risk, with forward simulation and Bayesian conditioning over a network of risks, resources, and tasks; general-purpose Bayesian network inference in \proglang{R} remains the province of \pkg{bnlearn} and \pkg{gRain}.

\pkg{PRA} also exposes every analytical function as a tool over the Model Context Protocol, so that LLM agents can drive the package without any embedded model, the first project risk analysis package to do so. All exported functions are annotated against the rOpenSci SRR general standards, with a test suite verifying correctness, edge-case handling, and numerical stability across three operating systems. The package is licensed under MIT and the source is available at \url{https://github.com/paulgovan/PRA}; documentation and a companion online book are at \url{https://paulgovan.github.io/PRA/}.

## Limitations and Future Work

Several limitations apply. First, \pkg{PRA} currently propagates uncertainty at the task level and does not yet model schedule logic, critical path, activity precedence, and resource leveling. Full schedule-network risk analysis, in which schedule Monte Carlo runs over an activity-precedence network, such as Program Evaluation and Review Technique / Critical Path Method (PERT/CPM) to identify the critical path and criticality indices under uncertainty, is a planned future direction that would extend \pkg{PRA} from task-level to network-level schedule risk. Second, \code{mcs()} induces task correlation through the Cholesky
decomposition of the correlation matrix, which preserves every marginal
exactly, but which fixes the dependence structure to the Gaussian family: tail dependence stronger
than a multivariate normal implies, such as tasks that fail together only in
extreme scenarios, cannot be represented. Richer dependence structures are
planned. Third,
the probabilistic network module is newer than the analytical core: its API may
still evolve. Fourth,
the package currently accepts data as \proglang{R} vectors and lists. Direct
import from project management file formats (Microsoft Project, Primavera P6,
CSV exports from Jira or Asana) is not supported and is planned for a future
release.

# Computational details {-}

The results in this article were produced using \proglang{R} \citep{RCore}
with the \pkg{PRA} package (version 0.5.0) and its hard dependencies
\pkg{mc2d} \citep{mc2d} and \pkg{minpack.lm} \citep{Elzhov2016}. The optional
Model Context Protocol server additionally uses \pkg{ellmer} \citep{ellmer},
\pkg{mcptools}, and \pkg{jsonlite}. \proglang{R} itself and all packages used
are available from the CRAN at \url{https://CRAN.R-project.org/}. A random seed of 42 is set at the start of the session so that the simulation output is reproducible.

# References {-}
