---
title: "An AI Agent Framework for Quantitative Project Risk Analysis: Integrating Large Language Models with Validated Probabilistic Tools"
author:
  - name: Paul Govan
    email: paul.govan2@gmail.com
    orcid: 0000-0002-1821-8492
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: null
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
bibliography: manuscript.bib
csl: https://www.zotero.org/styles/risk-analysis
abstract: |
  Quantitative methods for project risk analysis — including Monte Carlo
  simulation, Earned Value Management, and Bayesian inference — are
  well-established in theory but underutilized in practice due to their
  technical complexity and fragmented tooling. Large language model (LLM)
  agents capable of calling domain-validated code functions offer a promising
  path toward closing this gap. This paper presents an AI agent framework,
  implemented as the open-source R package PRA, that exposes a validated suite
  of project risk analysis functions as tools callable by LLMs through a
  three-mode routing architecture: deterministic slash commands for immediate
  reproducible results, LLM tool-calling for flexible natural-language
  queries, and Retrieval-Augmented Generation (RAG) for conceptual questions
  grounded in domain literature. We evaluate the framework across 25 scenarios
  in three tiers of complexity using the llama3.2 model (3.2B parameters,
  local inference). With RAG enabled, the agent achieves 72.0% overall
  tool-selection accuracy (87.5% on single-tool, 55.6% on multi-tool chains,
  75.0% on open-ended queries), compared to 68.0% without RAG. RAG
  substantially improves multi-tool chain accuracy (+11.1%) and open-ended
  response quality (+12.5%) at the cost of reduced single-tool precision
  (−12.5%), a trade-off fully mitigated by the slash command interface for
  deterministic workflows. The framework is available on CRAN, supports the
  Model Context Protocol for interoperability with Claude, Claude Code, and
  other AI clients, and runs fully offline via Ollama — requiring no cloud API
  keys or data transmission.
keywords:
  - project risk analysis
  - large language models
  - retrieval-augmented generation
  - Monte Carlo simulation
  - earned value management
  - AI agents
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# 1. Introduction

Quantitative risk analysis is central to sound project management. Methods
such as Monte Carlo simulation (MCS), Earned Value Management (EVM), and
Bayesian risk inference provide project managers and cost engineers with
principled tools for quantifying schedule uncertainty, diagnosing cost
overruns, and updating risk beliefs as projects evolve [@Damnjanovic2020;
@Vose2008; @Fleming2010]. Yet surveys of project management practice
consistently show that quantitative risk techniques remain underutilized, with
most organizations relying on qualitative risk registers rather than
probabilistic analysis [@Damnjanovic2020]. The barriers are well understood:
these methods require statistical expertise to apply correctly, their
implementations are scattered across commercial tools (Palisade \@RISK,
Primavera Risk Analysis) that are expensive and not reproducible, and
practitioners often lack the time or training to connect the right tool to
each type of problem.

Large language models (LLMs) have emerged as a potentially transformative
interface for domain expertise. Through *tool calling* — the ability to
invoke external functions in response to natural language queries —
LLMs can act as intelligent intermediaries that translate a practitioner's
intent into a validated computational pipeline [@Patil2023]. Rather than
requiring users to know which function to call and how to parameterize it, the
agent selects and invokes the appropriate tool automatically. This capability
has been demonstrated in general software engineering contexts, but its
application to specialized engineering analysis domains — particularly project
risk — remains nascent.

Retrieval-Augmented Generation (RAG) complements tool calling by grounding
agent responses in domain-specific knowledge [@Lewis2020]. For conceptual
queries ("what is the difference between SPI and CPI?" or "how does Bayesian
inference apply to cost risk?"), RAG retrieves relevant passages from a curated
knowledge base and presents them as context for the LLM's response, reducing
hallucination and improving source attribution [@RAGReview2024; @CHI2025RAG].
Together, tool calling and RAG address the two primary failure modes of
general-purpose LLMs in engineering contexts: incorrect tool selection or
parameterization, and fabrication of quantitative claims.

This paper presents a novel AI agent framework for quantitative project risk
analysis, implemented as part of the open-source R package PRA [@PRA] and
available on the Comprehensive R Archive Network (CRAN). The framework makes
four contributions:

1. A **three-mode routing architecture** — deterministic slash commands, LLM
   tool-calling, and RAG — that accommodates different user workflows and
   reliability requirements within a single interface.

2. A **validated analytical tool layer** comprising 14 LLM-callable tools and
   9 deterministic commands, wrapping a peer-reviewed suite of project risk
   analysis functions that cover MCS, SMM, EVM, contingency analysis,
   sensitivity analysis, Bayesian inference, learning curves, and dependency
   structure matrices.

3. **Model Context Protocol (MCP) integration** via the `mcptools` package
   [@mcptools], enabling PRA tools to be called natively from Claude, Claude
   Code, and any MCP-compatible AI client without additional configuration.

4. A **systematic evaluation** of agent performance across 25 scenarios in
   three tiers of complexity, providing the first published benchmark of an
   LLM agent for quantitative project risk analysis.

The remainder of the paper is organized as follows. Section 2 reviews the
relevant background. Section 3 describes the framework architecture. Section 4
presents the evaluation methodology and results. Section 5 discusses
implications and limitations. Section 6 concludes.

# 2. Background

## 2.1 Quantitative Project Risk Methods

Project risk analysis encompasses a range of techniques for quantifying
uncertainty in schedule, cost, and performance. Monte Carlo simulation samples
from probability distributions assigned to individual tasks and aggregates them
over many iterations to produce a project-level distribution from which
contingency reserves and risk-adjusted completion probabilities can be derived
[@Vose2008]. The Second Moment Method (SMM) provides a computationally cheaper
analytical approximation, propagating means and variances through the task
network without simulation [@Damnjanovic2020]. Earned Value Management tracks
project performance against a baseline cost and schedule plan using a set of
standardized metrics — Cost Performance Index (CPI), Schedule Performance
Index (SPI), Estimate at Completion (EAC), and others — defined in the
ANSI/EIA-748 standard [@Fleming2010]. Bayesian risk inference applies the law
of total probability and Bayes' theorem to estimate and update risk event
probabilities as observations arrive [@Gelman2013]. Design Structure Matrices
(DSM) model task coupling and risk propagation in complex project networks
[@Browning2001].

Despite their theoretical grounding, these methods see limited practical
adoption. Each requires a different data format, a different software tool, and
a different conceptual model. Commercial tools such as Palisade \@RISK and
Oracle Primavera Risk Analysis cover subsets of this space but are expensive,
proprietary, and not amenable to reproducible research workflows. The PRA
R package [@PRA] was developed to consolidate these methods in a single,
open-source, peer-reviewed implementation, but even a well-designed
programmatic interface requires users to understand which function addresses
their question and how to format its inputs.

## 2.2 LLM Tool Calling

The ability for LLMs to invoke external functions — variously called tool
calling, function calling, or code execution — has emerged as a key capability
for building AI agents that act on the world rather than only describing it
[@Patil2023]. When a user's query contains structured data that can be
processed by a validated function, the LLM selects the appropriate tool from
a registered set, constructs the call arguments, invokes the function, and
incorporates the result into its response. This paradigm separates the natural
language understanding task (which LLMs perform well) from the computation task
(which validated code performs reliably), combining the strengths of both.

The `ellmer` R package [@ellmer] provides a unified interface for LLM tool
calling in R, supporting Ollama local models, OpenAI, Anthropic, and other
backends through a consistent `tool()` registration API. Tools defined with
`ellmer::tool()` can be registered on a chat object and reused across
conversational turns, making them a natural fit for the PRA agent framework.

## 2.3 Retrieval-Augmented Generation

Retrieval-Augmented Generation augments LLM responses with passages retrieved
from an external knowledge base, enabling models to answer domain-specific
questions without requiring that knowledge to be encoded in model weights
[@Lewis2020]. A retrieval system indexes a corpus of documents as vector
embeddings; at query time, the most semantically similar chunks are retrieved
and inserted into the LLM's context window as grounding evidence. Reviews of
RAG systems find that hybrid retrieval — combining dense vector similarity
search with sparse keyword matching (BM25) — generally outperforms either
approach alone, particularly for domain-specific terminology [@RAGReview2024].

For engineering decision support, the transparency and source attribution
properties of RAG are especially important. @CHI2025RAG find that domain
experts are more willing to trust AI-generated responses when retrieved source
passages are visible, allowing them to verify claims against primary literature.
In the project risk context, where practitioners may need to defend their
analyses to stakeholders, the ability to cite a specific EVM standard or
simulation methodology is a meaningful quality criterion.

## 2.4 AI-Assisted Engineering Decision Support

AI-assisted decision support has been explored across a range of engineering
risk contexts. @DeLLMa2024 demonstrate that LLMs can support principled
decision making under uncertainty by combining probabilistic reasoning with
chain-of-thought prompting. @FMEA2025 show that LLM integration accelerates
Failure Mode and Effects Analysis (FMEA) — a structured risk identification
technique closely related to project risk assessment — while maintaining
accuracy comparable to expert analysts. @SafetyRAG2025 combine RAG with LLMs
to provide automated safety risk guidance, demonstrating that domain-grounded
generation substantially outperforms base LLM responses on compliance-critical
queries.

A common theme across these works is that LLM agents are most effective when
tightly coupled with validated domain tools or knowledge sources rather than
relying on the model's parametric knowledge alone. The PRA agent framework
adopts this design principle explicitly: the LLM is responsible for
understanding user intent and selecting tools, while all quantitative
computation is delegated to validated R functions.

# 3. Framework Design

## 3.1 Architecture Overview

The PRA agent framework is organized in three layers (Figure 1):

- **Interface layer**: The user interacts through one of three surfaces — a
  programmatic R chat object (`pra_chat()`), a Shiny web application
  (`pra_app()`), or any MCP-compatible AI client via `pra_mcp_server()`.

- **Routing layer**: User input is classified into one of three modes —
  *command*, *tool*, or *rag* — by a heuristic classifier that examines the
  input for slash prefixes, numerical data patterns, and conceptual question
  markers.

- **Execution layer**: Classified inputs are dispatched to the appropriate
  handler: the command registry (deterministic), the LLM tool-calling loop
  (probabilistic), or the RAG retrieval pipeline (knowledge-grounded).

This architecture separates concerns cleanly: reliability-critical workflows
use the command path, exploratory analysis uses the tool-calling path, and
conceptual learning uses the RAG path. A single entry point handles all three
modes, so users do not need to choose a different application for each workflow.

```{r fig1-architecture, fig.width=7, fig.height=5, fig.cap="Figure 1. Three-layer architecture of the PRA AI agent framework. User input enters through the interface layer, is classified by the routing layer into one of three modes, and dispatched to the appropriate execution handler.", echo=FALSE}
library(grid)
grid.newpage()

# Color palette
header_bg  <- "#d0d8e8"
box_bg     <- "#f7f9fc"
border_col <- "#4a5568"
arrow_col  <- "#4a5568"
text_col   <- "#1a202c"
label_col  <- "#2d3748"

# Layer y-positions (top, middle, bottom) and height
layer_tops <- c(0.93, 0.62, 0.31)
layer_h    <- 0.26
header_h   <- 0.055

# Column x-positions (left edge) and width
col_xs <- c(0.04, 0.37, 0.70)
col_w  <- 0.28

layer_labels <- c("INTERFACE LAYER", "ROUTING LAYER", "EXECUTION LAYER")
box_titles <- list(
  c("pra_chat()\nR Console", "pra_app()\nShiny Web", "MCP Clients\nClaude / Claude Code"),
  c("/command prefix\n→ COMMAND", "Numeric patterns\n→ TOOL", "Conceptual query\n→ RAG"),
  c("Command Registry\n9 slash commands\nDeterministic", "LLM Tool Loop\n14 ellmer tools\nProbabilistic", "RAG Pipeline\nDuckDB + BM25/VSS\nKnowledge-grounded")
)

for (layer in 1:3) {
  top <- layer_tops[layer]

  # Layer header bar
  grid.rect(x = 0.5, y = top - header_h / 2,
            width = 0.94, height = header_h,
            gp = gpar(fill = header_bg, col = border_col, lwd = 1),
            just = "centre")
  grid.text(layer_labels[layer],
            x = 0.5, y = top - header_h / 2,
            gp = gpar(fontsize = 8, fontface = "bold", col = label_col))

  # Three boxes per layer
  box_top <- top - header_h
  box_h   <- layer_h - header_h

  for (col in 1:3) {
    bx <- col_xs[col] + col_w / 2
    by <- box_top - box_h / 2

    grid.rect(x = bx, y = by,
              width = col_w - 0.02, height = box_h,
              gp = gpar(fill = box_bg, col = border_col, lwd = 0.8),
              just = "centre")
    grid.text(box_titles[[layer]][col],
              x = bx, y = by,
              gp = gpar(fontsize = 7, col = text_col),
              just = "centre")
  }
}

# Downward arrows between layers (three per gap)
for (col in 1:3) {
  ax <- col_xs[col] + col_w / 2
  for (gap in 1:2) {
    y_top    <- layer_tops[gap]   - layer_h + 0.005
    y_bottom <- layer_tops[gap + 1] - 0.005
    grid.lines(x = c(ax, ax), y = c(y_top, y_bottom),
               gp = gpar(col = arrow_col, lwd = 1.2),
               arrow = arrow(length = unit(0.08, "inches"), type = "closed",
                             angle = 20))
  }
}
```

## 3.2 Input Routing

The routing classifier (`route_input()`) operates as follows. If the input
begins with a forward slash, it is classified as a *command* and dispatched to
the command registry without involving the LLM. Otherwise, the input is scanned
for numerical patterns: distribution specifications (e.g., `normal(10, 2)`),
JSON arrays, monetary amounts (`$500,000`), EVM keywords (`BAC`, `schedule`),
and structured parameter assignments (`mean=`, `costs=`). If any pattern
matches, the input is classified as a *tool* call. If no numerical patterns are
found, the input is classified as *rag* and the knowledge base is queried for
relevant context.

The LLM is involved in both the *tool* and *rag* modes but plays different
roles. In *tool* mode, the system prompt encodes tool-selection heuristics —
for example, "if the user provides distribution specifications, call
`mcs_tool`, then `contingency_tool` and `sensitivity_tool`" — and the LLM
selects and invokes the appropriate tools. In *rag* mode, retrieved passages
are inserted into the context window and the LLM synthesizes a response with
source attribution.

The classification heuristic is intentionally lightweight: it trades recall
for precision, occasionally routing tool-call queries to RAG when numerical
patterns are absent, but never routing conceptual questions through the
tool-calling path (which would produce spurious function invocations). Users
can always override classification by using a slash command directly.

## 3.3 Analytical Tool Layer

The tool layer comprises 14 `ellmer::tool()` definitions, each wrapping one
or more PRA analytical functions. Table 1 summarizes the tools.

```{r table1-tools, echo=FALSE}
tools_df <- data.frame(
  Tool = c(
    "`mcs_tool`", "`smm_tool`", "`contingency_tool`", "`sensitivity_tool`",
    "`evm_analysis_tool`", "`risk_prob_tool`", "`risk_post_prob_tool`",
    "`cost_pdf_tool`", "`cost_post_pdf_tool`",
    "`fit_and_predict_sigmoidal_tool`",
    "`parent_dsm_tool`", "`grandparent_dsm_tool`"
  ),
  `R Function(s)` = c(
    "`mcs()`", "`smm()`", "`contingency()`", "`sensitivity()`",
    "EVM suite", "`risk_prob()`", "`risk_post_prob()`",
    "`cost_pdf()`", "`cost_post_pdf()`",
    "`fit_sigmoidal()` + `predict_sigmoidal()`",
    "`parent_dsm()`", "`grandparent_dsm()`"
  ),
  Description = c(
    "Monte Carlo simulation with task distributions and correlations",
    "Second Moment Method analytical project estimate",
    "Contingency reserve derived from simulation percentile",
    "Variance contribution by task (tornado chart)",
    "Full ANSI/EIA-748 dashboard: 11 EVM metrics",
    "Bayesian prior risk probability from root causes",
    "Bayesian posterior probability after observed evidence",
    "Prior cost risk distribution (Beta-Binomial)",
    "Posterior cost distribution after observations",
    "Sigmoidal learning curve fit and completion forecast",
    "Resource-task dependency matrix (parent DSM)",
    "Risk-resource-task dependency matrix (grandparent DSM)"
  ),
  check.names = FALSE
)
knitr::kable(tools_df,
  caption = "Table 1. LLM-callable tools in the PRA agent registry.",
  align = c("l", "l", "l")
)
```

All tool inputs are passed as JSON strings (matching the format LLMs naturally produce),
and a fault-tolerant JSON parser (`parse_json()`) handles common LLM output
artifacts such as smart quotes, trailing commas, and missing array brackets.

Tool outputs are designed for dual use: plain-text summaries are returned to
the LLM for incorporation into conversational responses, while rich HTML
outputs (inline charts and formatted tables) are rendered in the Shiny
interface. Full simulation results (e.g., the 10,000-element distribution
vector from `mcs_tool`) are stored in a package-level environment
(`.pra_agent_env`) rather than returned verbatim, which keeps the LLM's
context window manageable and enables tool chaining — the `contingency_tool`
and `sensitivity_tool` read from `.pra_agent_env$last_mcs` rather than
requiring the user to re-specify the simulation.

In addition to the LLM-callable tools, nine deterministic slash commands
(Table 2) provide the same analytical capabilities through a structured
argument syntax that bypasses the LLM entirely. Commands use a
bracket-aware argument parser (`parse_command_args()`) that handles nested
JSON values in command arguments without regex fragility. The commands produce
identical numerical results to the tools but execute in milliseconds and
guarantee reproducibility.

```{r table2-commands, echo=FALSE}
cmds_df <- data.frame(
  Command = c(
    "`/mcs`", "`/smm`", "`/contingency`", "`/sensitivity`",
    "`/evm`", "`/risk`", "`/risk_post`", "`/learning`", "`/dsm`"
  ),
  `Equivalent Tool` = c(
    "`mcs_tool`", "`smm_tool`", "`contingency_tool`", "`sensitivity_tool`",
    "`evm_analysis_tool`", "`risk_prob_tool`", "`risk_post_prob_tool`",
    "`fit_and_predict_sigmoidal_tool`", "`parent_dsm_tool`"
  ),
  `Example Syntax` = c(
    '/mcs n=10000 tasks=[{"type":"normal","mean":10,"sd":2}]',
    "/smm means=[10,15,20] vars=[4,9,16]",
    "/contingency phigh=0.95 pbase=0.50",
    '/sensitivity tasks=[{"type":"normal","mean":10,"sd":2}]',
    "/evm bac=500000 schedule=[0.2,0.4,0.6,0.8,1.0] period=3 complete=0.55 costs=[90000,195000,310000]",
    "/risk causes=[0.3,0.2] given=[0.8,0.6] not_given=[0.2,0.4]",
    "/risk_post causes=[0.3,0.2] given=[0.8,0.6] not_given=[0.2,0.4] observed=[1,null]",
    "/learning model=logistic x=[1,2,3,4,5,6,7,8] y=[2,8,25,50,72,88,95,98]",
    "/dsm matrix=[[1,1,0],[0,1,1],[0,0,1]]"
  ),
  check.names = FALSE
)
knitr::kable(cmds_df,
  caption = "Table 2. Deterministic slash commands. Each command executes the corresponding analytical function directly without LLM involvement.",
  align = c("l", "l", "l")
)
```

## 3.4 RAG Knowledge Layer

The knowledge base consists of six curated Markdown documents (~19 KB total)
covering the core project risk analysis methods:

- `mcs_methods.md` — distribution selection, correlation, convergence criteria
- `evm_standards.md` — ANSI/EIA-748 metrics and EAC forecasting methods
- `bayesian_risk.md` — prior/posterior probability, Beta-Binomial model
- `sensitivity_contingency.md` — variance decomposition, reserve calculation
- `learning_curves.md` — logistic, Gompertz, and Pearl sigmoidal models
- `pra_functions.md` — comprehensive function reference

Documents are chunked using Markdown-aware splitting via `ragnar::markdown_chunk()`
and stored in a DuckDB database through the `ragnar` package [@ragnar]. At
query time, `retrieve_context()` performs hybrid retrieval combining BM25
keyword matching and vector similarity search over 768-dimensional embeddings
produced by the `nomic-embed-text` model running locally via Ollama [@Ollama2023].
The top-3 chunks are returned with source filenames for attribution. The
database is built on first use and cached locally, so subsequent calls operate
from the cached store without re-embedding.

The knowledge base is user-extensible via `add_documents()`, which ingests
additional `.md` or `.txt` files and rebuilds the index. This allows
organizations to add project-specific risk registers, standards documents, or
historical performance data as grounding context.

## 3.5 MCP Integration

The Model Context Protocol [@MCP2024] defines a standard interface through
which AI clients can discover and invoke external tools via JSON-RPC 2.0
over stdio or HTTP/SSE transports. The `pra_mcp_server()` function launches
PRA as an MCP server, exposing the same 14 tools as the `ellmer` layer through
this protocol:

```r
pra_mcp_server()  # starts stdio MCP server
```

The implementation delegates to `mcptools::mcp_server(tools = pra_tools())`
[@mcptools], reusing the existing `ellmer::tool()` definitions without
duplication. Claude Code, Claude Desktop, and other MCP clients can connect
by adding PRA to their MCP server configuration. A project-level
`.claude/mcp.json` file is included in the package repository for zero-
configuration Claude Code integration:

```json
{
  "mcpServers": {
    "pra": {
      "command": "Rscript",
      "args": ["-e", "PRA::pra_mcp_server()"]
    }
  }
}
```

MCP integration decouples the PRA tool layer from the R-based agent
orchestration: any MCP-compatible AI client gains access to validated project
risk analysis functions without requiring R session management or Shiny.

## 3.6 Implementation

The framework is implemented as part of the PRA R package (v0.4.0), available
on CRAN. The agent components span approximately 2,400 lines of R across three
source files: `R/chat.R` (agent initialization, system prompt, MCP server),
`R/rag.R` (knowledge base construction and retrieval), and `R/tools.R` (tool
wrappers, command registry, and input router). Hard dependencies are limited to
`mc2d`, `minpack.lm`, and `stats`; all AI components are in `Suggests` and
loaded on demand.

By default, PRA uses locally running Ollama models [@Ollama2023] — specifically
`llama3.2` (3.2B parameters) for generation and `nomic-embed-text` (137M
parameters) for embeddings. This design is deliberate: local inference
preserves data privacy (no project data leaves the user's machine), eliminates
per-query API costs, and operates without network connectivity. Users who
prefer higher accuracy for complex tool-calling scenarios can supply a
pre-configured `ellmer` chat object using a cloud model (e.g.,
`ellmer::chat_anthropic(model = "claude-opus-4-7")`); the agent architecture
is model-agnostic.

The Shiny application (`pra_app()`) provides a browser-based interface with
streaming responses, inline visualizations from tool outputs, and example
prompt chips that demonstrate all three input modes. It is launched with a
single R call and requires no web server configuration.

# 4. Evaluation

## 4.1 Design

We evaluate the agent's ability to correctly identify and invoke the
appropriate analytical tools in response to user queries, using a benchmark
of 25 scenarios spanning three tiers of complexity:

- **Tier 1 — Single-tool** (8 scenarios): Each scenario requires exactly one
  tool call with correctly specified parameters. Examples include "Run a Monte
  Carlo simulation with 10,000 iterations for a 3-task project with Normal(10,2),
  Triangular(5,10,15), and Uniform(8,12) distributions" and "Fit a logistic
  learning curve to this data: x = [1..8], y = [2,8,25,50,72,88,95,98]."

- **Tier 2 — Multi-tool chain** (9 scenarios): Each scenario requires two or
  more tool calls in sequence. Examples include MCS followed by contingency
  reserve calculation, full EVM analysis with forecast interpretation, and
  Bayesian prior computation followed by posterior updating with observed
  evidence.

- **Tier 3 — Open-ended** (8 scenarios): Conceptual queries that do not
  require tool calls but should be answered with domain-accurate content
  citing the knowledge base. Examples include "How do Bayesian risk methods
  differ from Monte Carlo simulation?" and "My project's CPI is 0.85 and SPI
  is 0.92. Should I be concerned?"

Scoring follows a three-level rubric for Tiers 1 and 2: **Correct** (all
expected tools called), **Partial** (some expected tools called), and
**Incorrect** (no expected tools called). For Tier 3, responses are scored on
keyword coverage: **Correct** if ≥50% of domain-relevant keywords appear,
**Partial** if any keywords match, **Incorrect** otherwise. The evaluation
framework is implemented using the `vitals` R package [@vitals].

We test two configurations: `llama3.2` with RAG enabled (the default) and
`llama3.2` with RAG disabled (ablation). The `llama3.2` model (Meta, 3.2B
parameters) was chosen as a representative small local model accessible to
practitioners without GPU hardware. All inference runs locally via Ollama with
a 16,384-token context window. Each configuration is run in a single trial.

## 4.2 Results

Table 3 summarizes the evaluation results.

```{r results-table, echo=FALSE}
results <- data.frame(
  Configuration     = c("llama3.2 + RAG", "llama3.2 (no RAG)", "RAG delta"),
  Overall           = c("72.0%", "68.0%", "+4.0%"),
  `Single-tool (n=8)`  = c("87.5%", "100.0%", "-12.5%"),
  `Multi-tool (n=9)`   = c("55.6%", "44.4%",  "+11.1%"),
  `Open-ended (n=8)`   = c("75.0%", "62.5%",  "+12.5%"),
  check.names = FALSE
)
knitr::kable(results,
  caption = "Table 3. Tool-selection accuracy by configuration and tier.",
  align = c("l", "r", "r", "r", "r")
)
```

With RAG enabled, the agent achieves 72.0% overall accuracy: 87.5% on
single-tool scenarios, 55.6% on multi-tool chains, and 75.0% on open-ended
queries. Without RAG, overall accuracy falls to 68.0%, with 100.0% on
single-tool, 44.4% on multi-tool, and 62.5% on open-ended.

## 4.3 Discussion of Results

The RAG ablation reveals a clear pattern. RAG substantially improves
performance on multi-tool chains (+11.1 percentage points) and open-ended
conceptual queries (+12.5 pp), while reducing performance on single-tool
scenarios (−12.5 pp). We interpret this as follows.

For **multi-tool chains**, the retrieved context about method sequences (e.g.,
the `mcs_methods.md` document describes the typical workflow of
simulation → contingency → sensitivity) provides the LLM with planning context
that improves its ability to construct the correct tool call sequence. For
**open-ended queries**, the retrieved passages directly supply the domain
content the model needs, reducing reliance on the model's parametric knowledge
and improving keyword coverage. For **single-tool scenarios**, however, the
additional context appears to prime the model toward explanation over action:
when a passage about Monte Carlo methods is retrieved alongside a query
requesting a simulation, the model sometimes generates an explanatory response
about the method rather than immediately calling `mcs_tool`.

This trade-off is practically mitigated by the slash command interface.
Single-tool scenarios correspond precisely to the use case for which slash
commands were designed: the user knows which computation they want and provides
the data. Typing `/mcs tasks=[...]` guarantees correct tool invocation in
zero latency with no LLM involvement. The 100% single-tool accuracy of the
no-RAG configuration would be matched — and exceeded in speed and reliability
— by the deterministic command path.

Multi-tool chain performance (55.6% with RAG) represents the most significant
remaining limitation. Failures in this tier are predominantly attributable to
the model omitting the second tool in a chain — for example, running MCS but
not following through to `contingency_tool` — rather than calling the wrong
tool. This suggests that multi-step planning is the principal challenge for
a 3.2B parameter model at the task complexity level tested. We expect larger
models (8B+) and cloud-hosted models (GPT-4o, Claude Opus) to show
substantially higher accuracy on this tier, which we leave for future work.

# 5. Discussion

## 5.1 Implications for Risk Practice

The PRA agent framework lowers the barrier to quantitative risk analysis in
two complementary ways. First, the natural language interface means
practitioners no longer need to memorize function signatures or data formats:
they can describe their project in plain language and let the agent select
and invoke the appropriate analytical pipeline. Second, the slash command
interface means practitioners who have learned a workflow can execute it
instantly and reproducibly, without LLM latency or variability.

These two affordances serve different user profiles. A risk analyst encountering
a new project type benefits from the exploratory, conversational interface;
an experienced analyst running weekly EVM updates benefits from the
deterministic `/evm` command. The three-mode architecture accommodates both
within a single tool.

## 5.2 Comparison to Prior Approaches

Compared to **commercial tools** (Palisade \@RISK, Primavera Risk Analysis),
PRA offers full reproducibility through open-source R, no licensing cost, and
programmatic integration with the broader R statistical ecosystem. The AI
agent layer adds a natural language interface that these tools lack.

Compared to **general-purpose LLMs** (GPT-4, Claude, Gemini), PRA offers
validated quantitative computation. General-purpose LLMs can reason about
project risk in natural language but cannot reliably perform numerical
simulation, EVM calculation, or Bayesian updating; when they attempt to, they
frequently produce plausible-looking but incorrect numbers [@DeLLMa2024]. By
routing computation to validated R functions, PRA eliminates this failure mode.

Compared to **general R agent frameworks**, PRA provides domain-specific
tool definitions, a curated knowledge base, and a Shiny interface optimized
for project risk workflows. The MCP integration additionally enables users
to call PRA functions from Claude Code and other AI clients without writing
any R code.

## 5.3 Design Trade-offs

The choice to use local Ollama models by default prioritizes privacy and
accessibility over accuracy. The llama3.2 (3.2B) model achieves 72.0% overall
accuracy — adequate for exploratory use and reliable for single-tool scenarios,
but below what larger models would likely achieve for complex multi-tool chains.
Organizations with cloud access can substitute a more capable model; the
architecture is model-agnostic. A systematic model comparison benchmark is
left for future work.

The hybrid BM25 + vector similarity RAG retrieval performs well on the
domain-specific terminology in the knowledge base, but the current knowledge
corpus is intentionally small (~19 KB across six documents). Expanding the
corpus with project-specific standards, historical project data, or additional
risk frameworks would increase retrieval quality but also increase the risk of
retrieval noise degrading response coherence.

## 5.4 Generalizability

The three-mode routing architecture — deterministic commands, LLM tool-calling,
and RAG — is not specific to project risk analysis. Any domain that has (1) a
set of validated computational functions in R and (2) a corpus of domain
knowledge in text form can adopt the same architecture. The `ellmer` + `ragnar`
+ `mcptools` stack used here provides a reusable implementation pattern.

## 5.5 Limitations and Future Work

Several limitations of the current work point toward future research directions.

*Evaluation scope.* The 25-scenario benchmark uses a single model (llama3.2)
in a single trial. Multi-trial evaluation with confidence intervals, and
systematic comparison across model families and sizes (e.g., llama3.1 8B,
Qwen2.5, GPT-4o, Claude Opus), would provide a more complete picture of the
accuracy-model-size trade-off.

*Scoring methodology.* The tool-selection accuracy metric captures whether
the correct functions were called, but not whether the parameters were
correctly specified or whether the final answer was numerically correct. A
complementary end-to-end correctness metric — comparing tool output values
to reference solutions — would provide a more rigorous evaluation.

*Human baseline.* The evaluation does not include a human-baseline condition.
Comparing agent performance to that of domain experts operating the same
tools manually would provide the most meaningful measure of practical utility.

*Multi-turn conversations.* The current evaluation tests single-turn
performance. Real project risk assessments typically involve iterative
refinement — running a simulation, examining the results, adjusting
assumptions, re-running. Multi-turn benchmark design is an important next step.

*Data integration.* Future versions could integrate with project management
data sources (Microsoft Project, Primavera, ERP systems) to ingest schedule
and cost actuals automatically, removing the data-entry burden that is one
of the main practical barriers to quantitative risk analysis.

# 6. Conclusion

This paper presented an AI agent framework for quantitative project risk
analysis that integrates large language model tool-calling with a validated
suite of probabilistic analysis functions. The framework's three-mode routing
architecture — deterministic slash commands, LLM tool-calling, and
Retrieval-Augmented Generation — serves a range of user workflows: from
instant reproducible computation for experienced analysts to exploratory
natural language interaction for practitioners new to quantitative methods.

Systematic evaluation across 25 scenarios demonstrates 72.0% tool-selection
accuracy with a small local model (llama3.2, 3.2B parameters), with RAG
providing meaningful improvements for multi-tool chains (+11.1%) and
open-ended conceptual queries (+12.5%). The Model Context Protocol integration
extends the framework's reach to Claude, Claude Code, and other MCP-compatible
AI clients, enabling quantitative project risk analysis from any LLM interface.

The framework is fully open source, available on CRAN as the PRA package
(version 0.4.0), and operates entirely locally with no cloud API dependencies.
By combining the accessibility of natural language with the reliability of
validated quantitative tools, it offers a practical path toward broader
adoption of evidence-based quantitative risk analysis in project management.

# Acknowledgments

The author thanks the developers of the `ellmer`, `ragnar`, `mcptools`, and
`vitals` R packages, whose work provided the foundational infrastructure for
this framework.

# References
