---
title: "Probabilistic Networks"
output:
  html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Bayesian Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42)
```

## Introduction

Bayesian networks are probabilistic graphical models that represent variables and their conditional dependencies via a directed acyclic graph (DAG). Nodes represent random variables and edges represent dependencies between them. Bayesian networks are well-suited to project risk analysis because they can model how risk events propagate through resources and tasks to affect total project cost.

This vignette walks through a simple toy project to illustrate how to build, simulate, update, and learn from a Bayesian network using the `PRA` package. For a more advanced example covering causal inference, graph surgery, and the see-versus-do distinction across a full project portfolio, see the [Probabilistic Networks for Project Portfolio Risk Analysis](network2.html) vignette.

## Project

### Tasks

Consider a small software development project with three tasks.

```{r results='asis'}
tasks <- data.frame(
  ID    = c("F", "G", "H"),
  Label = c("Task-1", "Task-2", "Task-3"),
  Task  = c("Requirements and Design", "Development", "Testing and Handover")
)
knitr::kable(tasks, caption = "Project Tasks")
```

### Resources

Each task draws on one primary resource. The table below shows the baseline cost estimate (mean and standard deviation) for each resource.

```{r results='asis'}
resources <- data.frame(
  ID       = c("C", "D", "E"),
  Label    = c("Resource-1", "Resource-2", "Resource-3"),
  Resource = c("Business Analyst", "Developer", "QA Engineer"),
  Task_ID  = c("F", "G", "H"),
  Task     = c("Requirements and Design", "Development", "Testing and Handover"),
  Mean     = c(15000, 50000, 20000),
  SD       = c(3000, 10000, 4000)
)
knitr::kable(resources, caption = "Project Resources")
```

### Risks

Two risk events can escalate resource costs if they occur.

```{r results='asis'}
risks <- data.frame(
  Risk_ID           = c("A", "B"),
  Name              = c("Risk-1", "Risk-2"),
  Risk              = c("Requirements Scope Creep", "Technical Complexity"),
  Probability       = c(0.70, 0.60),
  Resource_ID       = c("C", "D"),
  Resource_Impacted = c("Business Analyst", "Developer"),
  Mean_if_occurs    = c(30000, 80000),
  SD_if_occurs      = c(8000, 20000)
)
knitr::kable(risks, caption = "Project Risks")
```

If Risk-1 (Requirements Scope Creep) occurs, the Business Analyst cost rises from a baseline of \$15,000 to a risk-adjusted mean of \$30,000. If Risk-2 (Technical Complexity) occurs, the Developer cost rises from \$50,000 to \$80,000. The QA Engineer cost is not directly affected by either risk.

## Bayesian Network

A Bayesian network models the full dependency chain from risks through resources to total project cost.

### Nodes

Nodes represent all variables in the network: risk events, resources, tasks, and the project total.

```{r}
nodes <- data.frame(
  id = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  label = c(
    "Risk-1", "Risk-2",
    "Resource-1", "Resource-2", "Resource-3",
    "Task-1", "Task-2", "Task-3",
    "Project"
  ),
  group = c(
    "Risk", "Risk",
    "Resource", "Resource", "Resource",
    "Task", "Task", "Task",
    "Project"
  ),
  stringsAsFactors = FALSE
)
```

### Edges

Edges encode the causal dependencies: risks affect resources, resources drive tasks, and tasks roll up to the project total.

```{r}
links <- data.frame(
  source = c("A", "B", "C", "D", "E", "F", "G", "H"),
  target = c("C", "D", "F", "G", "H", "I", "I", "I"),
  value  = rep(1, 8),
  stringsAsFactors = FALSE
)
```

### Distributions

Each risk is a binary discrete node (1 = occurs, 0 = does not occur). Each resource is a conditional node that follows a higher-cost distribution if its associated risk occurs, and a lower-cost baseline otherwise. Tasks and the project total are aggregate nodes that sum their inputs.

```{r}
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(
    type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)
  ),
  D = list(
    type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)
  ),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = c("C")),
  G = list(type = "aggregate", nodes = c("D")),
  H = list(type = "aggregate", nodes = c("E")),
  I = list(type = "aggregate", nodes = c("F", "G", "H"))
)
```

### Graph

```{r}
library(PRA)
graph <- prob_net(nodes, links, distributions = distributions)
```

The network can be visualized with the `igraph` and `networkD3` packages.

```{r message=FALSE}
library(igraph)
library(networkD3)
g   <- graph_from_data_frame(graph$links, vertices = graph$nodes, directed = TRUE)
d3g <- igraph_to_networkD3(g, group = graph$nodes$group)
forceNetwork(
  Links = d3g$links, Nodes = d3g$nodes, NodeID = "name", Group = "group",
  Value = "value", zoom = TRUE, legend = TRUE, arrows = TRUE,
  opacity = 0.8, fontSize = 14
)
```

## Inference

Use `prob_net_sim()` to forward-simulate the network and estimate the total project cost distribution.

```{r}
sim_results <- prob_net_sim(graph, num_samples = 10000)
```

```{r}
hist(sim_results$I, breaks = 60,
     main = "Total Project Cost", xlab = "Cost ($)",
     col = "skyblue", border = "white")
```

The spread of the distribution reflects compounded uncertainty from both risk events. The right tail represents scenarios where both risks occur simultaneously.

## Learning

Use `prob_net_learn()` to clamp one or more nodes to observed values and re-simulate. This shows the downstream effect of new information — for example, learning that Technical Complexity (Risk-2) did not materialise.

```{r}
# Numeric 0 or the string "No" can both be used for binary discrete nodes.
learn_results <- prob_net_learn(
  graph,
  observations = list(B = "No"),
  num_samples  = 10000
)
```

Comparing the Developer cost before and after the observation makes the shift clear.

```{r}
hist_before <- hist(sim_results$D,   breaks = 60, plot = FALSE)
hist_after  <- hist(learn_results$D, breaks = 60, plot = FALSE)

plot(
  hist_before,
  main   = "Developer Cost: Before vs. After Observing Risk-2 = No",
  xlab   = "Cost ($)",
  col    = "skyblue",
  border = "white",
  ylim   = c(0, max(hist_before$counts, hist_after$counts))
)
plot(hist_after, col = rgb(0, 0, 1, 0.5), border = "white", add = TRUE)
legend(
  "topright",
  legend = c("Before (Risk-2 uncertain)", "After (Risk-2 = No)"),
  fill   = c("skyblue", rgb(0, 0, 1, 0.5)),
  bty    = "n"
)
```

With Risk-2 ruled out, the Developer cost collapses to the lower baseline distribution.

## Updating

Use `prob_net_update()` to modify the network structure or distributions. Suppose a design review eliminates Requirements Scope Creep as a concern: remove the arc from Risk-1 to Resource-1 and replace the conditional distribution with a fixed normal.

```{r}
updated_graph <- prob_net_update(
  graph,
  remove_links = data.frame(source = "A", target = "C", stringsAsFactors = FALSE),
  update_distributions = list(
    C = list(type = "normal", mean = 15000, sd = 3000)
  )
)
updated_results <- prob_net_sim(updated_graph, num_samples = 10000)
```

```{r}
hist_before <- hist(sim_results$C,      breaks = 60, plot = FALSE)
hist_after  <- hist(updated_results$C,  breaks = 60, plot = FALSE)

plot(
  hist_before,
  main   = "Business Analyst Cost: Before vs. After Removing Risk-1",
  xlab   = "Cost ($)",
  col    = "skyblue",
  border = "white",
  ylim   = c(0, max(hist_before$counts, hist_after$counts))
)
plot(hist_after, col = rgb(0, 0, 1, 0.5), border = "white", add = TRUE)
legend(
  "topright",
  legend = c("Before (Risk-1 possible)", "After (Risk-1 removed)"),
  fill   = c("skyblue", rgb(0, 0, 1, 0.5)),
  bty    = "n"
)
```

With the risk arc removed, the Business Analyst cost tightens around the baseline mean and the heavy right tail disappears.

## Conclusion

This vignette demonstrated the core `PRA` Bayesian network workflow on a minimal toy project:

- `prob_net()` constructs the network from nodes, edges, and distributions.
- `prob_net_sim()` forward-simulates to estimate cost distributions.
- `prob_net_learn()` clamps observed nodes and re-simulates to propagate new evidence.
- `prob_net_update()` modifies network structure and distributions as the project evolves.

For a more advanced example covering a full project portfolio with shared enterprise risks, causal graph surgery, and the see-versus-do distinction, see the [Probabilistic Networks for Project Portfolio Risk Analysis](network2.html) vignette.
