# Your AI Co-Pilot: Agentic Risk Analysis {#sec-agent}

> *"The best way to predict the future is to invent it."*
> — Alan Kay

You've made it through eight chapters of quantitative risk analysis. You know Monte Carlo, SMM, EVM, Bayesian inference, learning curves, design structure matrices, and probabilistic networks. Now imagine not having to remember the syntax for all of them.

Imagine asking an AI assistant to "run Monte Carlo for three tasks with Normal(10,2), Triangular(5,10,15), Uniform(8,12)" and getting back a P95 date, a contingency reserve, and a tornado chart. That's the promise of the `PRA` agentic layer: every analytical function in this book is exposed as a tool that an AI agent can call directly, guided by a set of skills that tell it *when* and *how* to use each one.

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Explain how PRA exposes its functions as tools over the Model Context Protocol (MCP)
2. Start the PRA MCP server and register it with Claude Code or Claude Desktop
3. Describe what Agent Skills are and how they guide tool use
4. Map a natural-language request to the underlying PRA tool and its arguments
:::

## How PRA Talks to AI Agents

Rather than embedding a language model inside the package, PRA takes a simpler and more portable approach: it publishes its analytical functions as **tools** that any capable AI agent can call. Two pieces make this work:

1. **An MCP server** (`pra_mcp_server()`) that advertises each PRA function — Monte Carlo simulation, EVM, Bayesian risk, learning curves, DSM, and the rest — as a callable tool over the [Model Context Protocol](https://modelcontextprotocol.io).
2. **Agent Skills** — bundled `SKILL.md` files that document, in plain language, when to use each method, which tool to call, the argument format, and how to read the result.

The agent (Claude Code, Claude Desktop, or any MCP-compatible client) supplies the intelligence; PRA supplies the trustworthy, validated computation. Because the tools are the same functions you've used throughout the book, the numbers are identical whether you call `mcs()` yourself or ask an agent to run it.

## Prerequisites

The MCP server relies on a few optional packages:

```{r}
#| eval: false
install.packages(c("ellmer", "mcptools", "jsonlite"))
```

The analytical functions themselves need none of these — they are required only to serve the tools to an agent.

## The MCP Server

### Starting the server

```{r}
#| eval: false
library(PRA)
pra_mcp_server()
```

The server communicates over stdio, the standard transport for local MCP servers, and reuses the tool definitions from `pra_tools()` so any update to a tool is reflected automatically.

### Connecting from Claude Code

```bash
claude mcp add -s project pra -- Rscript -e "PRA::pra_mcp_server()"
```

Once registered, Claude Code can call PRA tools in any conversation:

> "Run a Monte Carlo simulation for three tasks: Task A normal(10, 2), Task B triangular(5, 10, 15), Task C uniform(8, 12). What is the contingency reserve at the 90th percentile?"

### Connecting from Claude Desktop

Add the following to your Claude Desktop config file:

```json
{
  "mcpServers": {
    "pra": {
      "command": "Rscript",
      "args": ["-e", "PRA::pra_mcp_server()"]
    }
  }
}
```

## Agent Skills

An agent works best when it knows not just *what* tools exist but *when* to reach for each one. PRA ships a set of Agent Skills for exactly this. Each is a `SKILL.md` file with structured guidance — when to use the method, which tool to call, the JSON argument format, and how to interpret the result.

Find the installed skills with:

```{r}
#| eval: false
list.files(system.file("skills", package = "PRA"))
```

| Skill | Covers | Primary tools |
|-------|--------|---------------|
| `pra-overview` | Routing / entry point | (directs to the others) |
| `pra-monte-carlo` | Schedule/cost uncertainty | `mcs_tool`, `smm_tool` |
| `pra-sensitivity-contingency` | Reserves & risk drivers | `contingency_tool`, `sensitivity_tool` |
| `pra-earned-value` | Performance & forecasting | `evm_analysis_tool` |
| `pra-bayesian-risk` | Risk probability & cost | `risk_prob_tool`, `risk_post_prob_tool`, `cost_pdf_tool`, `cost_post_pdf_tool` |
| `pra-learning-curves` | S-curve fitting/forecasts | `fit_and_predict_sigmoidal_tool` |
| `pra-dsm` | Task dependency structure | `parent_dsm_tool`, `grandparent_dsm_tool` |

## From Request to Tool Call

When you ask an agent to run an analysis, it selects a tool and fills in the arguments. All array and matrix arguments are passed as JSON strings. For example, the request "simulate three tasks — Normal(10,2), Triangular(5,10,15), Uniform(8,12), 10,000 runs" maps to `mcs_tool` with a `task_dists_json` of:

```json
[{"type":"normal","mean":10,"sd":2},
 {"type":"triangular","a":5,"b":10,"c":15},
 {"type":"uniform","min":8,"max":12}]
```

Under the hood, that is exactly the `mcs()` call you already know:

```{r}
library(PRA)
set.seed(42)
task_distributions <- list(
  list(type = "normal",     mean = 10, sd = 2),
  list(type = "triangular", a = 5, b = 10, c = 15),
  list(type = "uniform",    min = 8, max = 12)
)
result <- mcs(10000, task_distributions)
```

```{r}
#| fig-cap: "Monte Carlo results. The same output an agent receives when it calls `mcs_tool`."
hist(result$total_distribution,
  freq = FALSE, breaks = 50,
  main = "Monte Carlo Simulation Results",
  xlab = "Total Project Duration/Cost",
  col  = "#18bc9c80", border = "white"
)
curve(dnorm(x, mean = result$total_mean, sd = result$total_sd),
  add = TRUE, col = "#2c3e50", lwd = 2
)
abline(
  v   = quantile(result$total_distribution, c(0.50, 0.95)),
  col = c("#3498db", "#e74c3c"), lty = 2, lwd = 1.5
)
legend("topright",
  legend = c("Normal fit", "P50", "P95"),
  col    = c("#2c3e50", "#3498db", "#e74c3c"),
  lty    = c(1, 2, 2), lwd = c(2, 1.5, 1.5),
  cex    = 0.8, bg = "white"
)
```

A typical agent workflow chains tools the same way you would chain functions: `mcs_tool` to simulate, then `contingency_tool` for the reserve and `sensitivity_tool` for the risk drivers. The `pra-overview` skill lists these common workflows so the agent knows what to do next.

## Tool Reference

| Tool | Description |
|------|-------------|
| `mcs_tool` | Monte Carlo simulation with task distributions |
| `smm_tool` | Second Moment Method (analytical estimate) |
| `contingency_tool` | Contingency reserve from the last MCS |
| `sensitivity_tool` | Variance contribution per task |
| `evm_analysis_tool` | Full Earned Value Management analysis |
| `risk_prob_tool` | Bayesian prior risk probability |
| `risk_post_prob_tool` | Bayesian posterior risk after observations |
| `cost_pdf_tool` / `cost_post_pdf_tool` | Prior / posterior cost distributions |
| `fit_and_predict_sigmoidal_tool` | Sigmoidal learning curve fit and prediction |
| `parent_dsm_tool` / `grandparent_dsm_tool` | Design Structure Matrices |

## Summary

::: {.callout-tip icon=false}
## Key Takeaways

- PRA exposes every analytical function as a **tool over MCP** via `pra_mcp_server()`; the numbers are identical to calling the functions directly.
- **Agent Skills** (`SKILL.md` files under `system.file("skills", package = "PRA")`) tell an agent when and how to use each tool, including the JSON argument format.
- The package embeds **no language model** — only the MCP server needs the optional `ellmer`, `mcptools`, and `jsonlite` packages.
- Agents chain tools just as you chain functions: `mcs_tool` → `contingency_tool` + `sensitivity_tool`.
:::

Congratulations, you've completed the toolkit. You can now quantify project uncertainty (MCS, @sec-mcs; SMM, @sec-smm), track project performance (EVM, @sec-evm), update risk estimates with new evidence (Bayesian, @sec-bayes; Networks, @sec-network), model structural dependencies (DSM, @sec-dsm; Portfolio, @sec-network2), forecast learning (Sigmoidal, @sec-sigmoidal), and hand it all to an AI agent through MCP (@sec-agent). The rest is practice.

## Exercises

1. **Set up the server.** Install `ellmer`, `mcptools`, and `jsonlite`, then start `pra_mcp_server()` and register it with Claude Code or Claude Desktop using the instructions above. Confirm the client lists the PRA tools.

2. **Request to arguments.** For the request "simulate two tasks, Normal(20,4) and Uniform(10,15), with 5,000 runs," write out the `mcs_tool` call and its `task_dists_json` argument by hand. Then run the equivalent `mcs()` in R and compare.

3. **Chaining.** Ask an agent (or reason it out yourself) how to produce a P80 contingency reserve and a tornado chart for a 3-task project. Which tools are called, and in what order? Which skill documents that workflow?

4. **EVM from a prompt.** Using `evm_analysis_tool`'s arguments, translate this into a tool call: BAC = \$300K, schedule = [0.2, 0.4, 0.6, 0.8, 1.0], period = 4, complete = 0.55, costs = [\$55K, \$115K, \$175K, \$240K]. Then verify with the EVM functions from @sec-evm.

5. **Skill authoring.** ★ Read one of the bundled `SKILL.md` files, then draft a short skill for a project-specific workflow of your own (for example, a standard reserve policy). What guidance would an agent need to apply it correctly?
