# Everything Is Connected: Design Structure Matrices {#sec-dsm}

> *"You can't just fix Task 7. Task 7 shares three resources with Tasks 2, 4, and 9."*

Projects are not a list of independent activities. They are a web of interdependencies, and understanding that web is the key to understanding where delays and cost overruns actually come from. A delay in one task may ripple into five others — not because the tasks are sequentially linked, but because they share the same crew, the same materials, or the same equipment.

The Design Structure Matrix (DSM) makes these hidden connections visible and quantifiable [@browning2001; @steward1981].

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Construct a Resource-Task matrix **S** and a Risk-Resource matrix **R**
2. Compute the Parent DSM (resource-based coupling) using `parent_dsm()`
3. Compute the Grandparent DSM (risk-based coupling) using `grandparent_dsm()`
4. Interpret diagonal vs. off-diagonal DSM entries
5. Use DSM coupling values to prioritize risk mitigation decisions
:::

## What Is a DSM?

A DSM is a square matrix that maps dependencies between tasks in a project. In PRA, DSMs are derived from bipartite relationships: resources are shared across tasks, and risks are shared across resources. These shared dependencies create coupling between tasks that can propagate delays, cost overruns, or failures.

PRA provides two DSM functions:

- **`parent_dsm()`** — the Resource-based "Parent" DSM: shows how many resources are shared between each pair of tasks.
- **`grandparent_dsm()`** — the Risk-based "Grandparent" DSM: shows how many risks are shared between task pairs (via the resource layer).

The key idea from the resource-based view of project management [@govan2016]: shared resources are the structural pathway through which risks propagate. If you want to understand why two tasks always seem to get delayed together, look at what resources they share.

## The Resource-Task Matrix

The starting point is the Resource-Task Matrix **S**, where rows represent resources and columns represent tasks. An entry `S[i, j] = 1` means resource $i$ is used by task $j$.

```{r setup}
#| code-fold: false
library(PRA)
```

```{r}
# 4 resources x 5 tasks
S <- matrix(c(
  1, 0, 1, 0,
  1, 1, 0, 0,
  0, 1, 0, 1,
  0, 0, 1, 1,
  0, 1, 1, 0
), nrow = 4, ncol = 5)
rownames(S) <- paste0("R", 1:4)
colnames(S) <- paste0("T", 1:5)
S
```

## Parent DSM

The Parent DSM **P** = t(**S**) %*% **S** is a tasks-by-tasks matrix. The diagonal entry `P[j, j]` counts how many resources task $j$ uses. The off-diagonal entry `P[j, k]` counts how many resources tasks $j$ and $k$ share — a measure of coupling.

```{r}
p <- parent_dsm(S)
print(p)
```

```{r}
#| fig-cap: "Parent DSM heatmap. Darker cells indicate more shared resources between tasks — and therefore more potential for correlated disruption."
plot(p)
```

Tasks that share more resources are more tightly coupled. If a resource is delayed or constrained, all tasks that depend on it are affected simultaneously.

## The Risk-Resource Matrix

The Risk-Resource Matrix **R** adds a second layer. Rows represent risks and columns represent resources. An entry `R[i, j] = 1` means risk $i$ affects resource $j$.

```{r}
# 3 risks x 4 resources
R <- matrix(c(
  1, 0, 1,
  1, 1, 0,
  0, 1, 0,
  0, 0, 1
), nrow = 3, ncol = 4)
rownames(R) <- paste0("Risk", 1:3)
colnames(R) <- paste0("R", 1:4)
R
```

## Grandparent DSM

The Grandparent DSM traces the dependency chain from risks through resources to tasks. The intermediate matrix **T** = **R** %*% **S** gives a risks-by-tasks mapping, and the Grandparent DSM is **G** = t(**T**) %*% **T**. Off-diagonal entries count how many risks are shared between each pair of tasks.

```{r}
g <- grandparent_dsm(S, R)
print(g)
```

```{r}
#| fig-cap: "Grandparent DSM heatmap. Shows risk-based coupling: tasks that share risk exposure through common resources tend to fail together."
plot(g)
```

## Interpreting the DSM

- **Diagonal values** indicate the total number of resources (Parent) or risks (Grandparent) associated with each task. Higher values mean a task has more dependencies.
- **Off-diagonal values** indicate coupling between task pairs. Higher values mean more shared dependencies and greater potential for correlated disruption.
- **Symmetric structure**: both DSMs are always symmetric since shared dependencies are bidirectional — if task A shares a resource with task B, then task B shares that resource with task A.

A high off-diagonal value between tasks T2 and T5, for example, means those tasks share multiple resources. When risk strikes any of those resources, both tasks are hit simultaneously. This is not a coincidence — it is structural.

## From DSM to Decision

::: {.callout-tip}
## Using the DSM to Prioritize Mitigation

1. **Find the highest off-diagonal values** in the Parent DSM — these task pairs share the most resources and are most likely to fail together.
2. **Identify those shared resources** — they are the structural bottlenecks through which risk propagates.
3. **Inspect the Grandparent DSM** for those same pairs — if they also share risks, the coupling is doubly reinforced.
4. **Prioritize mitigation** on the shared resources with the highest combined Parent + Grandparent coupling: add contingency, build redundancy, or create buffers.
:::

The Grandparent DSM goes one level deeper: it shows which pairs of tasks are exposed to the same risks. If you can eliminate or mitigate a risk, you simultaneously reduce coupling between all tasks that share that risk.

## Summary

::: {.callout-tip icon=false}
## Key Takeaways

- The DSM makes **structural task dependencies** visible: not sequential dependencies, but resource-sharing dependencies that create correlated risk.
- **Parent DSM** (resource-based): `P = t(S) %*% S`. Off-diagonal `P[j,k]` = number of shared resources between tasks j and k.
- **Grandparent DSM** (risk-based): `G = t(T) %*% T` where `T = R %*% S`. Off-diagonal entries count shared risks via the resource layer.
- Both DSMs are always **symmetric** — shared dependencies are bidirectional.
- High off-diagonal values point directly to the resources and risks that deserve the most mitigation investment.
:::

The DSM quantifies *structural* coupling, but not the probabilistic dynamics of how risk propagates through it. For the full probabilistic treatment — including simulation and Bayesian updating over the same structure — see @sec-network and @sec-network2.

## Exercises

1. **Reading the DSM.** In the Parent DSM computed above, which pair of tasks has the highest coupling (highest off-diagonal value)? What does this mean in practical terms — what would happen if the shared resource became unavailable?

2. **Extend the matrix.** Add a 5th resource (R5) to the Resource-Task matrix **S** such that it is used by tasks T1 and T5. Recompute the Parent DSM. Which entries change, and why?

3. **Risk propagation.** ★ Look at the Grandparent DSM. Which task pair shares the most risks? Describe in plain English the risk scenario where this coupling causes the most damage.

4. **DSM symmetry.** Both the Parent and Grandparent DSMs are symmetric. Why? Is there a project structure where the DSM would not be symmetric? (Hint: think about directed dependencies — what if one task uses a resource but another task produces it?)

5. **From DSM to Monte Carlo.** ★ The off-diagonal values in the Parent DSM can be interpreted as a basis for correlation. Design a correlation matrix for MCS (from @sec-mcs) based on the Parent DSM: normalize the off-diagonal values so they range between 0 and 1, and use them as correlation coefficients. Run the MCS with this correlation structure and compare the total variance to the case with a zero correlation matrix. What did the structural coupling add to the risk picture?
