# Roll the Dice: Monte Carlo Simulation {#sec-mcs}

> *"Anyone who considers arithmetical methods of producing random digits is, of course, in a state of sin."*
> — John von Neumann

Here is the core insight behind Monte Carlo simulation: instead of guessing a single answer, roll the dice thousands of times and let the distribution speak for itself. The method was named after the famous casino in Monaco — not because it involves gambling, but because it involves random sampling. If you can describe your uncertainty with a probability distribution, Monte Carlo can turn that uncertainty into a full picture of possible project outcomes.

Monte Carlo simulation turns that randomness into insight.

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Explain the five steps of a Monte Carlo simulation
2. Run `mcs()` with multiple distributions and a correlation matrix
3. Interpret P50, P80, and P95 percentiles in project schedule context
4. Compute a schedule contingency reserve from simulation output
5. Identify which tasks drive the most variance using a tornado chart
:::

## How Monte Carlo Simulation Works

Monte Carlo (MC) simulation is a quantitative risk analysis technique that models uncertainty by running thousands of simulated project outcomes. Instead of using single-point estimates for task durations or costs, each task is described by a probability distribution. The simulation draws random samples from these distributions, sums them to get a total outcome, and repeats this thousands of times to build a full picture of possible project results [@vose2008].

::: {.callout-note}
## The Five Steps of Monte Carlo Simulation

1. **Model Definition** — Define project tasks and the variables that drive uncertainty (durations, costs).
2. **Assign Distributions** — Choose a probability distribution for each uncertain variable (e.g., triangular for tasks with optimistic/likely/pessimistic estimates).
3. **Specify Correlations** — If tasks are related (e.g., both affected by a shared risk), set a correlation coefficient between them.
4. **Run Simulation** — Draw random samples and compute the total outcome for each iteration (typically 10,000+).
5. **Analyze Results** — Summarize the distribution of totals using percentiles, mean, and variance.
:::

## Example

```{r setup}
#| code-fold: false
library(PRA)
library(ggplot2)
set.seed(42)
```

We model a 3-task project (in weeks). Task A follows a normal distribution, Task B has a triangular distribution (optimistic/most-likely/pessimistic), and Task C is uniformly distributed.

```{r}
num_simulations <- 10000
task_distributions <- list(
  list(type = "normal", mean = 10, sd = 2),        # Task A
  list(type = "triangular", a = 5, b = 10, c = 15), # Task B
  list(type = "uniform", min = 8, max = 12)          # Task C
)
```

### Correlation Matrix

Tasks often move together due to shared resources or external risks. The correlation matrix captures this. Values range from −1 (perfectly opposed) to +1 (perfectly aligned); 0 means independent. Here Tasks A and B have moderate positive correlation (0.5), meaning delays in one tend to coincide with delays in the other.

```{r}
correlation_matrix <- matrix(c(
  1.0, 0.5, 0.3,
  0.5, 1.0, 0.4,
  0.3, 0.4, 1.0
), nrow = 3, byrow = TRUE)
```

### Run the Simulation

```{r}
results <- mcs(num_simulations, task_distributions, correlation_matrix)
```

```{r}
cat("Mean Total Duration:  ", round(results$total_mean, 2), "weeks\n")
cat("Variance of Duration: ", round(results$total_variance, 2), "\n")
cat("Std Dev of Duration:  ", round(results$total_sd, 2), "weeks\n")
```

### Distribution of Outcomes

The histogram below shows all 10,000 simulated total durations. The overlaid density curve reveals the shape of the distribution — notice the slight right skew from the triangular task.

```{r}
#| fig-cap: "Monte Carlo simulation results showing the full distribution of possible total project durations."
hist_data <- results$total_distribution

hist(hist_data,
  breaks = 50, freq = FALSE,
  main = "Monte Carlo Simulation — Total Project Duration",
  xlab = "Total Duration (weeks)", col = "steelblue", border = "white"
)
lines(density(hist_data), col = "tomato", lwd = 2)
abline(v = results$total_mean, col = "black", lty = 2, lwd = 1.5)
legend("topright",
  legend = c("Density", paste0("Mean = ", round(results$total_mean, 1), " wks")),
  col = c("tomato", "black"), lty = c(1, 2), lwd = 2, bty = "n"
)
```

## Interpreting Percentiles

The `mcs()` function returns key percentiles of the total distribution. These answer the question: *"What duration has X% probability of not being exceeded?"*

```{r}
knitr::kable(
  data.frame(
    Percentile = c("P5", "P50 (Median)", "P95"),
    Duration   = round(results$percentiles, 1),
    Meaning    = c(
      "5% chance of finishing this fast or faster",
      "Equal chance of finishing above or below this",
      "95% chance of finishing by this date"
    )
  ),
  caption = "Simulation Percentiles"
)
```

The P50 is your base estimate — the median, not the mean (though for symmetric distributions they're close). The P95 is the date you can be 95% confident in. The gap between P50 and P95 is your schedule risk.

## Contingency Analysis

Contingency is the buffer added above the base estimate to cover uncertainty. A common approach is to use the difference between the P95 (or chosen confidence level) outcome and the P50 (base estimate) [@pmbok2021].

```{r}
contingency_val <- contingency(results, phigh = 0.95, pbase = 0.50)
cat("Schedule contingency (P95 − P50):", round(contingency_val, 2), "weeks\n")
cat(
  "There is a 95% chance the project will finish within",
  round(results$percentiles["95%"], 1), "weeks.\n"
)
```

Adding this contingency to the P50 estimate gives a 95% confidence of on-time delivery. Teams with low risk tolerance should use P95; those with higher tolerance might use P80.

## Sensitivity Analysis

Sensitivity analysis identifies which tasks drive the most variability in the total outcome — the tasks that deserve the most management attention. The result is a **tornado chart**: the wider the bar, the bigger the impact on total schedule risk.

```{r}
#| fig-cap: "Tornado chart showing each task's contribution to total schedule variance. Focus risk mitigation on the tallest bars."
sensitivity_results <- sensitivity(task_distributions, correlation_matrix)

sens_data <- data.frame(
  Task        = c("Task A (Normal)", "Task B (Triangular)", "Task C (Uniform)"),
  Sensitivity = sensitivity_results
)

p <- ggplot2::ggplot(
  sens_data,
  ggplot2::aes(x = Sensitivity, y = reorder(Task, Sensitivity))
) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::geom_text(ggplot2::aes(label = round(Sensitivity, 3)),
    hjust = -0.1, size = 3.5
  ) +
  ggplot2::labs(
    title = "Tornado Chart — Task Sensitivity",
    x     = "Sensitivity Coefficient",
    y     = NULL
  ) +
  ggplot2::xlim(0, max(sensitivity_results) * 1.2) +
  ggplot2::theme_minimal()

print(p)
```

Tasks with larger bars contribute more variance to the total. Prioritize risk mitigation efforts on the highest-sensitivity task. Even a small reduction in its uncertainty can meaningfully reduce overall project risk.

## Summary

::: {.callout-tip icon=false}
## Key Takeaways

- Monte Carlo simulation runs thousands of iterations to produce a full distribution of project outcomes, not just a single number.
- **Percentiles** (P50, P80, P95) translate simulation output into concrete schedule or cost commitments at a stated confidence level.
- **Contingency** = P_high − P50: the buffer needed to achieve a given confidence level above the base estimate.
- **Sensitivity analysis** (tornado chart) reveals which tasks drive the most variance — focus mitigation there first.
- Positive correlation between tasks *always* increases total variance; ignoring it leads to underestimating risk.
:::

For early-stage estimates when full distributions are unavailable, the Second Moment Method (@sec-smm) provides a fast analytical alternative using only means and variances.

## Exercises

1. **Conceptual check.** What does increasing the number of simulations from 10,000 to 100,000 do to the result? Does it change the mean? The P95? The smoothness of the histogram? Try it.

2. **New task.** Add a 4th task with a Beta distribution (hint: look at `?mcs` to see if Beta is supported, or use a normal approximation). Run `mcs()` with your updated task list and compare the P80 to the 3-task version. Did the project get riskier or safer?

3. **Correlation experiment.** Re-run the simulation with all off-diagonal correlation values set to 0 (fully independent tasks). Then re-run with all off-diagonal values set to 0.9 (strongly correlated). How does total variance change in each case? Explain intuitively why strong positive correlation increases portfolio risk. ★

4. **Choosing a confidence level.** Your client asks for a "90% confident" delivery date. What percentile do you use, and what contingency does that imply? Recalculate using `contingency(results, phigh = 0.90, pbase = 0.50)`.

5. **Real-world application.** ★ Think of a project you know. Identify three tasks with significant uncertainty. Estimate (or guess) a distribution type and rough parameters for each. Run `mcs()` and report the P50 and P80 schedule. What's the contingency at P80?
