# It's a Small World After All: Probabilistic Networks {#sec-network}

> *"Everything is connected to everything else."*
> — Barry Commoner's First Law of Ecology

A risk register lists risks. A probability network shows you how those risks talk to each other, and to your project. It's the difference between knowing that "Technical Complexity" is a risk and understanding exactly how it flows through your developer costs into your total project budget — and what happens to that flow when you learn new information.

Bayesian networks are the tool. A Bayesian network is a specific type of probabilistic network — a directed acyclic graph (DAG) in which each node represents a random variable and each edge encodes a conditional probability relationship. The "probabilistic network" framing in this chapter's title is intentional: the core concepts (conditioning, propagation, graph structure) apply to the broader class, and `PRA` uses the Bayesian network formulation specifically. They combine graph theory with probability theory to model the full dependency structure of a project: which risks drive which resources, which resources drive which tasks, and how uncertainty propagates all the way up to the project total.

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Build a probabilistic network using `prob_net()` with nodes, edges, and distributions
2. Run forward simulations with `prob_net_sim()` and interpret the cost distribution
3. Incorporate new evidence using `prob_net_learn()` and observe downstream shifts
4. Modify network structure and distributions with `prob_net_update()`
5. Explain the difference between learning (conditioning) and updating (graph surgery)
:::

## What Is a Bayesian Network?

::: {.callout-note}
## Bayesian Network vs. DSM

The Design Structure Matrix (@sec-dsm) shows *structural* coupling — how many resources two tasks share. A Bayesian network goes further: it encodes **probabilistic dependencies**, meaning the actual conditional distributions of costs given risk states. Where the DSM counts connections, the network simulates consequences.

Use a DSM for rapid structural triage. Use a Bayesian network when you need actual cost distributions and the ability to condition on observed evidence.
:::

A Bayesian network is a directed acyclic graph (DAG) where:

- **Nodes** represent random variables (risks, resources, tasks, project totals)
- **Edges** represent conditional dependencies (Risk A affects Resource C)
- **Distributions** encode the uncertainty at each node

Bayesian networks are well-suited to project risk analysis because they explicitly model how risk events propagate through resources and tasks to affect total project cost [@govan2014].

For a more advanced example covering causal inference, graph surgery, and the see-versus-do distinction across a full project portfolio, see @sec-network2.

## Project Setup

### Tasks

Consider a small software development project with three tasks.

```{r setup}
#| code-fold: false
library(PRA)
set.seed(42)
```

```{r}
tasks <- data.frame(
  ID    = c("F", "G", "H"),
  Label = c("Task-1", "Task-2", "Task-3"),
  Task  = c("Requirements and Design", "Development", "Testing and Handover")
)
knitr::kable(tasks, caption = "Project Tasks")
```

### Resources

Each task draws on one primary resource. The table below shows the baseline cost estimate for each resource.

```{r}
resources <- data.frame(
  ID       = c("C", "D", "E"),
  Label    = c("Resource-1", "Resource-2", "Resource-3"),
  Resource = c("Business Analyst", "Developer", "QA Engineer"),
  Task_ID  = c("F", "G", "H"),
  Mean     = c(15000, 50000, 20000),
  SD       = c(3000,  10000,  4000)
)
knitr::kable(resources, caption = "Project Resources")
```

### Risks

Two risk events can escalate resource costs if they occur.

```{r}
risks <- data.frame(
  Risk_ID        = c("A", "B"),
  Risk           = c("Requirements Scope Creep", "Technical Complexity"),
  Probability    = c(0.70, 0.60),
  Resource       = c("Business Analyst", "Developer"),
  Mean_if_occurs = c(30000, 80000),
  SD_if_occurs   = c(8000,  20000)
)
knitr::kable(risks, caption = "Project Risks")
```

If Risk-1 (Requirements Scope Creep) occurs, the Business Analyst cost rises from \$15,000 to \$30,000. If Risk-2 (Technical Complexity) occurs, the Developer cost rises from \$50,000 to \$80,000. The QA Engineer is unaffected by either risk.

## Building the Bayesian Network

### Nodes

```{r}
nodes <- data.frame(
  id    = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
  label = c(
    "Risk-1", "Risk-2",
    "Resource-1", "Resource-2", "Resource-3",
    "Task-1", "Task-2", "Task-3",
    "Project"
  ),
  group = c(
    "Risk", "Risk",
    "Resource", "Resource", "Resource",
    "Task", "Task", "Task",
    "Project"
  ),
  stringsAsFactors = FALSE
)
```

### Edges

Edges encode the causal dependencies: risks affect resources, resources drive tasks, and tasks roll up to the project total.

```{r}
links <- data.frame(
  source = c("A", "B", "C", "D", "E", "F", "G", "H"),
  target = c("C", "D", "F", "G", "H", "I", "I", "I"),
  value  = rep(1, 8),
  stringsAsFactors = FALSE
)
```

### Distributions

```{r}
distributions <- list(
  A = list(type = "discrete", values = c(1, 0), probs = c(0.70, 0.30)),
  B = list(type = "discrete", values = c(1, 0), probs = c(0.60, 0.40)),
  C = list(
    type = "conditional", condition = "A",
    true_dist  = list(type = "normal", mean = 30000, sd = 8000),
    false_dist = list(type = "normal", mean = 15000, sd = 3000)
  ),
  D = list(
    type = "conditional", condition = "B",
    true_dist  = list(type = "normal", mean = 80000, sd = 20000),
    false_dist = list(type = "normal", mean = 50000, sd = 10000)
  ),
  E = list(type = "normal", mean = 20000, sd = 4000),
  F = list(type = "aggregate", nodes = c("C")),
  G = list(type = "aggregate", nodes = c("D")),
  H = list(type = "aggregate", nodes = c("E")),
  I = list(type = "aggregate", nodes = c("F", "G", "H"))
)
```

### Build the Graph

```{r}
graph <- prob_net(nodes, links, distributions = distributions)
```

The network can be visualized with the `igraph` and `networkD3` packages.

```{r}
#| label: fig-network-graph-setup
library(igraph)
library(networkD3)
g   <- graph_from_data_frame(graph$links, vertices = graph$nodes, directed = TRUE)
d3g <- igraph_to_networkD3(g, group = graph$nodes$group)
```

```{r}
#| eval: !expr knitr::is_html_output()
#| fig-cap: "Probabilistic network of risks, resources, tasks, and project cost."
forceNetwork(
  Links = d3g$links, Nodes = d3g$nodes, NodeID = "name", Group = "group",
  Value = "value", zoom = TRUE, legend = TRUE, arrows = TRUE,
  opacity = 0.8, fontSize = 14
)
```

```{r}
#| eval: !expr "!knitr::is_html_output()"
#| fig-cap: "Probabilistic network of risks, resources, tasks, and project cost."
plot(
  g,
  vertex.color = as.factor(graph$nodes$group),
  vertex.size = 14, vertex.label.cex = 0.7,
  edge.arrow.size = 0.4, layout = layout_with_sugiyama(g)$layout
)
```

## Inference: Forward Simulation

Use `prob_net_sim()` to forward-simulate the network and estimate the total project cost distribution.

```{r}
sim_results <- prob_net_sim(graph, num_samples = 10000)
```

```{r}
#| fig-cap: "Total project cost distribution from 10,000 forward simulations. The right tail represents scenarios where both risks occur simultaneously."
hist(sim_results$I, breaks = 60,
     main = "Total Project Cost",
     xlab = "Cost ($)", col = "skyblue", border = "white")
```

The spread reflects compounded uncertainty from both risk events. The right tail represents the worst case: both risks occur.

## Learning: Incorporating New Evidence

Use `prob_net_learn()` to clamp one or more nodes to observed values and re-simulate. This shows the downstream effect of new information — for example, learning that Technical Complexity (Risk-2) did not materialize.

```{r}
learn_results <- prob_net_learn(
  graph,
  observations = list(B = "No"),
  num_samples  = 10000
)
```

```{r}
#| fig-cap: "Developer cost before and after observing Risk-2 = No. The distribution collapses to the lower baseline."
hist_before <- hist(sim_results$D,   breaks = 60, plot = FALSE)
hist_after  <- hist(learn_results$D, breaks = 60, plot = FALSE)

plot(
  hist_before,
  main  = "Developer Cost: Before vs. After Observing Risk-2 = No",
  xlab  = "Cost ($)", col = "skyblue", border = "white",
  ylim  = c(0, max(hist_before$counts, hist_after$counts))
)
plot(hist_after, col = rgb(0, 0, 1, 0.5), border = "white", add = TRUE)
legend(
  "topright",
  legend = c("Before (Risk-2 uncertain)", "After (Risk-2 = No)"),
  fill   = c("skyblue", rgb(0, 0, 1, 0.5)), bty = "n"
)
```

With Risk-2 ruled out, the Developer cost collapses to the lower baseline distribution — and the total project cost shifts left accordingly.

## Updating: Modifying the Network

Use `prob_net_update()` to modify the network structure or distributions. Suppose a design review eliminates Requirements Scope Creep as a concern: remove the arc from Risk-1 to Resource-1 and replace the conditional distribution with a fixed normal.

```{r}
updated_graph <- prob_net_update(
  graph,
  remove_links = data.frame(source = "A", target = "C", stringsAsFactors = FALSE),
  update_distributions = list(
    C = list(type = "normal", mean = 15000, sd = 3000)
  )
)
updated_results <- prob_net_sim(updated_graph, num_samples = 10000)
```

```{r}
#| fig-cap: "Business Analyst cost before and after removing the Risk-1 arc. The heavy right tail disappears."
hist_before <- hist(sim_results$C,      breaks = 60, plot = FALSE)
hist_after  <- hist(updated_results$C,  breaks = 60, plot = FALSE)

plot(
  hist_before,
  main  = "Business Analyst Cost: Before vs. After Removing Risk-1",
  xlab  = "Cost ($)", col = "skyblue", border = "white",
  ylim  = c(0, max(hist_before$counts, hist_after$counts))
)
plot(hist_after, col = rgb(0, 0, 1, 0.5), border = "white", add = TRUE)
legend(
  "topright",
  legend = c("Before (Risk-1 possible)", "After (Risk-1 removed)"),
  fill   = c("skyblue", rgb(0, 0, 1, 0.5)), bty = "n"
)
```

## The Four Core Functions

| Function | What it does |
|----------|-------------|
| `prob_net()` | Constructs the network from nodes, edges, and distributions |
| `prob_net_sim()` | Forward-simulates to estimate cost distributions |
| `prob_net_learn()` | Clamps observed nodes and re-simulates to propagate evidence |
| `prob_net_update()` | Modifies network structure and distributions as the project evolves |

## Summary

::: {.callout-tip icon=false}
## Key Takeaways

- A probabilistic network models the full dependency chain from risks → resources → tasks → project cost, encoding uncertainty at every layer.
- **`prob_net_sim()`** draws Monte Carlo samples in topological order — each run propagates risk states through the network to produce a total cost distribution.
- **`prob_net_learn()`** conditions on observed evidence (*seeing*): clamping a node to its observed value and re-simulating the downstream distribution.
- **`prob_net_update()`** performs structural changes (*doing*): removing edges or replacing distributions, enabling causal interventions beyond what conditioning can express.
- The difference between *seeing* and *doing* becomes consequential when nodes share upstream parents — see @sec-network2 for the full treatment with shared enterprise risks.
:::

## Exercises

1. **Learning effect.** In the example, we observed Risk-2 = No. What do you expect happens to the total project cost distribution (node I) when Risk-2 is observed as "Yes" instead? Test your prediction by running `prob_net_learn()` with `list(B = "Yes")` and plotting the result.

2. **Modify a risk.** Change the probability of Risk-1 from 0.70 to 0.30. How does this affect the mean total project cost from `prob_net_sim()`? Is the change proportional to the probability change?

3. **Add a QA risk.** Modify the network so that a new Risk-3 (with probability 0.40) affects the QA Engineer (Resource-3), increasing their cost from a mean of \$20K to \$35K if it occurs. Update the distributions and re-simulate. How much does this add to the expected project cost? ★

4. **Seeing vs. doing.** Explain the difference between `prob_net_learn(observations = list(B = "No"))` and `prob_net_update(remove_links = ...)`. When would you use each? Which corresponds to "seeing" and which to "doing" in the causal inference sense? (See @sec-network2 for the full treatment.)

5. **Project structure.** ★ This example has a simple layered structure (Risk → Resource → Task → Project). Design a more complex network with two risks that both affect the same resource. What does that mean for the correlation between the two downstream tasks? Build the network and verify with a correlation matrix of `sim_results`.
