# Write Once, Trust Always: Reproducible Risk Analysis with Quarto {#sec-quarto}

> *"The first principle is that you must not fool yourself — and you are the easiest person to fool."*
> — Richard Feynman

You've just run a Monte Carlo simulation, computed a P80 schedule, and printed the contingency reserve for your project. Now imagine sending that result to a colleague, then trying to recreate it six months later. The spreadsheet is gone. The file is named "Final_v3_REVISED_use_this_one.xlsx." Three cells have broken formulas and nobody remembers which inputs were authoritative.

This is the reproducibility gap in project risk analysis, and it is more common than anyone likes to admit.

Quarto closes that gap. It lets you write your analysis and your prose in the same document, so that when the document renders, R executes the code and embeds the actual output: the numbers, the plots, the tables. Not a screenshot. Not a copy-paste. The live result, freshly computed, every time.

This chapter teaches you how to build reproducible risk reports with Quarto and the `PRA` package. By the end, you will have a workflow you can hand to a colleague, run on a new project, or re-execute in two years and get the same answer.

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Explain why reproducibility is a professional standard in quantitative risk analysis
2. Create a Quarto document that embeds live `PRA` code and narrative prose
3. Use parameterized reports to run the same analysis across multiple project scenarios
4. Version-control a risk analysis workflow with Git
5. Describe how this book itself is built using the same Quarto framework
:::

## The Spreadsheet Problem

Spreadsheets are the default tool for project risk analysis. They are flexible, familiar, and require no installation. They are also, quietly, one of the biggest threats to analytical credibility.

Here is a realistic timeline:

1. A risk analyst builds a Monte Carlo model in Excel. It works. The P90 cost is $4.2M.
2. The file is emailed, renamed, and forwarded. Three versions exist by Tuesday.
3. Six months later, the project is audited. The analyst is on a new project. Nobody can open the file without `#REF!` errors.
4. The P90 estimate can no longer be verified. The basis of the contingency reserve is, in practice, lost.

This is not a spreadsheet problem specifically — it is a **reproducibility problem**. The analysis was never in a form that could be reliably re-executed.

::: {.callout-note}
## Reproducible vs. Repeatable

**Reproducible** means the same analyst, on the same machine, can re-run the analysis and get the same result.

**Repeatable** means a *different* analyst, on a *different* machine, can run the analysis and get the same result, given the same inputs.

A Quarto document checked into version control with a declared package environment achieves both. A spreadsheet achieves neither reliably.
:::

The alternative is to make your analysis a **document that also runs**. That is what Quarto does.

## What is Quarto?

Quarto is an open-source publishing system that combines prose and executable code in a single `.qmd` file. You write in Markdown. You embed code chunks in R, Python, Julia, or Observable. When you render the document, Quarto runs the code and weaves the results into the output — HTML, PDF, Word, or a full website or book.

::: {.callout-note}
## Quarto vs. R Markdown

If you have used R Markdown (`.Rmd`), Quarto will feel immediately familiar. Quarto is the next-generation version: it supports multiple languages, has a cleaner YAML syntax, and produces higher-quality output formats out of the box.

If you have not used either: Quarto is where to start. Install it once from [quarto.org](https://quarto.org) and it works with RStudio, VS Code, or any text editor.
:::

A minimal Quarto document looks like this:

```yaml
---
title: "Bridge Rehabilitation: Schedule Risk Analysis"
author: "Your Name"
date: today
format: html
---
```

````markdown
## Overview

This report summarizes the Monte Carlo schedule simulation for the
bridge rehabilitation project.

```{{r}}
library(PRA)

tasks <- list(
  list(dist = "normal",   params = list(mean = 30, sd = 4)),
  list(dist = "lognormal", params = list(meanlog = log(20), sdlog = 0.2)),
  list(dist = "uniform",   params = list(min = 10, max = 25))
)

results <- mcs(tasks)
print(results)
```
````

Run `quarto render report.qmd` in your terminal and you have a finished HTML report, numbers and all, generated from a single source file.

## Your First Risk Report

Let us build one from scratch.

### Step 1: Install Quarto

Download and install Quarto from [quarto.org](https://quarto.org/docs/get-started/). If you use RStudio, version 2022.07 or later includes Quarto support out of the box.

### Step 2: Create the Document

Create a new file called `risk_report.qmd` and add this header:

```yaml
---
title: "Project Schedule Risk Analysis"
author: "Risk Analyst"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---
```

### Step 3: Write the Analysis

Below the YAML header, write your prose and embed your R code:

````markdown
## Background

This analysis quantifies schedule uncertainty for a three-phase
infrastructure project using Monte Carlo simulation.

```{{r}}
library(PRA)

task_dists <- list(
  list(type = "normal",     mean = 45, sd = 6),          # Design
  list(type = "triangular", a = 20, b = 30, c = 50),     # Permits
  list(type = "uniform",    min = 50, max = 80)           # Construction
)

set.seed(42)
results <- mcs(10000, task_dists)
print(results)
```

## Contingency Reserve

At the P80 confidence level, the recommended contingency is:

```{{r}}
contingency(results, phigh = 0.80, pbase = 0.50)
```
````

### Step 4: Render

From your terminal:

```bash
quarto render risk_report.qmd
```

Or from RStudio, click **Render**. Quarto runs both code chunks, embeds the output inline, and writes `risk_report.html`. Open it in any browser.

Here is what the embedded output looks like when you run those chunks live:

```{r}
#| label: first-report
library(PRA)

task_dists <- list(
  list(type = "normal",     mean = 45, sd = 6),
  list(type = "triangular", a = 20, b = 30, c = 50),
  list(type = "uniform",    min = 50, max = 80)
)

set.seed(42)
results <- mcs(10000, task_dists)
print(results)
```

```{r}
#| label: contingency-result
contingency(results, phigh = 0.80, pbase = 0.50)
```

That output was not typed by hand. It was computed when this book was built.

## Parameterized Reports

So far the analysis is fixed to one project. Real workflows require running the same model against multiple scenarios — different task durations, different confidence targets, different projects entirely.

Quarto's `params:` block makes this straightforward.

### Defining Parameters

Add a `params:` section to your YAML header:

```yaml
---
title: "Schedule Risk Analysis"
format: html
params:
  mean_design:   45
  mean_permits:  30
  mean_construct: 60
  confidence:    0.80
---
```

### Using Parameters in Code

Inside any code chunk, access parameters via `params$`:

````markdown
```{{r}}
library(PRA)

task_dists <- list(
  list(type = "normal",     mean = params$mean_design,   sd = 6),
  list(type = "normal",     mean = params$mean_permits,  sd = 4),
  list(type = "normal",     mean = params$mean_construct, sd = 8)
)

set.seed(42)
results <- mcs(10000, task_dists)
contingency(results, phigh = params$confidence, pbase = 0.50)
```
````

### Rendering Multiple Scenarios

You can render the same `.qmd` with different parameter values from the command line:

```bash
# Optimistic scenario
quarto render risk_report.qmd \
  -P mean_design:35 -P mean_permits:22 -P mean_construct:50 \
  --output optimistic.html

# Conservative scenario
quarto render risk_report.qmd \
  -P mean_design:55 -P mean_permits:40 -P mean_construct:75 \
  --output conservative.html
```

Two reports, one template, zero copy-paste. The same audit trail applies to both.

::: {.callout-note}
## Parametric Reporting in Practice

A common workflow on large programs is to maintain one `risk_template.qmd` and a small data file (CSV or JSON) that drives scenario generation. A short R script loops over the data file and calls `quarto render` once per row. You end up with one authoritative report per project phase or scenario, all traceable to the same model.
:::

## Version Control with Git

A reproducible analysis that lives only on your laptop is not fully reproducible — it is one hard drive failure away from being lost. The second half of the reproducibility story is **version control**.

### What to Commit

The rule of thumb for a Quarto risk analysis:

| File | Commit? | Reason |
|------|---------|--------|
| `risk_report.qmd` | **Yes** | This is the source of truth |
| `data/inputs.csv` | **Yes** | Inputs must be versioned with the analysis |
| `risk_report.html` | No | Derived output; regenerate on demand |
| `risk_report_files/` | No | Auto-generated by Quarto |
| `.Rprofile`, `renv.lock` | **Yes** | Locks package versions for repeatability |

The `.qmd` file and its inputs are the authoritative record. Everything else is a by-product.

### A Minimal Workflow

```bash
git init
echo "*.html" >> .gitignore
echo "*_files/" >> .gitignore

git add risk_report.qmd data/inputs.csv renv.lock
git commit -m "Initial Monte Carlo schedule analysis, P80 = 142 days"
```

When you update the analysis — new task added, input data revised — commit again:

```bash
git add -A
git commit -m "Add permits phase; P80 increases to 158 days"
```

The git log becomes an audit trail. Anyone can check out any prior commit and re-render the analysis to verify a historical result.

::: {.callout-note}
## Package Environments with renv

Reproducibility also requires locking the versions of R packages your analysis depends on. The `renv` package does this automatically:

```r
install.packages("renv")
renv::init()   # creates renv.lock — commit this file
renv::restore() # anyone can restore the exact same packages later
```

Without `renv`, an analysis that ran on `PRA` version 0.3.0 might silently produce different results on version 0.4.0 if a function changes behavior.
:::

## This Book as a Living Example

This book is itself a Quarto project built on exactly the framework described in this chapter.

Every code block you have read throughout this book — the `mcs()` call in @sec-mcs, the `evm` calculations in @sec-evm, the Bayesian updates in @sec-bayes — was executed when the book was built. There is no "trust us, this is what you'd see." The numbers are real, freshly computed from the current version of the `PRA` package.

The book's source is a directory of `.qmd` files coordinated by a single `_quarto.yml`:

```yaml
project:
  type: book

book:
  title: "Project Risk Analysis: A Practical Guide"
  chapters:
    - index.qmd
    - intro.qmd
    - ch-mcs.qmd
    - ch-smm.qmd
    # ... one file per chapter ...
    - ch-quarto.qmd   # this chapter
    - ch-agent.qmd
```

To reproduce the entire book from scratch:

```bash
git clone https://github.com/paulgovan/PRA
cd PRA
quarto render book/
```

That single command runs every code chunk in every chapter, in a clean R session, and writes the full HTML site, PDF, and EPUB to `book/_book/`. If any code breaks — because a function changed behavior or a dependency was removed — the render fails loudly. The book cannot silently go out of date.

::: {.callout-note}
## What "Live Book" Means for You

Because the book is reproducible, you can:

- **Verify any result.** Copy a code chunk, change a parameter, and see immediately how the output shifts.
- **Trust the numbers.** They were not typed by hand and are not screenshots from an old run.
- **Adapt the structure.** The chapter files are plain Markdown with embedded R. Clone the repo, swap in your project's data, and render your own custom risk analysis book.

The same `.qmd` + `PRA` + `quarto render` workflow that produced this book can produce your project's quarterly risk report.
:::

## Exercises

1. **Create a minimal report.** Write a `risk_report.qmd` with a YAML header, a one-paragraph background section, and a single code chunk that runs `mcs()` on three tasks of your choice. Render it to HTML and verify the output appears inline.

2. **Add parameterization.** Extend the report from Exercise 1 with a `params:` block. Define `confidence` as a parameter (default `0.80`). Use `params$confidence` in the `contingency()` call. Re-render with `-P confidence:0.90` from the terminal and confirm the contingency increases.

3. **Render to PDF.** Add `pdf: default` to your `format:` block and re-render. Open the PDF. Note any differences in how the plots and tables appear. Try the `geometry: margin=1in` option to control margins.

4. **Inspect this book's source.** Clone the `PRA` repository from GitHub and open `book/ch-mcs.qmd` in a text editor. Identify: (a) the YAML header, (b) the learning objectives callout, (c) the code chunk that calls `mcs()`, and (d) the exercises section. Then run `quarto render book/ch-mcs.qmd` and compare the rendered output to the source. ★

5. **Version-control your analysis.** Initialize a Git repository in the folder containing your `risk_report.qmd`. Commit the `.qmd` file. Make a change (add a task, adjust a parameter) and make a second commit. Run `git log` and confirm you have a two-commit history that documents what changed and when. ★
